% ============================================================
%  ch1_introduction.tex
%  Chapter 1: Introduction
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% !TEX root = Thesis.tex
\chapter{Introduction}
\label{chap:intro}

Pattern matching is the process of analyzing a sequence of data
points whether rows in a relational table or events in a data
stream with the goal of identifying specific shapes, trends, or
behavior sequences within the data.  It plays a crucial role
across many analytical domains: detecting chains of transactions
associated with money laundering in financial
systems~\cite{petkovic2022fraud}, revealing multi-step fraudulent
behavior sequences in security event logs, identifying equipment
degradation patterns in IoT sensor streams, and tracking multi-step
user journeys in e-commerce clickstreams.  Unlike conventional SQL
predicates that evaluate each row in isolation, pattern matching
takes the temporal ordering and relative values of successive rows
into account.  Patterns are described using a regular-expression-like
syntax applied at the row level, enabling analysts to match and
extract meaningful sequences across ordered partitions.

% ----------------------------------------------------------------
\section{Problem Statement}
\label{sec:intro-problem}
% ----------------------------------------------------------------

The \texttt{MATCH\_RECOGNIZE} (MR) clause was introduced in the
SQL:2016 standard~\cite{iso2016,ISO2021}, enabling users to define
and search for patterns within ordered sequences of rows directly
in SQL queries.  Although standardized in 2016, only
a limited number of database management systems have implemented it.
Among established relational database systems, Oracle~\cite{oracle2022,book}
was the first to provide a production-quality implementation (in
Database~12c, 2013), providing an early reference implementation
for table-expression row pattern recognition.
Several modern analytical platforms followed: Trino~\cite{trino2022,trino2021row},
Snowflake~\cite{snowflake2025match_recognize}, and Apache Flink~\cite{flink}
each provide their own implementations, though with varying degrees
of conformance to the SQL:2016 feature levels (R010, R020, R030)
and with important practical restrictions.  Notably, PostgreSQL
and DuckDB---two widely used open-source systems---do not yet
support \texttt{MATCH\_RECOGNIZE}~\cite{howe2020matchrecognize,findling2025bringing}.

This gap is especially important for local analytical workflows.
Many users prepare data, test hypotheses, and build downstream
models directly in Python before moving to larger database or
cloud systems.  For data that fits comfortably in memory, requiring
a separate SQL engine only to express row-pattern logic can add
deployment and data-movement overhead.  The focus of this thesis is
therefore not to replace enterprise or streaming systems, but to
bring a practical subset of declarative row-pattern recognition into
the Pandas environment where many analytical workflows already
begin.

Without this declarative clause, detecting multi-row patterns
requires complex combinations of SQL window functions, self-joins,
or procedural loops over rows outside the database.  As the pattern
grows beyond two or three steps, the corresponding code expands
into dozens of lines filled with row-number calculations, timestamp
comparisons, and cascading \texttt{JOIN} conditions.  Such code
is difficult to write, harder to maintain, and typically forces
multiple table scans with poor
performance~\cite{cao2012optimizationanalyticwindowfunctions}.

In Python's data science ecosystem, the limitation is even more
acute.  Pandas~\cite{McKinney2010} is a widely used library for
in-process tabular data analysis, yet it provides no native
sequential pattern primitive.  Analysts who need to detect
multi-step patterns must manually compose low-level operations:
creating multiple shifted column copies (\texttt{df['value'].shift(-1)},
\texttt{shift(-2)}, \ldots), building Boolean masks, stitching
subsequences with \texttt{groupby().apply()}, and pivoting or
renaming columns.  Listing~\ref{lst:ch1-udf} illustrates this
approach for a simple three-step decreasing-price pattern (\texttt{A~B~C}).

\begin{lstlisting}[language=Python,
    caption={Detecting a three-step price-decrease pattern in
             Pandas \emph{without} \texttt{MATCH\_RECOGNIZE}},
    label={lst:ch1-udf}]
import pandas as pd

df['order_date'] = pd.to_datetime(df['order_date'])
df = df.sort_values(['customer_id', 'order_date']).reset_index(drop=True)

def match_pattern_ABC(group: pd.DataFrame):
    matches = []
    for i in range(len(group) - 2):
        A, B, C = group.iloc[i], group.iloc[i+1], group.iloc[i+2]
        if (B['price'] < A['price']) and (C['price'] < B['price']):
            matches.append({
                'customer_id':  A['customer_id'],
                'start_date':   A['order_date'],
                'end_date':     C['order_date'],
                'bottom_price': C['price']
            })
    return matches

all_matches = []
for _, grp in df.groupby('customer_id'):
    all_matches.extend(match_pattern_ABC(grp))

result_df = pd.DataFrame(all_matches)
\end{lstlisting}

This hand-crafted approach is verbose (over 20 lines for a
three-step pattern), prone to off-by-one alignment errors,
and does not scale gracefully to longer or conditional patterns.
The same query expressed with \texttt{MATCH\_RECOGNIZE} is shown
in Listing~\ref{lst:ch1-mr}: a single declarative statement that
specifies the partition, ordering, pattern grammar, variable definitions, and output measures.

\begin{lstlisting}[language=Python,
    caption={The same pattern expressed with \texttt{MATCH\_RECOGNIZE}
             on a Pandas DataFrame},
    label={lst:ch1-mr}]
import pandas as pd
from pandas_match_recognize import match_recognize

df['order_date'] = pd.to_datetime(df['order_date'])

query = """
SELECT customer_id, start_date, end_date, bottom_price
FROM orders
MATCH_RECOGNIZE (
  PARTITION BY customer_id
  ORDER BY     order_date
  MEASURES
    A.order_date AS start_date,
    C.order_date AS end_date,
    C.price      AS bottom_price
  ONE ROW PER MATCH
  AFTER MATCH SKIP TO NEXT ROW
  PATTERN (A B C)
  DEFINE
    B AS B.price < A.price,
    C AS C.price < B.price
)
"""

output_df = match_recognize(query, df)
\end{lstlisting}

The declarative form is not merely shorter syntactically: it
separates what to find (the pattern) from how to
find it (the engine), making patterns composable, auditable, and
easy to extend.  A more complex pattern, such as four or five steps
with conditional alternation, requires no change to the surrounding
Python code, only an update to the \texttt{PATTERN} and
\texttt{DEFINE} clauses~\cite{michels2018new,petkovic2022rpr}.
It also allows the implementation to compile the row pattern into
automata, cache repeated pattern structures, and process ordered
partitions without exposing this execution logic to the user.

The full structure and sub-clauses of \texttt{MATCH\_RECOGNIZE}
are introduced in Chapter~\ref{chap:background}.  Chapter~\ref{chap:related}
surveys the existing implementations and identifies the research
gap.

% ----------------------------------------------------------------
\section{Research Objectives and Contributions}
\label{sec:intro-objectives}
% ----------------------------------------------------------------

This thesis addresses the gap between the SQL:2016
\texttt{MATCH\_RECOGNIZE} standard and Python's Pandas ecosystem,
bridging the expressiveness of relational pattern queries with the
flexibility of in-memory analytics.  Three objectives guided the
work; each is stated below together with the contribution that
fulfills it.

\begin{enumerate}

  \item \textbf{An in-process engine supporting core SQL:2016
    Feature~R010-style capabilities for Pandas DataFrames.}
    Without a declarative alternative in Python, a data scientist
    must write the twenty-line hand-crafted loop in
    Listing~\ref{lst:ch1-udf} merely to detect a simple
    three-step declining-price pattern.  This thesis delivers
    an in-process \texttt{MATCH\_RECOGNIZE}-style implementation
    supporting the core Feature~R010 behavior evaluated in this
    thesis.  It runs entirely alongside Pandas---no external
    database server or cloud subscription required---reducing that
    loop to a single declarative call in Listing~\ref{lst:ch1-mr}
    and eliminating the need for complex Python code or external
    SQL-engine dependencies.

  \item \textbf{Broad evaluated coverage of SQL:2016-style pattern
    constructs.}
    The engine implements and evaluates a broad set of pattern
    constructs: quantifiers (\texttt{*}, \texttt{+}, \texttt{?},
    \texttt{\{n,m\}}), alternation (\texttt{|}), grouping,
    anchors, exclusions (\texttt{\{- -\}}), and \texttt{PERMUTE}.
    For example, a fraud analyst can express a probe-and-drain
    sequence as

    \medskip
    \texttt{PATTERN (PROBE+ (TRANSFER | WITHDRAW)\{2,5\} DRAIN)}
    \medskip

    \noindent
    and the engine parses, compiles, and executes it.  Functional
    correctness is assessed through targeted automated tests
    covering the supported features.  One acknowledged boundary:
    patterns with three or
    more levels of nested greedy quantifiers
    (e.g., \texttt{(A+B*)+C?}) can trigger exponential
    automata state-space growth; bounded quantifiers and simpler
    nesting behave efficiently~\cite{michels2018new}.

  \item \textbf{Safe predicate evaluation and empirical
    validation at a real-world scale.}
    User-written \texttt{DEFINE} predicates must be evaluated
    \emph{securely}: passing strings directly to Python's
    \texttt{eval()} is a code-injection vulnerability (a predicate
    such as \texttt{\_\_import\_\_('os').system('echo hacked')}
    would execute an arbitrary system command).  The engine instead parses
    predicates into an AST and evaluates them through a controlled
    interpreter that allows only permitted operations.  Correctness
    is assessed through targeted automated tests, while performance
    is evaluated using the Amazon UK Products~2023
    catalog~\cite{asaniczka2023amazon}.  The performance study
    uses five pattern families at five input sizes
    (25K, 35K, 50K, 75K, and 100K rows) sampled from the
    2.2-million-row catalog; all 25 performance runs completed
    successfully, with throughput up to approximately
    \textbf{12\,799~rows/sec} on a commodity laptop.
    Future work will address performance on larger datasets,
    memory consumption for high-variable patterns, and integration
    with query-optimizer plan-level improvements.

\end{enumerate}

% ----------------------------------------------------------------
\section{Organization of the Thesis}
\label{sec:thesis-organization}
% ----------------------------------------------------------------

The remainder of this thesis is organized as follows.

\textbf{Chapter~\ref{chap:background} --- Background and Concepts}
presents the theoretical and technical foundations on which this work is built.  It introduces sequential pattern matching and its
real-world applications, explains the SQL:2016
\texttt{MATCH\_RECOGNIZE} clause and its eight sub-clauses in
detail, and introduces the SQL:2016 feature-level classification
(R010, R020, R030)
used throughout the thesis.

\textbf{Chapter~\ref{chap:related} --- Related Work}
surveys each major \texttt{MATCH\_RECOGNIZE} implementation in
depth---Oracle, Trino, Snowflake, Apache Flink, and DuckDB---documenting
their documented feature scope, execution strategies, deployment
models, and practical
restrictions.  It also examines the Python and Pandas ecosystem
gap and formally positions the contribution of this thesis within
the broader pattern-matching research landscape.

\textbf{Chapter~4 --- System Design and Methodology}
describes the overall system architecture and methodology.  It
covers the SQL parsing pipeline, grammar and AST design, semantic
validation, logical plan construction, and the design choices that
connect declarative \texttt{MATCH\_RECOGNIZE} queries to the
execution layer.

\textbf{Chapter~5 --- Implementation and Optimization}
describes the concrete implementation of the engine.  It covers
finite-automata compilation, NFA and DFA construction, Pandas-based
execution, output materialization, measure evaluation, and
performance optimizations such as pattern caching, memory
management, and enhanced condition evaluation.

\textbf{Chapter~6 --- Experimental Evaluation}
presents the evaluation methodology, the benchmark dataset
(Amazon UK Products 2023, 2.2~million rows), and experimental
results covering functional correctness, performance across five
pattern families and five input sizes, scalability, pattern
complexity, and comparison with existing systems.

\textbf{Chapter~7 --- Conclusions and Future Work}
summarizes the main contributions, reflects on the limitations
of the current implementation, and outlines directions for
future research and further extensions.

\section{Publications in the Course of this Thesis}

This thesis has resulted in scientific outputs contributing to the
area of data systems and sequence-based pattern recognition.  The
following items summarize the published and submitted work produced
during the course of this research.

\begin{description}

\item[First Publication.]
The primary outcome of this thesis has been published in the
\textit{2026 IEEE 23rd Mediterranean Electrotechnical Conference
(MELECON)}.  The paper, entitled ``Row Pattern Recognition for
Pandas: Bringing MATCH\_\allowbreak RECOGNIZE to Python DataFrames''
\cite{lawande2026rpr}, addresses the absence of native row pattern
recognition (RPR) support in Python data analysis frameworks.  It
proposes an SQL-in-Pandas engine that enables the execution of
\texttt{MATCH\_RECOGNIZE}-like queries directly on \texttt{pandas}
DataFrames.  The system supports key SQL:2016 RPR constructs and
uses a non-deterministic finite automaton (NFA)-based matching
approach for processing sequential patterns.  Experimental results
demonstrate reduced implementation complexity compared with
conventional \texttt{pandas}-based workflows while maintaining
correctness across the evaluated cases.

\item[Second Publication.]
A second manuscript extending this work has been submitted to
\textit{Scientific Reports} and is currently under review.

\end{description}

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:intro-summary}
% ----------------------------------------------------------------

This chapter introduced the research problem addressed by the
thesis: the absence of a native, declarative row-pattern recognition
primitive in Pandas.  It explained why hand-written Pandas solutions
for multi-row patterns are verbose and error-prone, and it stated
the main objectives and contributions of the work.

The chapter also outlined the structure of the thesis and listed the
research outputs produced during the project.  The next chapter
provides the background needed to understand SQL row-pattern
recognition and the \texttt{MATCH\_RECOGNIZE} clause.
