% ============================================================
%  ch3_related_work.tex
%  Chapter 3: Related Work
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% !TEX root = Thesis.tex
\chapter{Related Work}
\label{chap:related}

This chapter positions the thesis within existing work on SQL
row-pattern recognition and Python-based tabular analytics.
Section~\ref{sec:related-native} reviews systems that provide native
\texttt{MATCH\_RECOGNIZE} or related row-pattern support.
Section~\ref{sec:related-pandas-gap} discusses the gap in Pandas and
similar DataFrame libraries.  Section~\ref{sec:related-contribution}
summarizes how this thesis addresses that gap.

Row pattern recognition was first proposed as a SQL standard feature
by Zemke et al.~\cite{zemke2007pattern} and was officially adopted
in SQL:2016~\cite{ISO2016SQL,iso2016}.
Subsequent editions refined and documented its
semantics~\cite{ISO2021,melton2017sqlrpr,michels2018new}.
The clause has since been studied both as a specification and in
terms of its practical
implementations~\cite{petkovic2022rpr,petkovic2022fraud}.

The \texttt{MATCH\_RECOGNIZE} clause is supported natively by
several SQL and stream-processing engines, including
Oracle~\cite{oracle2022,book},
Snowflake~\cite{snowflake2025match_recognize}, and
Trino~\cite{trino2022,trino2021row}; Apache Flink also exposes row
pattern recognition through Flink SQL~\cite{flink}.
By contrast, widely used systems such as PostgreSQL and DuckDB do
not currently expose a native \texttt{MATCH\_RECOGNIZE}
implementation~\cite{howe2020matchrecognize,findling2025bringing}.
In systems without native support, users typically emulate
row-pattern matching using window functions, recursive queries,
user-defined aggregates, or external procedural
layers~\cite{TURP2,TURP,cao2012optimizationanalyticwindowfunctions}.
These alternatives can express many of the same ideas, but they are
usually more verbose and harder to maintain than a declarative row
pattern clause.

% ----------------------------------------------------------------
\section{SQL and Stream-Processing Engines with Native Support}
\label{sec:related-native}
% ----------------------------------------------------------------

Oracle~\cite{oracle2022,book}, Trino~\cite{trino2022,trino2021row},
Snowflake~\cite{snowflake2025match_recognize}, and
Flink~\cite{flink} all provide SQL-based row pattern recognition,
but they differ in feature coverage, execution environment, and
practical deployment cost.
Although \texttt{MATCH\_RECOGNIZE} was already defined as part of the
SQL:2016 standard, it has only been implemented in a limited number
of DBMSs to date~\cite{petkovic2022rpr}.
Given the construct's ability to express sequence logic that would
otherwise require verbose combinations of joins, window functions,
or procedural code, it remains useful to extend row-pattern support
to more analytical environments.
Oracle remains a major reference implementation for SQL row pattern
matching~\cite{oracle2022,oracle2021dwhsg,book}.

% ................................
\subsection{Comparative Overview of Existing Implementations}
\label{subsec:related-existing-overview}
% ................................

The SQL standard defines three feature profiles for classifying row
pattern recognition implementations~\cite{ISO2021,michels2018new}.
\textbf{Feature~R010} covers row pattern recognition in the
\texttt{FROM} clause together with basic aggregate functions;
\textbf{Feature~R020} additionally permits row pattern recognition
inside \texttt{WINDOW}/\texttt{OVER} clauses; and
\textbf{Feature~R030} enables extended aggregate support within the
clause.  In practice, Feature~R010 is the common baseline, while
support beyond R010 varies across engines.  Current Trino
documentation, for example, also describes row pattern recognition
inside window structures~\cite{trino2022}.  Feature~R030 is not
widely adopted in publicly documented implementations.

Table~\ref{tab:engine-comparison} summarizes the main systems
considered in this chapter alongside the implementation developed in
this thesis.  The table uses the SQL feature levels where they map
cleanly to public documentation; otherwise, it describes the
documented implementation scope more cautiously.  The comparison is
therefore based on documented feature scope, deployment model, and
research relevance, rather than on a claim that all systems expose
identical SQL semantics.

\begin{table}[htbp]
  \centering
  \caption{Comparison of \texttt{MATCH\_RECOGNIZE} implementations}
  \label{tab:engine-comparison}
  \small
  \setlength{\tabcolsep}{3pt}
  \begin{tabularx}{\textwidth}{l>{\raggedright\arraybackslash}Xcll}
    \toprule
    \textbf{Engine} & \textbf{Documented scope} & \textbf{Open Source} & \textbf{Deployment} & \textbf{License} \\
    \midrule
    Oracle       & \texttt{FROM}-clause row patterns & No  & On-premise / Cloud & Commercial  \\
    Trino        & \texttt{FROM} + window row patterns & Yes & Distributed        & Apache 2.0  \\
    Snowflake    & \texttt{FROM}-clause row patterns + extensions & No  & Cloud-native       & Commercial  \\
    Apache Flink & SQL subset / streaming CEP-backed & Yes & Streaming cluster  & Apache 2.0  \\
    DuckDB       & No native implementation; research translation & Yes & In-process         & MIT         \\
    \midrule
    \textbf{This work} & \textbf{Core R010-style capabilities evaluated here} & \textbf{Yes} & \textbf{In-process Python} & \textbf{Open-source} \\
    \bottomrule
  \end{tabularx}
\end{table}

% ................................
\subsection{Oracle}
\label{subsec:related-oracle}
% ................................

Oracle has been a pioneer in SQL \texttt{MATCH\_RECOGNIZE}, having
introduced row pattern recognition in Database~12c.  It provides a
mature implementation for enterprise-scale relational analytics and
serves as a major reference point in discussions of SQL row pattern
recognition~\cite{oracle2022,oracle2021dwhsg,book}.  Publicly
documented usage focuses on the \texttt{MATCH\_RECOGNIZE} clause in
the table expression context; the window-pattern feature level
(R020) is therefore outside the documented Oracle usage considered
in this thesis.  Oracle also imposes implementation-specific
restrictions, including limitations on Boolean expressions in
\texttt{MEASURES} and restrictions around aggregate syntax.
Despite this technical maturity, Oracle's commercial deployment model
and operational requirements make it less accessible for users seeking
lightweight or open-source experimentation.

% ................................
\subsection{Trino}
\label{subsec:related-trino}
% ................................

Trino is an open-source distributed query engine designed for
parallel, federated query execution across large data
sources~\cite{trino2022,trino2021row}.  Its current documentation
covers \texttt{MATCH\_RECOGNIZE} in the \texttt{FROM} clause and
also documents row pattern recognition in window structures.  This
makes Trino one of the broader open-source implementations from a
SQL-feature perspective, but it is still a distributed query engine
rather than an in-process DataFrame library.  Using it therefore
requires server or cluster deployment and operational management that
may be unnecessary for smaller Pandas workflows.  In this thesis,
Trino is useful primarily as a semantic reference point for
validating selected query results, not as a substitute for a local
DataFrame-native execution path.

% ................................
\subsection{Snowflake}
\label{subsec:related-snowflake}
% ................................

Snowflake is a cloud-native data platform.  It provides documented
\texttt{MATCH\_\allowbreak RECOGNIZE} support with rich
\texttt{DEFINE} and \texttt{MEASURES} expression capabilities for
row-pattern queries~\cite{snowflake2025match_recognize,hoffa2021snowflake}.
Its documentation also notes implementation-specific behavior: the
engine uses backtracking for pattern matching and warns that some
pattern/data combinations can take a long time to execute.  Some
row-pattern functions are also restricted in particular sub-clauses.
Because Snowflake is a managed cloud warehouse, computational cost
scales with warehouse size and query workload.

% ................................
\subsection{Apache Flink}
\label{subsec:related-flink}
% ................................

Apache Flink~\cite{vsprem2024building,georgiou2025streaming} is
designed for real-time stream processing and exposes
\texttt{MATCH\_\allowbreak RECOGNIZE} through Flink SQL, backed by
its Complex Event Processing (CEP) runtime library~\cite{flink}.
Its documented implementation is a SQL subset oriented toward
streaming use cases.  The current Flink documentation explicitly
describes the implementation as a subset of the full standard and
lists unsupported features including pattern groups, some
alternations, \texttt{PERMUTE}, anchors, exclusions,
\texttt{ALL ROWS PER MATCH}, \texttt{SUBSET}, physical offsets
(\texttt{PREV}/\texttt{NEXT}), and \texttt{DISTINCT}
aggregations.  While Flink is highly scalable and effective for
high-velocity data, it assumes a streaming runtime and introduces
its own learning curve for users accustomed to lightweight,
batch-oriented analysis.
For analytical workflows that combine pattern extraction with
machine learning, these systems may also require additional
integration work between SQL execution and Python-oriented modeling
tools~\cite{ML_Workflow,Makrynioti2019sql4mlAD}.  In many practical
pipelines, this can introduce data movement, duplicated
intermediate data, or more complicated orchestration
steps~\cite{naphade2025evolution}.

% ................................
\subsection{Bringing \texttt{MATCH\_RECOGNIZE} to DuckDB}
\label{subsec:related-duckdb}
% ................................

DuckDB is a relational database management system that offers
several advantages over the DBMSs discussed
above~\cite{raasveldt2019duckdb}.
The most significant advantage is that it does not require a
separate server for operation; instead it can be embedded directly
into the host application, making it well-suited for small projects
and in-process analytics.
DuckDB is open-source, enabling users to contribute new
functionality.  Its Python API also makes it convenient to test and
debug generated SQL from translation-based
approaches~\cite{raasveldt2019duckdb}.

Although \texttt{MATCH\_RECOGNIZE} has not yet been implemented in
DuckDB natively, DuckDB supports relevant SQL building blocks such
as window functions and recursive common table expressions.  This
makes it a natural target for research that transpiles
\texttt{MATCH\_RECOGNIZE} queries into standard relational
constructs~\cite{findling2025bringing,mogensen2024compiler,TURP2,TURP}.
Such approaches typically use multiple CTEs and rule-based
rewriting to simulate pattern variables, ordering, skip behavior,
and output measures.  A DuckDB-oriented translation approach would
therefore aim to cover core Feature~R010 use cases without requiring
a native execution operator inside DuckDB itself.

% ................................
\subsection{Implementation Restrictions Across Systems}
\label{subsec:related-restrictions}
% ................................

Existing implementations of row pattern recognition impose
implementation-specific restrictions during use.  Examples reported
across systems~\cite{oracle2022,oracle2021dwhsg,trino2022,flink,snowflake2025match_recognize,petkovic2022rpr}
include:

\begin{itemize}
  \item supported contexts differ, with some systems documenting
    row patterns only in the \texttt{FROM} clause and others also
    documenting window-oriented row pattern recognition;
  \item the availability of pattern-recognition functions such as
    \texttt{CLASSIFIER()} and \texttt{MATCH\_NUMBER()} differs
    between \texttt{MEASURES}, \texttt{DEFINE}, and output modes;
  \item aggregate support differs across engines, including
    restrictions on \texttt{DISTINCT} aggregates and on expressions
    used as aggregate arguments;
  \item support for empty matches, exclusions, reluctant
    quantifiers, and skip modes is not uniform across all systems.
\end{itemize}

For many users---especially those with modest analytics needs or who
work independently---installing and managing large-scale systems
like Oracle, Trino, Snowflake, or Flink can be overly complicated.
These users may only need occasional pattern detection and may prefer
to avoid the overhead of learning different functions or dealing
with varying syntax for similar queries across
platforms~\cite{8509400}.  Table~\ref{tab:engine-comparison}
summarizes the key characteristics and documented scope of the
systems discussed above.

Additionally, for datasets that already fit comfortably in memory,
Pandas can offer a lower-friction workflow than deploying a separate
database server, cluster, or cloud warehouse~\cite{Evaluation,in-memory_processing}.
This does not replace the
strengths of Oracle, Trino, Snowflake, or Flink for large-scale or
streaming workloads; rather, it highlights a different use case:
lightweight, local, Python-centered analysis where data preparation,
pattern extraction, and downstream modeling can remain in the same
environment.

% ................................
\subsection{Synthesis of the Research Gap}
\label{subsec:related-gap-synthesis}
% ................................

The reviewed systems show that row-pattern recognition is a mature
and useful abstraction, but they also reveal a deployment and
workflow gap.  Oracle and Snowflake provide powerful managed or
commercial database environments; Trino provides a distributed
open-source SQL engine; and Flink targets streaming workloads.  These
systems are appropriate when data already resides in their execution
environment or when the workload requires distributed processing.
They are less convenient when the analyst is working with a Pandas
DataFrame that already fits in memory and wants to remain inside a
Python notebook, script, or experimental pipeline.

The research problem addressed in this thesis is therefore not the
absence of \texttt{MATCH\_RECOGNIZE} in SQL engines in general.
Rather, it is the absence of a Pandas-native row-pattern interface
that preserves the main declarative structure of SQL row-pattern
recognition while avoiding a round trip through an external database
or stream-processing system.  This distinction defines the scope of
the contribution: the proposed engine targets in-process, offline
DataFrame analytics and evaluates core R010-style behavior and
selected SQL:2016 pattern constructs.  It is not intended to replace
distributed engines for large-scale or streaming workloads.

% ----------------------------------------------------------------
\section{Pandas and the Python Ecosystem Gap}
\label{sec:related-pandas-gap}
% ----------------------------------------------------------------

Pandas is a popular, open-source Python library for data analysis
and manipulation~\cite{McKinney2010}.  Its straightforward installation process, broad
community adoption, and integration with the Python data-science
ecosystem make it attractive for students, researchers, and
practitioners who need to analyze tabular data without deploying a
separate database system.  Pandas provides useful primitives for
sorting, filtering, grouping, rolling windows, shifting columns, and
basic time-series analysis.

These primitives are powerful, but they are low-level building
blocks rather than a declarative row-pattern language.  Expressing a
multi-step sequence usually requires combinations of
\texttt{groupby}, \texttt{shift}, \texttt{rolling}, custom Python
loops, or intermediate Boolean columns.  Such implementations become
verbose and error-prone as patterns grow longer, require overlap
control, or involve conditions over previously matched rows.

Despite this broad adoption, Pandas does not provide
a native SQL:2016 \texttt{MATCH\_\allowbreak RECOGNIZE}-style primitive for row
pattern recognition.  In this setting, analysts often need to
manually implement stateful logic using
\texttt{DataFrame.\allowbreak apply()} chains, \texttt{iterrows()}, or external
database round-trips---all of which are error-prone, verbose, and
difficult to maintain.  Within the scope of this thesis, alternative
DataFrame libraries such as Dask, Polars, and Modin similarly lack a
direct equivalent to SQL row pattern recognition.  Even DuckDB, which
supports window functions and in-process analytics~\cite{findling2025bringing},
does not yet expose a native \texttt{MATCH\_RECOGNIZE} interface.

% ----------------------------------------------------------------
\section{Contribution of This Work}
\label{sec:related-contribution}
% ----------------------------------------------------------------

This thesis addresses the identified gap with the following
contributions:

\begin{enumerate}
  \item A \textbf{Pandas-native \texttt{MATCH\_RECOGNIZE} engine}
    supporting core SQL:2016 Feature~R010-style capabilities in an
    in-process Python workflow,A separate database server or cloud warehouse is not required.
  \item \textbf{Broad pattern construct support}, including
    quantifiers, alternation, grouping, anchors, exclusions, and
    \texttt{PERMUTE}, following the main row-pattern constructs used
    by SQL:2016 \texttt{MATCH\_RECOGNIZE}.
  \item An \textbf{optimization layer} comprising
    LRU pattern caching, partition-wise memory management, and
    thread-safe access to shared cache components.
  \item A \textbf{safe expression evaluation pipeline} that
    reduces code-injection risk by parsing user-defined predicates
    into a controlled representation rather than executing raw
    Python expressions.
  \item \textbf{Empirical validation} using targeted functional
    tests and experiments on samples from the Amazon UK
    Products~2023 catalog, with the detailed methodology and
    results reported in Chapter~\ref{chap:evaluation}.
\end{enumerate}

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:related-summary}
% ----------------------------------------------------------------

This chapter reviewed the main systems that currently support, or
partially support, SQL row-pattern recognition.  Oracle, Trino,
Snowflake, and Apache Flink provide useful reference points, while
DuckDB and Pandas illustrate the current gap for lightweight
in-process analytics.

The chapter showed that existing alternatives either require a
separate database or stream-processing engine, or force analysts to
write procedural logic in Python.  The resulting gap is a scoped but
important one: declarative row-pattern recognition for in-memory
Pandas workflows.  This motivates the system design presented in
Chapter~\ref{chap:methodology}.

% ................................................................
% END OF CHAPTER
% ................................................................
