% ============================================================
%  ch4_System Design and Methodology.tex
%  Chapter 4: System Design and Methodology
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% Provide \mr if not already defined by the main document
\providecommand{\mr}{\texttt{MATCH\_RECOGNIZE}}
\chapter{System Design and Methodology}
\label{chap:methodology}

The \texttt{MATCH\_RECOGNIZE} implementation follows a modular,
five-stage architecture specifically designed for SQL row-pattern
recognition.  The engine executes SQL queries directly on
\texttt{pandas} DataFrames~\cite{monierashraf2025rowmatch} while
preserving the supported SQL:2016 row-pattern semantics~\cite{iso2016,ISO2021,melton2017sqlrpr}.
The pipeline consists of five phases: \emph{parsing},
\emph{semantic validation}, \emph{logical plan construction},
\emph{pattern compilation}, and \emph{execution}
(Fig.~\ref{fig:system-architecture}).

In the parsing stage, the SQL query is transformed into an abstract
syntax tree (AST) that captures both the \texttt{SELECT}/\texttt{FROM}
clauses and \texttt{MATCH\_RECOGNIZE} sub-clauses~\cite{tarazona2021antlr,parr2013definitive}.
Semantic validation then checks variable usage, predicates, skip
policies, output modes, and the consistency of referenced
row-pattern constructs.  The validation rules are scoped to the
features supported by the implementation: they enforce correct use of
navigation functions, consistent measure semantics, and predictable
handling of edge cases such as empty matches, partition boundaries,
and skip behavior.

The validated plan is translated into a logical plan composed of
modular operators for partitioning, sorting, matching, measuring,
and skipping.  From this plan, the system compiles a
nondeterministic finite automaton (NFA) with guarded transitions
that support quantifiers, alternation, and \texttt{PERMUTE}
constructs~\cite{salomaa1996nfa}.  The NFA is constructed using
Thompson's construction, extended for SQL:2016 semantics by
embedding predicates from the \texttt{DEFINE} clause when present,
organizing $\varepsilon$-transitions, and applying subset
construction with state deduplication to mitigate unnecessary state
growth in practical patterns.

Finally, execution is performed over ordered partitions using the
compiled automaton.  Matches are materialized with the correct
\texttt{RUNNING}/\texttt{FINAL} semantics.  The evaluation engine
manages context separation, applies navigation functions
(\texttt{FIRST}, \texttt{LAST}, \texttt{PREV}, \texttt{NEXT}) with
partition awareness, and computes aggregates such as \texttt{COUNT},
\texttt{SUM}, and \texttt{AVG} across pattern variables.  This
preserves the supported SQL:2016 behavior while producing results
directly as \texttt{pandas} DataFrames.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    node distance=0.7cm and 0.7cm,
    every node/.style={font=\small},
    phase/.style={rectangle, draw, rounded corners=3pt,
                  minimum width=2.8cm, minimum height=0.75cm,
                  align=center},
    artifact/.style={rectangle, draw, dashed, rounded corners=3pt,
                     minimum width=2.6cm, minimum height=0.6cm,
                     align=center, fill=gray!10},
    arrow/.style={->, thick}
  ]
    \node[phase, fill=blue!10] (parse) {SQL Parsing};
    \node[phase, fill=blue!10, right=of parse] (validate) {Semantic\\Validation};
    \node[phase, fill=blue!10, right=of validate] (plan) {Logical Plan\\Construction};
    \node[phase, fill=blue!10, right=of plan] (compile) {Pattern\\Compilation};
    \node[phase, fill=blue!10, right=of compile] (execute) {Pandas\\Execution};

    \node[artifact, below=0.65cm of parse] (ast) {AST};
    \node[artifact, below=0.65cm of validate] (checked) {Validated\\AST};
    \node[artifact, below=0.65cm of plan] (logical) {Logical\\Plan};
    \node[artifact, below=0.65cm of compile] (automata) {NFA/DFA};
    \node[artifact, below=0.65cm of execute] (dfout) {Output\\DataFrame};

    \draw[arrow] (parse) -- (validate);
    \draw[arrow] (validate) -- (plan);
    \draw[arrow] (plan) -- (compile);
    \draw[arrow] (compile) -- (execute);

    \draw[arrow] (parse) -- (ast);
    \draw[arrow] (validate) -- (checked);
    \draw[arrow] (plan) -- (logical);
    \draw[arrow] (compile) -- (automata);
    \draw[arrow] (execute) -- (dfout);
  \end{tikzpicture}
  \caption{Five-stage architecture of the Pandas
    \texttt{MATCH\_RECOGNIZE} engine}
  \label{fig:system-architecture}
\end{figure}

The remainder of this chapter is organized as follows.
Section~\ref{sec:meth-overview} provides a high-level overview of
the three primary runtime components.
Section~\ref{sec:meth-compiler} describes the SQL parsing pipeline
in detail.
Section~\ref{sec:meth-semantic} covers the semantic validation
checks applied after parsing.
Section~\ref{sec:meth-logical} presents the logical plan
construction step.  The following chapter continues from this
logical design into automata compilation, Pandas execution,
implementation details, and optimization techniques.

% ----------------------------------------------------------------
\section{System Architecture Overview}
\label{sec:meth-overview}
% ----------------------------------------------------------------

The engine's five-phase pipeline is designed so that each stage
produces a well-defined artifact consumed by the next.
The following three subsections give a brief description of the
primary runtime components.  This chapter then details the parsing,
validation, and logical-planning stages; Chapter~\ref{chap:implementation}
continues with automata compilation, Pandas execution, and
optimization.

% ................................
\subsection{SQL Parsing and AST Construction}
\label{subsec:overview-parsing}
% ................................

The process starts by analyzing the user's SQL query and identifying
its logical components.  A parsing process explores the query,
recognizes its main components---grouping (\texttt{PARTITION BY}),
ordering (\texttt{ORDER BY}), the pattern to be searched, and any
specific conditions or output metrics---and organizes these into an
Abstract Syntax Tree (AST).  The AST provides a clear, structured
representation of the query.  This approach eases query validation
and prepares it for pattern matching in the following stages.

% ................................
\subsection{Pattern Translation to Finite Automata}
\label{subsec:overview-automata}
% ................................

Once the AST is complete, the system focuses on the pattern
component of the SQL query.  The pattern is converted into a finite
automaton that defines a set of states and transitions for detecting
complex sequences or behaviors in the data.  The process begins by
constructing a non-deterministic finite automaton (NFA) from the
pattern and then, for improved efficiency, transforms it into a
deterministic finite automaton
(DFA)~\cite{georgiou2025streaming}.  The DFA eliminates transition
ambiguity, enabling the system to scan data sequences quickly and
accurately.  This translation handles a wide variety of pattern
features---repetitions, alternatives, and specific sequences---in a
unified and efficient way.  The resulting automaton acts as the core
mechanism for scanning and detecting patterns within the data.

% ................................
\subsection{Execution Engine on Pandas}
\label{subsec:overview-execution}
% ................................

The execution phase uses the compiled DFA to perform pattern
matching directly on input data represented as Pandas DataFrames.
The engine begins by partitioning and ordering the data according to
the query's specifications.  It then applies the DFA to scan the
dataset, identifying matching sequences according to the configured
pattern and skip policy.
For every detected match the engine computes any requested measures
or calculations and compiles the results into a DataFrame formatted
to mirror standard SQL query outputs.  By utilizing Pandas, the
system retains both flexibility and accessibility, smoothly
incorporating advanced pattern matching into conventional Python
data-analysis workflows.

% ----------------------------------------------------------------
\section{SQL Parsing}
\label{sec:meth-compiler}
% ----------------------------------------------------------------

Transforming a \texttt{MATCH\_RECOGNIZE} query from plain text into
an executable representation requires a parsing pipeline that
handles both standard SQL syntax and the extended row-pattern
sub-language.  This section explains parsing fundamentals, surveys
existing SQL parsers and parser generators, motivates the design
choices made in this implementation, and describes the resulting
parser in detail.

% ................................
\subsection{What is Parsing?}
\label{subsec:meth-what-parsing}
% ................................

\emph{Parsing} refers to the analysis of an input string according to the
rules of a formal grammar and its transformation into a
machine-processable representation~\cite{mogensen2024compiler}.  In essence,
the parser examines a human-written query, determines which grammar rules
govern its syntactic construction, and produces a tree data structure that
reflects that derivation.

Parsing in this engine proceeds in two sequential stages:
\emph{lexical analysis} and \emph{syntactic analysis}~\cite{DragonBook}.

\paragraph{Lexical analysis.}
The raw character stream is segmented into a sequence of predefined,
meaningful units called \emph{tokens}.  Keywords
(\texttt{SELECT}, \texttt{MATCH\_RECOGNIZE}, \texttt{PATTERN}, \ldots),
identifiers, string and numeric literals, and punctuation are all examples
of tokens.  Whitespace and comments are discarded at this stage; only the
ordered token sequence is passed to the next stage.

\paragraph{Syntactic analysis.}
The token stream is matched against the grammar rules of the language,
and a valid input is translated into a tree representation.  A valid
input produces a \emph{parse tree} (also called a \emph{concrete syntax
tree}) whose internal nodes correspond to grammar rules and whose leaves
correspond to individual tokens.  An invalid input causes the parser to
report an error, indicating precisely where the text violated the grammar.

The two-stage design separates concerns cleanly: the lexer handles
regular-language aspects (token shapes), while the parser handles
context-free-language aspects (phrase structure).  Together they convert
a character string into a structured tree that subsequent compiler phases
can traverse and transform.

% ................................
\subsection{Parsing SQL: Existing Solutions}
\label{subsec:meth-existing-parsers}
% ................................

Several Python libraries offer SQL parsing out of the box.
\texttt{sqlparse}~\cite{sqlparse_tool} performs tokenization and basic structural
splitting (statements, clauses) but produces only a shallow token
stream, not a typed AST, and does not model \texttt{MATCH\_RECOGNIZE}
at all.  \texttt{sqlglot}~\cite{sqlglot_tool} is a more complete SQL transpiler that
supports multiple SQL dialects and exposes a typed AST; however,
it does not provide project-ready coverage for the required
SQL:2016 row-pattern sub-language (e.g.\ exclusions, anchors,
\texttt{PERMUTE}, bounded quantifiers).  \texttt{mo\_sql\_parsing}~\cite{mo_sql_parsing_tool}
converts SQL to JSON but similarly provides no project-ready
coverage for row-pattern constructs.

A widely used baseline for conventional SQL parsing is the
\textbf{PostgreSQL parser}, which has been extracted from the
PostgreSQL source code and released as a standalone C library via
the \texttt{libpg\_query} project~\cite{libpg_query_tool}.  Built on top of
\texttt{libpg\_query}, the Python library \texttt{pglast}~\cite{pglast_tool} wraps the
original C AST nodes as typed Python classes and provides
functionality to reconstruct SQL text from an AST.
Listing~\ref{lst:pglast-example} shows a minimal usage example:
the call to \texttt{parse\_sql} returns a tree of typed node
objects, and \texttt{RawStream} serializes it back to SQL.

\begin{lstlisting}[language=Python,
    caption={Parsing a SQL statement with \texttt{pglast} and
             reconstructing SQL text},
    label={lst:pglast-example}]
from pglast import parse_sql
from pglast.stream import RawStream

tree_root = parse_sql("SELECT 42;")
print(tree_root)          # typed AST node objects
print(RawStream()(tree_root))  # SELECT 42
\end{lstlisting}

Although \texttt{pglast} cannot parse \texttt{MATCH\_RECOGNIZE}, it
illustrates the difference between a conventional SQL AST and the
row-pattern-specific structure required here.  The project therefore
does not depend on \texttt{pglast}, \texttt{sqlparse},
\texttt{sqlglot}, or \texttt{mo\_sql\_parsing} for execution.
Instead, these tools were considered as alternatives, and a
grammar-based parser was selected because the required syntax
includes pattern constructs that are not handled uniformly by
general-purpose Python SQL parsers.

% ................................
\subsection{Grammars}
\label{subsec:meth-grammars}
% ................................

A grammar formally describes the syntax of a language~\cite{DragonBook}.
The elementary constituents of a grammar are:

\begin{itemize}
  \item \textbf{Terminal symbols} (tokens) --- the elementary,
    indivisible symbols of the language.  In this thesis the terms
    \emph{terminal} and \emph{token} are used interchangeably.
  \item \textbf{Non-terminals} --- symbols that represent sets of
    strings composed of terminals, used to name syntactic constructs
    (e.g.\ \emph{expression}, \emph{pattern}, \emph{clause}).
  \item \textbf{Production rules} --- rewriting rules of the form
    $N \to \alpha$, where $N$ is a non-terminal and $\alpha$ is a
    sequence of terminals and/or non-terminals.  Each rule describes
    one way in which a particular construct of the language may be
    written~\cite{wagenknecht2022formal}.
\end{itemize}

A grammar is specified as a list of production rules.  Because the
left-hand side of each rule contains exactly one non-terminal, the
grammar is \emph{context-free}: the applicable rules do not depend
on the surrounding context~\cite{DragonBook}.  For example, a grammar
for one-or-more occurrences of the letter $a$ can be written:

\[
  S \to a \quad\mid\quad aS
\]

This form is also known as \emph{Backus--Naur Form} (BNF).
The \emph{Extended} variant (EBNF) adds syntactic sugar for
optional elements (\texttt{[~]}), repetition (\texttt{\{~\}}), and
grouping (\texttt{(~)}), making large grammars more readable.  For
example, the \texttt{PARTITION~BY} clause can be expressed in EBNF
as:

\begin{center}
  \texttt{partitionBy~:~'PARTITION'~'BY'~expr~(','\ expr)*~;}
\end{center}

SQL is a large context-free language; the \texttt{MATCH\_RECOGNIZE}
sub-language adds additional syntactic constructs for pattern terms,
quantifiers, alternation, grouping, anchors, exclusions,
\texttt{PERMUTE}, and the eight sub-clauses.

% ................................
\subsection{Parsing Requirements}
\label{subsec:meth-parsing-needs}
% ................................

This section examines what must be parsed and what the output
should look like.  Listing~\ref{lst:mr-full-syntax} shows the
supported syntax of a query containing a \texttt{MATCH\_RECOGNIZE}
clause in this implementation.

\begin{lstlisting}[
    language   = {},
    caption    = {Syntax of a \texttt{MATCH\_RECOGNIZE} query},
    label      = {lst:mr-full-syntax},
    basicstyle = \ttfamily\footnotesize
]
SELECT <expression> [, ...]
FROM   <table>
MATCH_RECOGNIZE (
  [ PARTITION BY <column> [, ...] ]
  [ ORDER BY <column> [ASC|DESC] [, ...] ]
  [ MEASURES <expression> AS <alias> [, ...] ]
  [ ONE ROW PER MATCH
  | ALL ROWS PER MATCH
    [ SHOW EMPTY MATCHES
    | OMIT EMPTY MATCHES
    | WITH UNMATCHED ROWS ] ]
  [ AFTER MATCH SKIP
    { PAST LAST ROW
    | TO NEXT ROW
    | TO FIRST <variable>
    | TO LAST  <variable> } ]
  PATTERN ( <row_pattern> )
  [ SUBSET <subset_name> = ( <variable> [, ...] ) [, ...] ]
  [ DEFINE <variable> AS <condition> [, ...] ]
) [ AS <alias> ]
<rest>;
\end{lstlisting}

The \texttt{DEFINE} clause is optional in both the grammar and the
execution pipeline.  If a pattern variable appears in
\texttt{PATTERN} without a corresponding predicate in
\texttt{DEFINE}, the logical plan assigns that variable an implicit
always-true predicate.
Measure aliases require the \texttt{AS} keyword in the adopted
Trino grammar, while the alias after the
\texttt{MATCH\_RECOGNIZE} clause itself is optional.

\texttt{MATCH\_RECOGNIZE} is an optional sub-clause of the
\texttt{FROM} clause.  The implementation focuses on one
\texttt{MATCH\_RECOGNIZE} clause extracted from a top-level
\texttt{SELECT} query and supports an optional outer
\texttt{ORDER~BY}.  More complex SQL compositions such as full CTE
rewriting, set-operation planning, and general nested-subquery
execution are outside the main execution path, although selected
subquery-like and multi-query scenarios are covered by the test
suite as foundations for future extension.  This scope is sufficient
for the core \texttt{FROM}-clause row-pattern use cases evaluated in
this thesis.

The parser must extract the mandatory pattern clause and the
available optional sub-clauses of
\texttt{MATCH\_RECOGNIZE}---\texttt{PARTITION~BY},
\texttt{ORDER~BY}, \texttt{MEASURES}, rows-per-match,
after-match-skip, \texttt{PATTERN}, \texttt{SUBSET}, and
\texttt{DEFINE}---and deliver them to subsequent pipeline stages
through a well-defined AST interface, described next.

% ................................
\subsubsection{The \texttt{MATCH\_RECOGNIZE} AST Interface}
\label{subsubsec:meth-ast-interface}
% ................................

The AST exposed by the parser to downstream stages is defined by a
hierarchy of typed Python \texttt{@dataclass} objects.  The root
node, \texttt{FullQueryAST}, holds the outer \texttt{SelectClause},
\texttt{FromClause}, and a single \texttt{MatchRecognizeClause}
object that groups the parsed row-pattern sub-clauses.  The interface
contract is:

\begin{itemize}
  \item Each node is represented as a typed Python dataclass or
    structured object.  Some metadata fields, especially those on
    \texttt{PatternClause}, are normalized after construction so that
    defined variables, implicit variables, and subset components can
    be resolved consistently.
  \item Each \texttt{MATCH\_RECOGNIZE} sub-clause maps to a
    dedicated typed dataclass
    (\texttt{PartitionByClause}, \texttt{OrderByClause},
     \texttt{MeasuresClause}, \texttt{RowsPerMatchClause},
     \texttt{AfterMatchSkipClause}, \texttt{PatternClause},
     \texttt{SubsetClause}, \texttt{DefineClause}),
    enabling the NFA compiler to dispatch on type rather than
    parsing raw strings.
  \item \texttt{DEFINE} predicates and \texttt{MEASURES}
    expressions are captured as expression strings during the first
    pass and interpreted on demand through a controlled expression
    evaluator (Section~\ref{subsec:meth-safe-eval}), keeping the main
    grammar concise.
  \item \texttt{Measure} nodes carry computed flags
    (\texttt{is\_classifier}, \texttt{is\_match\_number}) that
    allow the execution layer to route output columns without
    additional string inspection.
\end{itemize}

% ................................
\subsection{Parser Generators}
\label{subsec:meth-parser-generators}
% ................................

Since productivity is a goal, the question arises whether the
creation of a lexer and parser can be automated.  The kind of
software addressing this issue is called a
\emph{parser generator}~\cite{mogensen2024compiler}: it takes a
grammar as input and produces the source code of a parser that
understands the syntax defined by that grammar.  The target
language of the generated source code must be Python, since that
is the language the engine is written in.

Candidates that support Python as a target language include:

\begin{itemize}
  \item \textbf{PLY} (Python Lex-Yacc) --- a Python implementation
    of the classic 1970 Lex/Yacc tools~\cite{ply_tool}.
  \item \textbf{Lark} --- described as ``a modern approach, with a
    focus on ergonomics, performance and
    modularity''~\cite{lark_tool}.
  \item \textbf{ANTLR4} --- a parser generator ``widely used to
    build languages, tools, and frameworks; from a grammar, ANTLR
    generates a parser that can build and walk parse
    trees''~\cite{Parr2013}.
\end{itemize}

Table~\ref{tab:parser-gen-comparison} compares the three tools
across criteria extracted from their respective documentation.

\begin{table}[htbp]
  \centering
  \small
  \caption{Comparison of Python-compatible parser generators}
  \label{tab:parser-gen-comparison}
  \begin{tabular}{p{2.8cm} p{3.0cm} p{3.0cm} p{3.0cm}}
    \toprule
    \textbf{Feature}
      & \textbf{PLY (Lex/Yacc)}
      & \textbf{Lark}
      & \textbf{ANTLR4} \\
    \midrule
    Algorithm
      & LALR(1)
      & LALR / Earley
      & Adaptive LL(*) \cite{parr2014adaptive} \\[4pt]
    Grammar notation
      & BNF
      & EBNF
      & EBNF \\[4pt]
    Ambiguity handling
      & Operator precedence declarations
      & Rule priorities; Earley resolves via priority
      & Production order \& semantic predicates \\[4pt]
    Left recursion
      & Yes
      & Yes (LALR mode)
      & Yes \\[4pt]
    Target language
      & Python native
      & Python native
      & Java runtime required to generate Python parser \\[4pt]
    AST / parse tree
      & No
      & Yes (parse tree)
      & Yes (parse tree) \\[4pt]
    Tree visualiser
      & No
      & Dot output
      & Java GUI (\texttt{antlr4-tools}) \\[4pt]
    Traversal scaffolding
      & No
      & Transformer class; methods written manually
      & Listener \& visitor generated automatically \\
    \bottomrule
  \end{tabular}
\end{table}

Due to the manageable size and complexity of the
\texttt{MATCH\_RECOGNIZE} statement, any of the three tools would
likely produce a satisfactory result.  Nevertheless, two reasons
motivated the choice of ANTLR4.  First, its parse-tree visualiser
(via the \texttt{antlr4-tools} Java GUI) greatly aids debugging
grammar rules.  Second, and more crucially, ANTLR4 automatically
generates a traversal skeleton---both listener and visitor base
classes---for the parse tree.  Because the raw parse tree must be
transformed into the typed AST described in
Section~\ref{subsubsec:meth-ast-interface}, a generated visitor
base class eliminates manual tree-traversal boilerplate.  An
empirical study further demonstrates significantly higher
performance and maintainability when using ANTLR compared to
Lex/Yacc~\cite{antlr_performance}.

In this thesis the workflow combining ANTLR4 with the typed
\texttt{MATCH\_RECOGNIZE} AST interface is evaluated for its
suitability in parsing the supported \texttt{MATCH\_RECOGNIZE}
statement.

% ................................
\subsection{Alternative Methods}
\label{subsec:meth-alternatives}
% ................................

Two non-generator approaches were considered and rejected.

\paragraph{Hand-written recursive descent.}
A recursive descent parser is straightforward to implement for
small grammars: each non-terminal becomes a Python method that
calls other methods for sub-rules.  However, SQL's grammar contains
hundreds of productions; maintaining a hand-written parser as the
grammar evolves is error-prone.  The approach was ruled out on
maintainability grounds.

\paragraph{Regex-based string extraction.}
Regular expressions cannot model recursive structures such as nested
grouping \texttt{((A+B)*|C)} or balanced parentheses, so they
cannot serve as a complete parser.  They can pre-screen queries or
extract simple fragments, but any production use requires a proper
parser as the primary mechanism.

% ................................
\subsection{ANTLR4}
\label{subsec:meth-antlr4}
% ................................

\textbf{ANTLR4} (ANother Tool for Language Recognition, version~4)
is a parser generator that produces top-down, Adaptive LL(*)
recursive-descent parsers from context-free grammar specifications
written in the ANTLR meta-language~\cite{Parr2013,parr2014adaptive,antlr_website,parr_antlr_docs}.  Given a \texttt{.g4} grammar
file the tool generates a lexer, a parser, and listener/visitor
scaffolding in the target language (Python in this project).

The implementation uses Trino's SQL grammar
(\texttt{TrinoLexer.g4} / \texttt{TrinoParser.g4}) from the ANTLR
community grammars project.  This grammar already contains the
\texttt{patternRecognition} productions needed for the
\texttt{MATCH\_RECOGNIZE} sub-clauses, as shown in
Listing~\ref{lst:grammar}.  The project-specific layer is therefore
the Python visitor and validation logic that extracts these parse
tree nodes into typed AST objects.  A custom error listener
intercepts syntax errors and reports them with line/column positions
and contextual suggestions.

% ................................
\subsection{Implementation of the Parser}
\label{subsec:meth-parser-impl}
% ................................

% ................................
\subsubsection{Designing the Grammar}
\label{subsubsec:meth-grammar-design}
% ................................

The implementation uses the Trino SQL grammar obtained from the
ANTLR community grammars project
(\texttt{TrinoLexer.g4} / \texttt{TrinoParser.g4},
approximately 1\,521 lines across the two grammar files in this
project)~\cite{antlr_trino_sql}.
This grammar natively contains the \texttt{patternRecognition} rule
that covers all eight \texttt{MATCH\_RECOGNIZE} sub-clauses; no
modifications to the grammar file were required.
Listing~\ref{lst:grammar} shows the key productions.

\begin{lstlisting}[
    language = {},
    caption  = {Key productions of the \texttt{patternRecognition} rule in \texttt{TrinoParser.g4}},
    label    = {lst:grammar},
    basicstyle = \ttfamily\footnotesize
]
patternRecognition
    : aliasedRelation (
        MATCH_RECOGNIZE '('
          (PARTITION BY expression (',' expression)*)?
          (ORDER BY sortItem (',' sortItem)*)?
          (MEASURES measureDefinition (',' measureDefinition)*)?
          rowsPerMatch?
          (AFTER MATCH skipTo)?
          (INITIAL | SEEK)?
          PATTERN '(' rowPattern ')'
          (SUBSET subsetDefinition (',' subsetDefinition)*)?
          (DEFINE variableDefinition (',' variableDefinition)*)?
        ')' (AS? identifier columnAliases?)?
      )?
    ;

measureDefinition : expression AS identifier ;

rowsPerMatch
    : ONE ROW PER MATCH
    | ALL ROWS PER MATCH emptyMatchHandling?
    ;

skipTo
    : SKIP TO NEXT ROW
    | SKIP TO (FIRST | LAST)? identifier
    | SKIP PAST LAST ROW
    ;

rowPattern
    : rowPattern rowPattern             # patternConcatenation
    | rowPattern '|' rowPattern         # patternAlternation
    | patternPrimary patternQuantifier? # quantifiedPrimary
    ;

patternPrimary
    : identifier                                        # patternVariable
    | '(' rowPattern ')'                                # groupedPattern
    | '{-' rowPattern '-}'                              # excludedPattern
    | PERMUTE '(' rowPattern (',' rowPattern)* ')'      # patternPermutation
    ;
\end{lstlisting}

% ................................
\subsubsection{Handling Expressions as Strings}
\label{subsubsec:meth-expr-strings}
% ................................

\texttt{DEFINE} predicates and \texttt{MEASURES} expressions contain
column references, navigation function calls, arithmetic, comparisons,
and selected scalar or aggregate functions.  Parsing every expression
into a fully typed SQL expression tree would require a much larger
expression subsystem.  The adopted strategy is therefore to preserve
the relevant expression text during parsing and interpret it later
through the controlled expression evaluators used by the matching and
measure layers.  This two-pass strategy keeps the main grammar concise
while still avoiding direct execution of raw user expressions (see
Section~\ref{subsec:meth-safe-eval}).

% ................................
\subsubsection{Parse Tree vs.\ Abstract Syntax Tree}
\label{subsubsec:meth-ast}
% ................................

Parse trees produced by ANTLR4 contain many structural nodes
(keywords, punctuation) that are unnecessary for execution.  A
visitor traverses the parse tree and produces a compact
\emph{Abstract Syntax Tree} (AST) whose node types correspond
directly to semantic constructs~\cite{DragonBook,jones2003ast}.  The key node
types are summarised in Table~\ref{tab:meth-ast-nodes}.

\begin{table}[htbp]
  \centering
  \caption{Principal AST node types}
  \label{tab:meth-ast-nodes}
  \begin{tabular}{ll}
    \toprule
    \textbf{Node Type}                  & \textbf{Represents} \\
    \midrule
    \texttt{FullQueryAST}               & Root of the full query (SELECT + FROM + MR) \\
    \texttt{MatchRecognizeClause}        & Root of the \texttt{MATCH\_RECOGNIZE} sub-clause \\
    \texttt{PartitionByClause}           & \texttt{PARTITION BY} column list \\
    \texttt{OrderByClause}              & \texttt{ORDER BY} sort items (\texttt{SortItem}) \\
    \texttt{MeasuresClause}             & \texttt{MEASURES} list (\texttt{Measure} nodes) \\
    \texttt{RowsPerMatchClause}          & \texttt{ONE / ALL ROWS PER MATCH} mode \\
    \texttt{AfterMatchSkipClause}        & \texttt{AFTER MATCH SKIP} policy \\
    \texttt{PatternClause}              & \texttt{PATTERN} string with tokenizer \\
    \texttt{SubsetClause}               & \texttt{SUBSET} variable definitions \\
    \texttt{DefineClause}               & \texttt{DEFINE} list (\texttt{Define} nodes) \\
    \bottomrule
  \end{tabular}
\end{table}

The AST is validated for semantic correctness before pattern
compilation: unresolved pattern-variable references, invalid subset
definitions, incompatible output modes, and illegal pattern syntax are
detected and reported at this stage.  Pattern variables without an
explicit \texttt{DEFINE} predicate remain valid and are assigned an
implicit always-true predicate downstream.

% ................................
\subsection{Visitor Implementation}
\label{subsec:meth-visitor}
% ................................

ANTLR4 generates an abstract visitor base class from the grammar~\cite{robson2022designpatterns}.
The parsing module is structured as two cooperating visitor classes
built on the generated \texttt{TrinoParserVisitor} base.

\paragraph{\texttt{MatchRecognizeExtractor}.}
The inner visitor overrides \texttt{visitPatternRecognition}
and extracts the available sub-clauses in grammar order:
\texttt{PARTITION~BY} column expressions,
\texttt{ORDER~BY} sort items with direction and nulls handling,
\texttt{MEASURES} definitions (handling \texttt{RUNNING}/\texttt{FINAL}
prefix and alias via a dedicated \texttt{smart\_split()} helper that
correctly traverses nested parentheses and quoted identifiers),
rows-per-match mode, after-match-skip policy, the raw pattern string,
\texttt{SUBSET} definitions, and \texttt{DEFINE} variable--condition
pairs.  After extraction, the parser applies validation routines such
as \texttt{validate\_clauses}, \texttt{validate\_identifiers},
\texttt{validate\_pattern\_clause},
\texttt{validate\_pattern\_variables\_defined}, and
\texttt{validate\_function\_usage}.  These checks verify required
clauses, balanced pattern syntax, incompatible output-mode
combinations such as exclusions with \texttt{WITH UNMATCHED ROWS},
table-prefix misuse, skip targets, and malformed uses of supported
navigation or aggregate functions.  Pattern variables that appear in
\texttt{PATTERN} but not in \texttt{DEFINE} are intentionally allowed
and later receive the implicit always-true predicate required by the
standard.

\paragraph{\texttt{FullQueryExtractor}.}
The outer visitor handles the top-level \texttt{singleStatement}
rule.  It extracts the outer \texttt{SELECT} column list, \texttt{FROM}
table reference, and any trailing \texttt{ORDER~BY} clause via regex,
then locates the \texttt{PatternRecognitionContext} node through a
recursive tree walk and delegates to \texttt{MatchRecognizeExtractor}.

\paragraph{Public API.}
The entry point for all callers is
\texttt{parse\_full\_query(query:~str)} $\to$ \texttt{FullQueryAST}.
This function preprocesses the query string
(e.g.\ rewriting empty \texttt{IN~()} predicates to a sentinel value
to satisfy the grammar), constructs
\texttt{InputStream} $\to$ \texttt{TrinoLexer} $\to$ \texttt{CommonTokenStream} $\to$ \texttt{TrinoParser},
replaces the default ANTLR4 error listeners with a
\texttt{CustomErrorListener} that raises a structured
\texttt{ParserError} carrying exact line, column, and source snippet
on any syntax violation, and visits the parse tree with
\texttt{FullQueryExtractor} to produce the typed \texttt{FullQueryAST}.

% ................................
\subsection{Testing and Evaluation of the Parser}
\label{subsec:meth-parser-testing}
% ................................

The parser implementation is validated through a structured test
suite located in the \texttt{tests/} directory.  Tests are organized
into parser-focused checks, integration tests that exercise the
\texttt{parse\_full\_query} pipeline, and SQL:2016-oriented tests
derived from reference scenarios and observed Trino behavior.

\paragraph{Parser-focused tests.}
The parser tests exercise the extraction and validation behavior of
the main \texttt{MATCH\_RECOGNIZE} sub-clauses, including
\texttt{MEASURES} expressions with \texttt{RUNNING}/\texttt{FINAL}
keywords, after-match skip targets, output-mode combinations, and
pattern syntax.  The validation chain is also tested
adversarially: supplying an undefined pattern-variable reference, a
prohibited function form in \texttt{MEASURES}, an invalid skip
target, or an incompatible output-mode combination triggers a
\texttt{ParserError} or an empty-result path as appropriate.  By
contrast, a pattern variable without a corresponding
\texttt{DEFINE} predicate is valid and receives an implicit
always-true condition.

\paragraph{Integration tests.}
The file \texttt{test\_match\_recognize.py} is adapted from Trino's
\texttt{TestRowPatternMatching.java} and exercises representative
end-to-end scenarios: simple queries, pattern quantifiers
(\texttt{*}, \texttt{+}, \texttt{?}, bounded \texttt{\{n,m\}}),
pattern exclusion syntax (\texttt{\{-\ldots-\}}), the supported
after-match-skip modes, one-row and all-rows output modes,
\texttt{CLASSIFIER()} and \texttt{MATCH\_NUMBER()} special columns,
navigation functions, partitioned and ordered input, \texttt{SUBSET}
aliasing, and \texttt{RUNNING}/\texttt{FINAL} semantics.

\paragraph{SQL:2016 compliance tests.}
\texttt{test\_sql2016\_compliance.py} contains targeted tests for
standard-oriented behavior: pattern syntax, navigation functions,
special aggregate functions, partition boundary handling, subset
variable resolution, output-mode semantics, and the supported skip
policies.  Empty-match handling is
covered separately in \texttt{test\_empty\_matches.py} with nine
test cases including \texttt{SHOW EMPTY MATCHES},
\texttt{OMIT EMPTY MATCHES}, \texttt{WITH UNMATCHED ROWS}, and
combinations with alternation and partitioning.

\paragraph{Edge-case tests.}
Additional test files cover \texttt{IN~()} predicate preprocessing
(\texttt{test\_in\_predicate.py}),
\texttt{IS~NULL} handling (\texttt{test\_is\_null\_handling.py}),
anchored patterns (\texttt{test\_anchor\_patterns.py}),
\texttt{PERMUTE} patterns (\texttt{test\_permute\_patterns.py}),
back-references (\texttt{test\_back\_reference.py}),
pattern cache correctness and thread safety
(\texttt{test\_pattern\_cache.py}), and aggregate-function behavior
(\texttt{test\_production\_aggregates.py}).

% ................................
\subsection{Discussion of the Parser}
\label{subsec:meth-parser-discussion}
% ................................

The grammar-driven approach provides several practical advantages:
the language specification remains in a single readable
\texttt{.g4} file; ANTLR4 provides structured parse-tree generation
and error-recovery mechanisms, while the custom listener reports
syntax errors with line and column context; and the generated
visitor scaffolding eliminates manual tree traversal boilerplate.

% ................................
\subsubsection{Limits}
\label{subsubsec:meth-parser-limits}
% ................................

Two limitations are known.  First, advanced SQL constructs outside
the supported row-pattern subset, such as recursive common table
expressions embedded inside predicate subqueries, are not part of the
main execution path; the expression subsystem recognizes the
constructs needed for \texttt{MATCH\_RECOGNIZE} evaluation.  Second,
because the grammar is based on Trino's dialect, vendor-specific
extensions of other systems (Oracle \texttt{RUNNING} keyword
placement differences, Snowflake aggregate variants) may require
minor grammar adjustments if cross-dialect compatibility is needed.

% ................................
\subsection{Safe SQL Expression Evaluation}
\label{subsec:meth-safe-eval}
% ................................

\texttt{DEFINE} predicates and \texttt{MEASURES} expressions are
user-supplied strings that must be interpreted at runtime.  Directly
calling \texttt{eval()} on arbitrary user input would create a
code-injection vulnerability.  The implementation instead translates
the supported SQL expression subset into controlled Python expression
forms and evaluates them through AST-aware evaluators with explicit
operator and function handling.  Listing~\ref{lst:safe-eval} shows
the simplified principle.

\begin{lstlisting}[
    language = Python,
    caption  = {Safe SQL expression evaluation using AST whitelisting},
    label    = {lst:safe-eval}
]
import ast

def safe_evaluate(sql_expr, row_data, context):
    # 1. Translate SQL syntax to Python equivalents
    py_expr = translate_sql_to_python(sql_expr)

    # 2. Parse into a Python AST (expression mode only)
    tree = ast.parse(py_expr, mode='eval')

    # 3. Whitelist-validate: allow comparisons, arithmetic,
    #    boolean ops, literals, subscripts -- forbid imports,
    #    function calls (except navigation functions), assignments
    ASTSecurityValidator().visit(tree)

    # 4. Evaluate by walking only the approved AST nodes
    evaluator = ControlledExpressionEvaluator(row_data, context)
    return evaluator.visit(tree.body)
\end{lstlisting}

The runtime evaluator follows the same principle but uses the
project's \texttt{ConditionEvaluator} and measure-evaluation helpers:
it walks supported AST nodes explicitly and exposes only the current
row context, pattern-variable bindings, supported navigation
functions (\texttt{PREV}, \texttt{NEXT}, \texttt{FIRST},
\texttt{LAST}), and selected scalar or aggregate operations.  It does
not expose Python built-ins, imports, file-system access, network
access, or interpreter internals.


% ----------------------------------------------------------------
\section{Semantic Validation}
\label{sec:meth-semantic}
% ----------------------------------------------------------------

After the AST is constructed, the engine performs a systematic
semantic validation pass before any automaton is built.  This pass
catches errors that are syntactically valid but logically
inconsistent with the supported SQL:2016 row-pattern subset, and
applies a set of normalization rules that simplify downstream
processing.

% ................................
\subsection{Schema and Variable Consistency Checks}
\label{subsec:sem-schema}
% ................................

The validation and execution setup verify the following properties,
raising parser or execution errors when a query cannot be mapped
safely to the input DataFrame:

\begin{itemize}
  \item Columns referenced in \texttt{PARTITION BY},
    \texttt{ORDER BY}, \texttt{DEFINE}, and \texttt{MEASURES}
    expressions must be resolvable against the input DataFrame during
    execution setup or expression evaluation.

  \item Pattern-variable references in \texttt{MEASURES} or
    \texttt{DEFINE} expressions must refer to variables declared in
    the \texttt{PATTERN} clause or made available through
    \texttt{SUBSET}.  Variables without explicit \texttt{DEFINE}
    predicates are allowed and are treated as implicit always-true
    variables.

  \item If a \texttt{SUBSET} clause is present, the subset mapping is
    normalized and made available to later stages.  References to
    subset aliases and their component variables are then resolved
    where they are used in measures, predicates, or automaton
    construction.

  \item The \texttt{AFTER MATCH SKIP TO FIRST/LAST $V$} options
    reference a pattern variable $V$ that actually exists in the
    pattern, catching typos at compile time rather than at runtime.
\end{itemize}

% ................................
\subsection{Navigation-Function Validation}
\label{subsec:sem-navfunc}
% ................................

Navigation functions impose inter-variable ordering constraints that
must be verified before the automaton is constructed.  The
\texttt{validate\_navigation\_conditions()} function in
\texttt{src/matcher/condition\_evaluator.py} performs this check by
building a \emph{variable timeline}: the order in which distinct
pattern variables first appear in the pattern string.  For each
\texttt{DEFINE} predicate the validator then checks:

\begin{itemize}
  \item Cross-variable \texttt{NEXT($v$.$c$)} references must point
    to a variable that appears later in the pattern timeline.
    Self-references and unqualified column navigation are handled by
    the evaluator according to the available row context.  If a
    navigation condition is determined to be invalid for the pattern,
    the engine returns an empty result set rather than raising an
    exception, matching the Trino-oriented behavior used in the
    tests.

  \item \texttt{PREV($v$.$c$, $k$)} references are accepted when
    the referenced variable is available in the pattern context; the
    evaluator also checks boundary conditions during execution.

  \item Nested navigation expressions (e.g.\
    \texttt{NEXT(PREV(A.price, 1), 1)}) are handled by the enhanced
    navigation-evaluation path; executor-side helper routines also
    validate nested forms where they are used.
\end{itemize}

% ................................
\subsection{Output-Mode and Skip-Mode Consistency}
\label{subsec:sem-modes}
% ................................

Several combinations of output mode, skip mode, and pattern
features are invalid in the supported SQL:2016-oriented subset.  The
validator enforces the following rules:

\begin{itemize}
  \item Pattern exclusion syntax (\texttt{\{- \ldots -\}}) is
    incompatible with \texttt{ALL ROWS PER MATCH WITH UNMATCHED
    ROWS}; this combination is rejected with a descriptive error
    message.

  \item \texttt{AFTER MATCH SKIP TO FIRST/LAST $V$} requires that
    $V$ be a pattern variable, not a subset alias.  Subset aliases
    are resolved at automaton-construction time, not at skip-mode
    computation time, so this restriction is necessary to ensure
    a deterministic next-start position.
\end{itemize}

\noindent
After all checks pass, the validator records the resolved
configuration—normalized output mode, skip mode, implicit-variable
defaults, and measure semantics—ready for the logical planning step.

% ................................................................
\section{Logical Plan Construction}
\label{sec:meth-logical}
% ................................................................

The logical plan translates the validated AST into a flat,
canonical data structure that drives all subsequent phases without
further reference to the raw SQL text.  In the implementation this
materializes as a set of typed Python objects extracted from the
\texttt{MatchRecognizeClause} AST node and passed as explicit
arguments to the execution functions.
Table~\ref{tab:logical-plan} lists the principal logical-plan
attributes and their types.

\begin{table}[htbp]
  \centering
  \caption{Principal logical-plan attributes extracted from the
    validated AST}
  \label{tab:logical-plan}
  \begin{tabular}{llp{6.5cm}}
    \toprule
    \textbf{Attribute} & \textbf{Type} & \textbf{Semantics} \\
    \midrule
    \texttt{partition\_by}   & \texttt{List[str]}
      & Column names for DataFrame partitioning; empty list means
        single global partition \\[4pt]
    \texttt{order\_by}       & \texttt{List[str]}
      & Column names for intra-partition sort order \\[4pt]
    \texttt{pattern\_text}   & \texttt{str}
      & Normalized pattern string forwarded to the tokenizer \\[4pt]
    \texttt{define}          & \texttt{Dict[str, str]}
      & Variable $\to$ predicate expression mapping; implicit
        variables (absent from \texttt{DEFINE}) are assigned
        the always-true predicate \texttt{"True"} \\[4pt]
    \texttt{subset\_dict}    & \texttt{Dict[str, List[str]]}
      & Subset alias $\to$ constituent variable list, extracted
        by \texttt{extract\_subset\_dict()} \\[4pt]
    \texttt{measures}        & \texttt{Dict[str, str]}
      & Measure alias $\to$ expression string \\[4pt]
    \texttt{measure\_semantics} & \texttt{Dict[str, str]}
      & Measure alias $\to$ \texttt{"RUNNING"} or
        \texttt{"FINAL"}, determined per the rules summarised in
        Table~\ref{tab:rows-per-match} \\[4pt]
    \texttt{rows\_per\_match} & \texttt{RowsPerMatch}
      & Output-row multiplicity enum value \\[4pt]
    \texttt{skip\_mode}      & \texttt{SkipMode}
      & After-match skip strategy enum value \\[4pt]
    \texttt{skip\_var}       & \texttt{Optional[str]}
      & Target variable for \textsc{To First}/\textsc{To Last};
        \texttt{None} for the other two modes \\
    \bottomrule
  \end{tabular}
\end{table}

% ................................
\subsection{Implicit Variable Defaults}
\label{subsec:logical-implicit}
% ................................

SQL:2016 permits pattern variables to be listed in the
\texttt{PATTERN} clause without a corresponding entry in the
\texttt{DEFINE} clause.  Such variables match any row
unconditionally.  During logical plan construction the engine
identifies these implicit variables by comparing the set of names
extracted from the pattern tokens against the keys of the
\texttt{define} dictionary.  Any variable present in the former
but absent from the latter is assigned the always-true predicate
(represented as the Python string \texttt{"True"}), so that later
stages can treat all variables uniformly without special-casing
implicit ones.

% ................................
\subsection{Measure Semantics Resolution}
\label{subsec:logical-measure-sem}
% ................................

Measure semantics (\textsc{Running} vs.\ \textsc{Final}) can be
specified explicitly using the keywords \texttt{RUNNING} and
\texttt{FINAL} in the \texttt{MEASURES} clause, or left implicit
for the engine to determine.  The resolution algorithm
(Figure~\ref{fig:measure-sem-resolution}) scans all measure
expressions in order:

\begin{enumerate}
  \item If any measure carries an explicit \texttt{RUNNING} or
    \texttt{FINAL} keyword, the query is in \emph{mixed-semantics
    mode} and all other measures without an explicit keyword
    default to \textsc{Final}.

  \item In single-semantics mode (no explicit keywords):
    \begin{itemize}
      \item All measures default to \textsc{Final} under
        \texttt{ONE ROW PER MATCH}.
      \item Under \texttt{ALL ROWS PER MATCH}: \texttt{LAST()}
        defaults to \textsc{Running}; \texttt{FIRST()},
        \texttt{PREV()}, and \texttt{NEXT()} default to
        \textsc{Final}; aggregate functions \texttt{COUNT()},
        \texttt{SUM()}, and \texttt{ARRAY\_AGG()} default to
        \textsc{Running}; all others default to \textsc{Final}.
    \end{itemize}
\end{enumerate}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    node distance=0.5cm and 1.8cm,
    every node/.style={font=\small},
    decision/.style={diamond, draw, aspect=2.2,
                     inner sep=1pt, align=center},
    result/.style={rectangle, draw, rounded corners=3pt,
                   minimum width=2.8cm, minimum height=0.6cm,
                   align=center},
    arrow/.style={->, thick}
  ]
    \node[decision] (explicit) {Explicit\\keyword?};
    \node[result, right=2.2cm of explicit] (useexp) {Use stated\\semantics};
    \node[decision, below=1.2cm of explicit] (mixed) {Any measure\\explicit?};
    \node[result, right=2.2cm of mixed] (final0) {\textsc{Final}};
    \node[decision, below=1.2cm of mixed] (onerow) {\texttt{ONE ROW}\\mode?};
    \node[result, right=2.2cm of onerow] (final1) {\textsc{Final}};
    \node[decision, below=1.2cm of onerow] (functype) {\texttt{LAST}/\texttt{COUNT}/\\
      \texttt{SUM}/\texttt{ARRAY\_AGG}?};
    \node[result, right=2.2cm of functype] (running) {\textsc{Running}};
    \node[result, below=0.9cm of functype] (final2) {\textsc{Final}};

    \draw[arrow] (explicit) -- node[above,font=\tiny]{yes} (useexp);
    \draw[arrow] (explicit) -- node[left,font=\tiny]{no} (mixed);
    \draw[arrow] (mixed) -- node[above,font=\tiny]{yes} (final0);
    \draw[arrow] (mixed) -- node[left,font=\tiny]{no} (onerow);
    \draw[arrow] (onerow) -- node[above,font=\tiny]{yes} (final1);
    \draw[arrow] (onerow) -- node[left,font=\tiny]{no} (functype);
    \draw[arrow] (functype) -- node[above,font=\tiny]{yes} (running);
    \draw[arrow] (functype) -- node[left,font=\tiny]{no} (final2);
  \end{tikzpicture}
  \caption{Measure semantics resolution algorithm applied to each
    measure expression individually.  The decision tree follows
    SQL:2016 default-semantics rules.}
  \label{fig:measure-sem-resolution}
\end{figure}

\noindent
Once the required attributes are resolved, the logical plan is
assembled and passed to the pattern-compilation phase described in
Chapter~\ref{chap:implementation}.

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:meth-summary}
% ----------------------------------------------------------------

This chapter has presented the system design and methodology of the
proposed \texttt{MATCH\_RECOGNIZE} engine for Pandas DataFrames.  It
introduced the five-stage architecture---parsing, semantic
validation, logical plan construction, pattern compilation, and
execution---and explained how the first three stages transform a
user query into a validated logical representation.

The chapter first described the overall architecture and then focused
on the front-end methodology: SQL parsing, grammar selection,
ANTLR4-based parse-tree processing, AST construction, safe expression
handling, semantic validation, and logical-plan normalization.  These
stages establish the boundary between declarative SQL input and the
lower-level execution machinery.

Chapter~\ref{chap:implementation} continues from this logical design
by describing the automata compilation pipeline, the Pandas execution
model, the concrete implementation components, and the optimization
techniques used to improve runtime performance and memory behavior.
