% ============================================================
%  ch5 Implementation and Optimization.tex
%  Chapter 5: Implementation and Optimization
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% Provide \mr if not already defined by the main document
\providecommand{\mr}{\texttt{MATCH\_RECOGNIZE}}
% !TEX root = Thesis.tex
\chapter{Implementation and Optimization}
\label{chap:implementation}

This chapter describes how the logical design introduced in
Chapter~\ref{chap:methodology} is realized as an executable
\texttt{MATCH\_RECOGNIZE} engine for Pandas DataFrames.  The focus is
on the implementation layers that translate a validated logical plan
into executable pattern matching: automata compilation, pattern
tokenization, NFA/DFA construction, Pandas-based execution, measure
materialization, and performance optimization.

The chapter is organized as follows.  Section~\ref{sec:meth-automata}
explains how row patterns are compiled into finite automata.
Section~\ref{sec:meth-pandas} describes execution over Pandas
DataFrames, including partitioning, row classification, skip modes,
output modes, and measure evaluation.  Section~\ref{sec:meth-impl}
summarizes the concrete implementation components.  Finally,
Section~\ref{sec:bg-performance} presents the optimization layer,
including LRU pattern caching, memory management, and enhanced
condition evaluation.

% ----------------------------------------------------------------
\section{Pattern Compilation with Finite Automata}
\label{sec:meth-automata}
% ----------------------------------------------------------------

Pattern matching in \texttt{MATCH\_RECOGNIZE} can be modeled as a
sequence-recognition problem.  A sequence of data rows, each labeled
with the pattern variable or variables it satisfies, constitutes a
string over the alphabet of pattern variables.  The problem of
determining whether this string is accepted by the pattern is
therefore naturally expressed using finite automata~\cite{HopcroftUllman}.

% ................................
\subsection{Non-deterministic Finite Automata (NFA)}
\label{subsec:meth-nfa}
% ................................

A \emph{Non-deterministic Finite Automaton} (NFA) is formally
defined as a 5-tuple $M = (Q,\,\Sigma,\,\delta,\,q_0,\,F)$, where:

\begin{itemize}
  \item $Q$ is a finite set of \emph{states};
  \item $\Sigma$ is the \emph{input alphabet} (pattern variables in
    this setting);
  \item $\delta : Q \times (\Sigma \cup \{\varepsilon\}) \to 2^{Q}$
    is the \emph{transition function}, mapping each state and symbol
    (or the empty string $\varepsilon$) to a \emph{set} of successor
    states;
  \item $q_0 \in Q$ is the \emph{initial state}; and
  \item $F \subseteq Q$ is the set of \emph{accepting states}.
\end{itemize}

NFAs naturally model alternation and optional constructs owing to
their ability to be in multiple states simultaneously and to use
$\varepsilon$-transitions (transitions that consume no input).

Figure~\ref{fig:meth-nfa} shows the NFA for pattern $A^{+}\,B?$.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
      >=stealth, shorten >=1pt,
      node distance=2.8cm, auto, thick,
      every state/.style={circle, draw, minimum size=1cm}
    ]
    \node[state, initial, initial text={}]  (q0) {$q_0$};
    \node[state]                            (q1) [right of=q0] {$q_1$};
    \node[state]                            (q2) [right of=q1] {$q_2$};
    \node[state, accepting]                 (q3) [right of=q2] {$q_3$};

    \path[->]
      (q0) edge              node {$A$}           (q1)
      (q1) edge [loop above] node {$A$}           (q1)
      (q1) edge              node {$\varepsilon$} (q2)
      (q2) edge              node {$B$}           (q3)
      (q2) edge [bend right=35] node [below] {$\varepsilon$} (q3);
  \end{tikzpicture}
  \caption{NFA for the pattern $A^{+}\,B?$ illustrating
    $\varepsilon$-transitions and nondeterminism}
  \label{fig:meth-nfa}
\end{figure}

% ................................
\subsection{Thompson's Construction}
\label{subsec:meth-thompson}
% ................................

Thompson's construction algorithm~\cite{Thompson1968} converts a
regular expression into an NFA in linear time $O(n)$, where $n$ is
the length of the expression.  The algorithm proceeds recursively
over the expression structure according to the rules in
Algorithm~\ref{alg:thompson}.

\begin{algorithm}[htbp]
\caption{Thompson's NFA Construction}
\label{alg:thompson}
\begin{algorithmic}[1]
\Procedure{Build}{$e$}
  \If{$e = \varepsilon$}
    \State \Return NFA with one $\varepsilon$-transition from start to accept
  \ElsIf{$e = a \in \Sigma$}
    \State \Return NFA with one $a$-labeled transition from start to accept
  \ElsIf{$e = r \cdot s$ \Comment{concatenation}}
    \State $N_r \gets \Call{Build}{r}$
    \State $N_s \gets \Call{Build}{s}$
    \State merge accept state of $N_r$ with start state of $N_s$
    \State \Return combined NFA
  \ElsIf{$e = r \mid s$ \Comment{alternation}}
    \State $N_r \gets \Call{Build}{r}$
    \State $N_s \gets \Call{Build}{s}$
    \State add new start $q_\text{new}$ with $\varepsilon$-edges to $q_0^{N_r}$ and $q_0^{N_s}$
    \State add new accept $f_\text{new}$ with $\varepsilon$-edges from $F^{N_r}$ and $F^{N_s}$
    \State \Return combined NFA
  \ElsIf{$e = r^{*}$ \Comment{Kleene star}}
    \State $N_r \gets \Call{Build}{r}$
    \State add $\varepsilon$-edge: start $\to$ accept \Comment{zero repetitions}
    \State add $\varepsilon$-edge: $F^{N_r} \to q_0^{N_r}$ \Comment{loop back}
    \State \Return modified NFA
  \ElsIf{$e = r^{\{n,m\}}$ \Comment{bounded quantifier}}
    \State construct $n$ mandatory copies followed by $(m{-}n)$ optional copies
    \State \Return combined NFA
  \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

In its classical form, Thompson's construction is linear in the size
of the regular expression~\cite{DragonBook}.  The implementation uses
the same construction principle, while adding wrapper states,
predicate metadata, priorities, and safety guards for SQL-specific
constructs such as \texttt{PERMUTE}.  Consequently, the exact state
count depends on the supported SQL pattern constructs rather than on
the textbook $2n$ bound alone.

% ................................
\subsection{Pattern Tokenization}
\label{subsec:meth-pattern-tokeniser}
% ................................

Before an NFA can be constructed the raw pattern string extracted
from the SQL query must be converted into a structured, traversable
representation.  The \texttt{tokenize\_pattern()} function in
\texttt{src/matcher/pattern\_tokenizer.py} performs this step in a
single left-to-right scan, producing a flat list of
\texttt{PatternToken} objects—the \emph{linearized} form of the
pattern's syntax tree.

\paragraph{PatternToken structure.}
Each \texttt{PatternToken} is a Python \texttt{dataclass} with the
following key fields:

\begin{itemize}
  \item \textbf{\texttt{type}} — a \texttt{PatternTokenType} enum
    value determining which NFA fragment builder is invoked
    (see Table~\ref{tab:token-types});
  \item \textbf{\texttt{value}} — the raw textual value extracted
    from the pattern string (variable name, operator character, etc.);
  \item \textbf{\texttt{quantifier}} — optional quantifier string
    (\texttt{'*'}, \texttt{'+'}, \texttt{'?'}, or
    \texttt{'\{$n$,$m$\}'}) copied from the immediately following
    quantifier when present; defaults to \texttt{None};
  \item \textbf{\texttt{greedy}} — boolean flag indicating greedy
    (\texttt{True}, the default) or reluctant (\texttt{False})
    quantification, set when a trailing \texttt{?} follows the
    quantifier;
  \item \textbf{\texttt{metadata}} — a dictionary carrying
    type-specific structured data (e.g.\ a \texttt{variables} list
    for \texttt{PERMUTE} tokens, an \texttt{excluded\_variables}
    list for \texttt{EXCLUSION} tokens); and
  \item \textbf{\texttt{priority}} — an integer priority field used
    by selected builders, especially alternation and \texttt{PERMUTE}
    handling, to order competing transitions; tokens normally default
    to priority zero unless a construction step assigns a more
    specific ordering.
\end{itemize}

\begin{table}[htbp]
  \centering
  \caption{The thirteen \texttt{PatternTokenType} values and their
    roles in the pattern string}
  \label{tab:token-types}
  \begin{tabular}{lp{8cm}}
    \toprule
    \textbf{Token type}         & \textbf{Role in pattern} \\
    \midrule
    \texttt{LITERAL}            & A pattern variable name (\texttt{A}, \texttt{UP},
                                  \texttt{STRT}, etc.) \\
    \texttt{ALTERNATION}        & The \texttt{|} operator \\
    \texttt{PERMUTE}            & A complete \texttt{PERMUTE(\ldots)} construct
                                  carrying its variable list in \texttt{metadata} \\
    \texttt{GROUP\_START}       & An opening parenthesis \texttt{(} \\
    \texttt{GROUP\_END}         & A closing parenthesis \texttt{)} \\
    \texttt{ANCHOR\_START}      & The start-of-partition anchor \texttt{\^{}} \\
    \texttt{ANCHOR\_END}        & The end-of-partition anchor \texttt{\$} \\
    \texttt{EXCLUSION}          & A complete \texttt{\{- \ldots -\}} exclusion block \\
    \texttt{EXCLUSION\_START}   & Opening delimiter \texttt{\{-} \\
    \texttt{EXCLUSION\_END}     & Closing delimiter \texttt{-\}} \\
    \texttt{QUANTIFIER}         & A standalone quantifier node \\
    \texttt{SUBSET}             & A \texttt{SUBSET} variable reference \\
    \texttt{SEQUENCE}           & A composite sequence node \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Tokenization algorithm.}
The internal function \texttt{\_tokenize\_pattern\_internal()} scans
the pattern string character by character, dispatching on the current
position:

\begin{enumerate}
  \item Whitespace is skipped unconditionally.
  \item The keyword \texttt{PERMUTE} (case-insensitive) triggers
    \texttt{\_parse\_permute\_pattern()}, which collects the
    comma-separated variable list inside the following parentheses
    and returns a single \texttt{PERMUTE} token whose
    \texttt{metadata[`variables']} holds the parsed variable tokens.
  \item The two-character sequence \texttt{\{-} triggers
    \texttt{\_parse\_exclusion\_pattern()}, which reads to the
    matching \texttt{-\}} (tracking nesting depth) and returns an
    \texttt{EXCLUSION} token.
  \item The structural characters \texttt{(}, \texttt{)},
    \texttt{|}, \texttt{\^{}}, and \texttt{\$} each produce a
    single-character token of the corresponding type.
  \item Quantifier characters \texttt{*}, \texttt{+}, \texttt{?},
    and \texttt{\{} are parsed by \texttt{parse\_quantifier\_at()}
    and \emph{attached} to the preceding token as its
    \texttt{quantifier} and \texttt{greedy} fields—no separate
    token is inserted.
  \item Identifier characters (\texttt{[A-Za-z0-9\_]}) are
    accumulated into a \texttt{LITERAL} token.
\end{enumerate}

After the scan, \texttt{\_optimize\_token\_sequence()} makes a
conservative pass that removes structural redundancies while
preserving the one-variable-per-token invariant that the NFA builder
requires: adjacent \texttt{LITERAL} tokens that look like pattern
variable names are deliberately \emph{not} merged.

\paragraph{Quantifier normalization.}
Quantifiers are stored as attributes of the token they modify rather
than as separate tokens.  The helper \texttt{parse\_quantifier()}
normalizes all surface forms to the triple
$(\textit{min},\,\textit{max},\,\textit{greedy})$:
\begin{center}
\begin{tabular}{lll}
  \texttt{?} & $\to$ & $(0,\;1,\;\mathit{True})$ \\
  \texttt{*} & $\to$ & $(0,\;\infty,\;\mathit{True})$ \\
  \texttt{+} & $\to$ & $(1,\;\infty,\;\mathit{True})$ \\
  \texttt{\{$n$\}} & $\to$ & $(n,\;n,\;\mathit{True})$ \\
  \texttt{\{$n$,$m$\}} & $\to$ & $(n,\;m,\;\mathit{True})$ \\
  append \texttt{?} to any & $\to$ & $\mathit{greedy} = \mathit{False}$
\end{tabular}
\end{center}

\noindent This triple is consumed directly by
\texttt{NFABuilder.\_apply\_quantifier()} during NFA construction.

% ................................
\subsection{NFA Construction from Tokens}
\label{subsec:meth-nfa-from-tokens}
% ................................

The \texttt{NFABuilder} class in \texttt{src/matcher/automata.py}
implements the \emph{visitor} pattern over the token list.  Its
central method \texttt{\_process\_sequence()} iterates the token
stream and dispatches on each token's \texttt{type} field to invoke
the appropriate NFA-fragment builder—mirroring the recursive
tree-traversal structure of classical Thompson's construction but
applied to the flat linearized token representation rather than an
explicit parse tree.

\paragraph{Overall build sequence.}
The public entry point \texttt{NFABuilder.build()} executes the
following steps:

\begin{enumerate}
  \item Check the LRU pattern cache (keyed on an MD5 hash of the
    token list and DEFINE context) for a previously compiled NFA.
    A cache hit returns a deep clone of the stored NFA to prevent
    state sharing.
  \item On a cache miss call \texttt{\_build\_nfa\_internal()}, which:
    \begin{enumerate}[label=(\alph*)]
      \item validates anchor semantics (e.g.\ rejecting patterns such
        as \texttt{A\^{}} where a start anchor follows a literal);
      \item creates a global start state $q_\text{start}$ and accept
        state $q_\text{accept}$;
      \item delegates to \texttt{\_process\_sequence()} for the
        pattern body; and
      \item connects the pattern fragment to the global states via
        $\varepsilon$-transitions.
    \end{enumerate}
  \item Apply \texttt{NFA.optimize()} to remove unreachable states,
    deduplicate transitions, and pre-compute DFA hot-paths.
  \item Cache the compiled NFA and return it.
\end{enumerate}

\subsubsection{Token-to-NFA Fragment Mapping}
\label{subsubsec:meth-token-to-nfa}

Table~\ref{tab:token-nfa-map} summarizes how the visitor translates
each token type into NFA structure.

\begin{table}[htbp]
  \centering
  \caption{Token types and their NFA fragment builders in
    \texttt{NFABuilder.\_process\_sequence()}}
  \label{tab:token-nfa-map}
  \begin{tabular}{lp{8.2cm}}
    \toprule
    \textbf{Token type}            & \textbf{NFA construction rule} \\
    \midrule
    \texttt{LITERAL}               & \texttt{create\_var\_states()} produces a
                                     two-state fragment
                                     $q_s \xrightarrow{C(v)} q_e$, where
                                     $C(v)$ is the callable compiled from the
                                     \texttt{DEFINE} predicate for variable $v$;
                                     the token's \texttt{quantifier} (if any) is
                                     applied by \texttt{\_apply\_quantifier()} \\[4pt]
    \texttt{ALTERNATION}           & The left branch is finalized; the right branch
                                     is processed by a recursive call to
                                     \texttt{\_process\_sequence()}; a fresh start
                                     state fans out with $\varepsilon$-transitions
                                     to both branch heads, and a fresh end state
                                     collects both branch tails \\[4pt]
    \texttt{GROUP\_START}          & Triggers a recursive call to
                                     \texttt{\_process\_sequence()} that reads
                                     until the matching \texttt{GROUP\_END}; the
                                     group quantifier (if any) is applied to the
                                     resulting fragment \\[4pt]
    \texttt{PERMUTE}               & \texttt{\_process\_permute()} expands all
                                     $n!$ orderings of the $n$ variables and
                                     builds a union of $n!$ concatenation
                                     fragments under a shared start/end pair \\[4pt]
    \texttt{EXCLUSION}             & \texttt{\_process\_exclusion\_token()} builds
                                     the contained sub-pattern normally but marks
                                     every state in the range with
                                     \texttt{is\_excluded = True}; the matcher
                                     filters excluded variables from output at
                                     match time \\[4pt]
    \texttt{ANCHOR\_START}/\texttt{END}
                                   & Creates a single anchor state with
                                     \texttt{is\_anchor = True} and the
                                     corresponding \texttt{anchor\_type};
                                     partition-boundary semantics are enforced
                                     by the matcher \\
    \bottomrule
  \end{tabular}
\end{table}

\subsubsection{Quantifier Expansion}
\label{subsubsec:meth-apply-quantifier}

\texttt{\_apply\_quantifier(start, end, }$\mathit{min}$\texttt{, }$\mathit{max}$\texttt{, }$\mathit{greedy}$\texttt{)}
wraps a fragment $(s,\,e)$ in outer envelope states $(q_s,\,q_e)$
and adds $\varepsilon$-transitions accordingly:

\begin{itemize}
  \item \textbf{\texttt{?}} $(\min=0,\,\max=1)$: adds
    $\varepsilon(q_s \to s)$ for the match case and
    $\varepsilon(q_s \to q_e)$ for the skip case.
  \item \textbf{\texttt{*}} $(\min=0,\,\max=\infty)$: adds
    $\varepsilon(q_s \to s)$, $\varepsilon(e \to q_e)$ (exit),
    $\varepsilon(q_s \to q_e)$ (skip), and
    $\varepsilon(e \to s)$ (loop for repetition).
  \item \textbf{\texttt{+}} $(\min=1,\,\max=\infty)$: as \texttt{*}
    but without the skip transition, enforcing at least one match.
  \item \textbf{\texttt{\{$n$,$m$\}}}: represents mandatory and
    optional repetitions using additional states and
    $\varepsilon$-transitions.  The implementation uses a simplified
    bounded-repetition expansion rather than a complete deep clone
    of an arbitrary sub-automaton for every repetition.
  \item \textbf{Reluctant} (\texttt{greedy = False}): the parser
    records reluctant quantifiers through the token's
    \texttt{greedy} flag.  The NFA construction currently preserves
    the same structural transitions and logs the reluctant case; the
    greedy/reluctant distinction is therefore only partially
    reflected in transition priorities.
\end{itemize}

\subsubsection{PERMUTE Expansion}
\label{subsubsec:meth-permute-expansion}

The \texttt{PERMUTE}$(v_1, v_2, \ldots, v_n)$ construct is
semantically equivalent to the alternation of all $n!$ orderings of
its arguments.  In the main NFA builder,
\texttt{\_process\_permute()} proceeds in three phases:

\begin{enumerate}
  \item \textbf{Complexity guard.}  Patterns with more than eight
    variables are rejected immediately (raising
    \texttt{RuntimeError}) because $8! = 40\,320$ is the largest
    feasible inline expansion; patterns estimated to produce more
    than 50\,000 combinations are redirected to a conservative
    fallback via \texttt{\_process\_permute\_limited()}.
  \item \textbf{Permutation generation.}
    The builder generates permutation branches with
    \texttt{itertools.permutations}; optional arguments are expanded
    through subset-permutation combinations.  A separate
    \texttt{ProductionPermuteHandler} utility also exists for
    reusable cached permutation expansion, but the main automaton
    construction path performs the branch expansion inside
    \texttt{NFABuilder}.
  \item \textbf{NFA assembly.}  For each permutation
    $\sigma = (\sigma_1, \ldots, \sigma_n)$, the builder constructs
    a concatenation fragment by calling
    \texttt{create\_var\_states()} for each $\sigma_i$ and connecting
    the fragments with $\varepsilon$-transitions.  All $n!$
    concatenation fragments are then united under a common fork/merge
    pair with the same $\varepsilon$-alternation structure used for
    the \texttt{|} operator.
\end{enumerate}

% ................................
\subsection{Properties of the Constructed NFA}
\label{subsec:meth-nfa-properties}
% ................................

The NFA returned by \texttt{NFABuilder.build()} satisfies the
following formal properties:

\begin{enumerate}
  \item \textbf{Controlled size for ordinary patterns.}  For
    ordinary concatenation, grouping, alternation, anchors, and
    simple quantifiers, the builder adds a small number of states per
    token or construct.  Constructs such as \texttt{PERMUTE} can grow
    factorially, so explicit complexity guards cap the number of
    expanded branches.

  \item \textbf{Single accept state.}  Exactly one accept state
    $q_\text{accept}$ exists; all fragment tails are connected to
    it via $\varepsilon$-transitions.

  \item \textbf{Priority-annotated $\varepsilon$-transitions.}
    Every $\varepsilon$-transition carries an integer priority
    stored in the \texttt{epsilon\_priorities} dictionary on its
    source \texttt{NFAState}.  Lower numeric values denote higher
    priority.  For greedy quantifiers the loop-back
    $\varepsilon$ has lower priority than the exit $\varepsilon$,
    so the $\varepsilon$-closure computation (which sorts targets by
    ascending priority) extends the match before terminating.
    These priorities are propagated through subset construction so
    that each DFA
    transition inherits the minimum (highest-priority) value of the
    contributing NFA transitions.  This priority layer supports the
    implementation's greedy and leftmost-preference behavior, while
    some SQL edge cases remain governed by executor-level heuristics.

  \item \textbf{Thread safety.}  Each \texttt{NFAState} and the
    \texttt{NFA} object itself are guarded by a
    \texttt{threading.RLock()}.  Epsilon-closure results are cached
    in a bounded dictionary (capped at 1\,000 entries) accessed
    under the same lock, protecting shared automaton data if the
    execution layer is extended to concurrent partition workers.

  \item \textbf{Structural validation.}  \texttt{NFA.validate()}
    validates transition and $\varepsilon$ targets, checks and
    reports accept-state reachability, and validates general
    automaton metadata before the NFA is returned to the caller.
    PERMUTE metadata is also checked for required fields where it
    appears, but this is a structural consistency check rather than a
    formal proof of semantic equivalence for every expanded
    permutation.
\end{enumerate}

% ................................
\subsection{Worked Example: Pattern \texttt{A\;B+\;C?}}
\label{subsec:meth-nfa-example}
% ................................

Consider the pattern \texttt{A B+ C?} with three pattern variables
\texttt{A}, \texttt{B}, \texttt{C} each associated with an
arbitrary Boolean DEFINE predicate.  The tokenizer produces the
flat list:
\begin{center}
  \texttt{[\,LITERAL('A'),\quad LITERAL('B', quantifier='+'),\quad
    LITERAL('C', quantifier='?')\,]}
\end{center}
The \texttt{NFABuilder} then processes this list as follows.

\begin{enumerate}
  \item A global start state $q_0$ and accept state $q_a$ are
    created; $q_a$ is marked \texttt{is\_accept = True}.

  \item \textbf{LITERAL \texttt{A} (no quantifier).}
    \texttt{create\_var\_states} allocates $q_1$ and $q_2$; adds
    transition $q_1 \xrightarrow{C(A)} q_2$.
    The visitor adds $\varepsilon(q_0 \to q_1)$ and advances
    \texttt{current} to $q_2$.

  \item \textbf{LITERAL \texttt{B} with quantifier \texttt{+}.}
    \texttt{create\_var\_states} allocates $q_3, q_4$ with
    $q_3 \xrightarrow{C(B)} q_4$.
    \texttt{\_apply\_quantifier}$(q_3,\,q_4,\,1,\,\infty,\,\mathit{True})$
    allocates envelope states $q_5, q_6$ and adds:
    $\varepsilon(q_5 \to q_3)$,\;
    $\varepsilon(q_4 \to q_6)$ (exit),\;
    $\varepsilon(q_4 \to q_3)$ (loop, greedy).
    The visitor adds $\varepsilon(q_2 \to q_5)$ and advances to $q_6$.

  \item \textbf{LITERAL \texttt{C} with quantifier \texttt{?}.}
    \texttt{create\_var\_states} allocates $q_7, q_8$ with
    $q_7 \xrightarrow{C(C)} q_8$.
    \texttt{\_apply\_quantifier}$(q_7,\,q_8,\,0,\,1,\,\mathit{True})$
    allocates $q_9, q_{10}$ and adds:
    $\varepsilon(q_9 \to q_7)$ (take $C$),\;
    $\varepsilon(q_8 \to q_{10})$ (exit after $C$),\;
    $\varepsilon(q_9 \to q_{10})$ (skip $C$, bypass).
    The visitor adds $\varepsilon(q_6 \to q_9)$ and advances to $q_{10}$.

  \item $\varepsilon(q_{10} \to q_a)$ connects the pattern's tail
    to the global accept state.
\end{enumerate}

\noindent
The constructed automaton accepts the language
$A \cdot B^{+} \cdot C^{?}$: one row satisfying $A$, one or more
rows satisfying $B$ (the loop $q_4 \to q_3$ implements the
\texttt{+} case), and zero or one row satisfying $C$ (the bypass
$q_9 \to q_{10}$ implements the \texttt{?}).  In the current
implementation, this example produces twelve NFA states after the
global start/accept states and quantifier envelopes are included.

After construction the NFA is passed to \texttt{DFABuilder.build()}
for subset construction as described in the next section.

% ................................
\subsection{Deterministic Finite Automata (DFA)}
\label{subsec:meth-dfa}
% ................................

A \emph{Deterministic Finite Automaton} (DFA) is a special case of
an NFA in which:
\begin{enumerate*}[label=(\roman*)]
  \item there are no $\varepsilon$-transitions, and
  \item for each state $q$ and symbol $a$, there is at most one
    successor state in the partial transition relation used by the
    implementation.
\end{enumerate*}
DFAs require $O(n)$ time to process an input of length $n$ with no
backtracking, making them ideal for row-by-row matching.

\paragraph{Subset Construction (NFA $\to$ DFA).}
The \emph{subset construction} (powerset construction) algorithm
converts an NFA to an equivalent DFA.  Each DFA state corresponds
to a \emph{set} of NFA states.  For a DFA state $S$ and input
symbol $a$:

\[
  \delta_{\text{DFA}}(S,\,a)
  \;=\;
  \varepsilon\text{-closure}\!\left(
    \bigcup_{q \in S} \delta_{\text{NFA}}(q,\,a)
  \right)
\]

where $\varepsilon\text{-closure}(T)$ is the set of all NFA states
reachable from any state in $T$ by zero or more
$\varepsilon$-transitions.  Although the resulting DFA may have up
to $2^{|Q|}$ states in the worst case, patterns arising from
\texttt{MATCH\_RECOGNIZE} syntax are structurally constrained and in
practice yield DFAs with far fewer states.

\paragraph{DFA Optimization.}
After subset construction, the implementation applies iterative
optimization passes: unreachable states are removed, equivalent
states are merged when their metadata and transitions are compatible,
and transitions are grouped for faster lookup.  This reduces memory
footprint and matching overhead in practice, but it is not presented
as a full canonical Hopcroft minimization.

\paragraph{State Prioritization for SQL Semantics.}
SQL:2016 mandates \emph{greedy} matching by default: the engine
must prefer the longest possible match, and in the case of
alternation, the leftmost alternative takes precedence.  Standard
DFA theory does not encode these preferences directly.  For the
supported pattern subset, the implementation approximates these
preferences by:
\begin{enumerate}
  \item Assigning a \emph{priority} to each NFA transition based on
    its position in the pattern and quantifier greediness.
  \item Propagating priorities through subset construction so that
    each DFA transition inherits the minimum (highest-priority)
    value of the contributing NFA transitions.
  \item Ordering outgoing DFA transitions by priority during
    match execution so that higher-priority paths are considered
    first in ambiguous cases.
\end{enumerate}


% ----------------------------------------------------------------
\section{Pandas DataFrame Execution Model}
\label{sec:meth-pandas}
% ----------------------------------------------------------------

Pandas~\cite{McKinney2010} is a Python library providing the
\texttt{DataFrame} data structure: a two-dimensional, labeled
table whose columns are stored in array-like structures, commonly
NumPy or Pandas extension arrays.  Its in-memory, column-oriented
storage enables vectorized operations over large row sets, making
it well suited for the row-by-row DFA simulation that
\texttt{MATCH\_RECOGNIZE} requires.

This section describes how the engine transforms a validated
\texttt{MATCH\_RECOGNIZE} clause and a compiled DFA into concrete
output rows.  The process passes through six distinct stages:
partition, order, row classification, DFA simulation,
measure evaluation, and output assembly.
Figure~\ref{fig:exec-pipeline} places these stages in context
of the full five-phase architecture introduced at the start of
this chapter.

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
    node distance=0.6cm and 1.6cm,
    every node/.style={font=\small},
    box/.style={rectangle, draw, rounded corners=3pt,
                minimum width=2.8cm, minimum height=0.75cm,
                align=center},
    arrow/.style={->, thick}
  ]
    \node[box, fill=gray!15] (parse) {SQL Parsing};
    \node[box, fill=gray!15, right=of parse] (sem) {Semantic\\Validation};
    \node[box, fill=gray!15, right=of sem] (plan) {Logical Plan};
    \node[box, fill=gray!15, right=of plan] (comp) {Pattern\\Compilation};
    \node[box, fill=blue!15, right=of comp] (exec) {\textbf{Execution}};
    \draw[arrow] (parse) -- (sem);
    \draw[arrow] (sem) -- (plan);
    \draw[arrow] (plan) -- (comp);
    \draw[arrow] (comp) -- (exec);
  \end{tikzpicture}
  \caption{Five-phase execution pipeline; the execution stage (shaded)
    is the subject of this section}
  \label{fig:exec-pipeline}
\end{figure}

% ................................
\subsection{Partition and Order}
\label{subsec:meth-exec-partition}
% ................................

Execution begins by splitting the input DataFrame $D$ into
independent \emph{partitions}.  The columns named in the
\texttt{PARTITION BY} clause are passed to
\texttt{DataFrame.groupby()}, which returns one sub-frame per
distinct combination of partition-key values.  When no
\texttt{PARTITION BY} clause is present the entire DataFrame
constitutes a single partition.

Each partition is then sorted by the \texttt{ORDER BY} columns
using \texttt{sort\_values()}, establishing the temporal order
within which the DFA simulation will proceed.  The underlying
NumPy index is preserved so that row references in navigation
functions remain stable.

Because partitions share no state, they can be processed
independently.  The implementation keeps this partition-level
independence explicit, and the optimization layer discussed in
Section~\ref{subsec:bg-parallel} provides a parallel-aware scheduling
boundary for future execution modes.  The remainder of this section
describes the processing of a single partition $P$ of $n$ rows,
indexed $r_0, r_1, \ldots, r_{n-1}$ in sorted order.

% ................................
\subsection{Row Classification}
\label{subsec:meth-exec-classify}
% ................................

Before the DFA simulation begins, every \texttt{DEFINE} predicate
is compiled into a callable Python function.  The
\texttt{compile\_condition()} function in
\texttt{src/matcher/condition\_evaluator.py} parses the predicate
expression—which may reference pattern variables via dot notation
(\texttt{A.price}), navigation functions, and SQL comparison
operators—and wraps it in a closure that accepts a row dictionary
and a \texttt{RowContext} object.
Consistent with Trino's documented behavior, pattern variables that
appear in \texttt{PATTERN} but are absent from \texttt{DEFINE} are
treated as implicit always-true variables, so they may match any
row unless constrained by surrounding pattern structure~\cite{trino2022}.

The \texttt{RowContext}
(Listing~\ref{lst:rowcontext-iface}) is the central evaluation
context.  It provides:
\begin{itemize}
  \item \texttt{rows}: the full sorted partition as a list of row
    dictionaries, enabling look-ahead and look-behind via index
    arithmetic;
  \item \texttt{variables}: the current match's variable bindings
    (a mapping from variable name to a list of matching row
    indices);
  \item \texttt{current\_idx}: the index of the row being tested;
  \item \texttt{current\_var}: the name of the pattern variable
    whose predicate is being evaluated; and
  \item \texttt{subsets}: the \texttt{SUBSET} variable definitions.
\end{itemize}

\begin{lstlisting}[
  language=Python,
  caption={Key fields of \texttt{RowContext} consulted during
    condition evaluation},
  label={lst:rowcontext-iface}]
@dataclass
class RowContext:
    rows:              List[Dict[str, Any]]   # full partition
    variables:         Dict[str, List[int]]   # var -> [row indices]
    current_idx:       int                    # row under test
    current_var:       Optional[str]          # variable being matched
    subsets:           Dict[str, List[str]]   # SUBSET definitions
    defined_variables: Set[str]               # DEFINE variable names
    pattern_variables: List[str]              # all pattern variables
\end{lstlisting}

\noindent
Predicates are evaluated lazily during DFA simulation: a condition
is called when the matcher attempts the corresponding transition.
For larger partitions, the engine enables a hybrid condition
preprocessing path that increases condition-cache capacity and, for
very large inputs, pre-warms common condition-evaluation paths on a
sample of rows.  The codebase also contains a condition-matrix
routine for precomputing Boolean arrays, but the main matching path
currently relies primarily on the hybrid cache/pre-warming strategy.

% ................................
\subsection{DFA Simulation}
\label{subsec:meth-exec-dfa-sim}
% ................................

The core of the execution stage is the DFA simulation loop, shown
as Algorithm~\ref{alg:dfa-sim}.  Starting from the DFA's initial
state $q_0$, the engine processes the sorted partition one row at a
time.  At each row $r_i$ it evaluates the \texttt{DEFINE} predicates
for every outgoing transition of the current state, collects the
transitions whose predicates evaluate to \texttt{True}, and selects
the highest-priority one (reflecting greedy or leftmost-match
semantics).  When the DFA reaches an accepting state, the engine
records the current variable bindings as a candidate match and
either terminates (for simple patterns) or continues to seek a
longer greedy match.

\begin{algorithm}[htbp]
\caption{DFA Simulation Loop for a Single Partition}
\label{alg:dfa-sim}
\begin{algorithmic}[1]
\Procedure{SimulatePartition}{$P,\, \mathit{dfa},\, \mathit{skip\_mode},\, \mathit{config}$}
  \State $\mathit{results} \gets [\,]$
  \State $i \gets 0$
  \While{$i < |P|$}
    \State $q \gets q_0$
    \State $\mathit{bindings} \gets \{\}$
    \State $\mathit{best} \gets \bot$
    \State $j \gets i$
    \While{$j < |P|$}
      \State $\mathit{fired} \gets
        \{(v,q') \mid (q \xrightarrow{v} q') \in \mathit{dfa},
          C(v, P[j], \mathit{ctx}_{j,\mathit{bindings}})\}$
      \If{$\mathit{fired} = \varnothing$}
        \State \textbf{break} \Comment{no transition -- end of match attempt}
      \EndIf
      \State $(v^*,q^*) \gets \Call{BestTransition}{\mathit{fired}}$
      \State $\mathit{bindings}[v^*] \gets
          \mathit{bindings}[v^*] \cup \{j\}$
      \State $q \gets q^*$
      \If{$\mathit{dfa.accepting}(q)$}
        \State $\mathit{best} \gets
          (\mathit{start}{=}i,\;\mathit{end}{=}j,\;
           \mathit{bindings}\text{ copy})$
        \If{$\mathit{skip\_mode} = \textsc{ToNextRow}$}
          \State \textbf{break}
        \EndIf
      \EndIf
      \State $j \gets j + 1$
    \EndWhile
    \If{$\mathit{best} \neq \bot$}
      \State $\mathit{results}.\textsc{Append}(\mathit{best})$
      \State $i \gets \Call{NextStart}{i,\;\mathit{best},\;
        \mathit{skip\_mode},\;\mathit{bindings}}$
    \Else
      \State $i \gets i + 1$
    \EndIf
  \EndWhile
  \State \Return $\mathit{results}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\paragraph{BestTransition selection.}
When multiple transitions fire simultaneously (e.g.\ when a row
satisfies more than one pattern variable's predicate) the engine
applies the following priority ordering, consistent with
SQL:2016 greedy-leftmost semantics:
\begin{enumerate}
  \item Transitions leading to accepting states are preferred over
    non-accepting ones, ensuring a valid prefix is recorded as a
    candidate match before later rows are considered.
  \item Among non-accepting transitions, those whose source and
    target state are the same (self-loops for repeating quantifiers
    such as \texttt{A+}) are preferred over state-advancing
    transitions—implementing greedy extension.
  \item When the pattern contains cross-variable references in
    \texttt{DEFINE} predicates (e.g.\ \texttt{B.price > A.price}),
    the engine further biases toward the self-loop for the
    prerequisite variable (\texttt{A}) before switching to the
    dependent one (\texttt{B}), ensuring $A$'s rows are fully
    consumed before $B$ begins.
  \item Remaining ties are broken by alternation order: the
    leftmost alternative in the original \texttt{PATTERN} string
    receives the lowest numeric priority and is therefore selected
    first.
\end{enumerate}

\paragraph{Variable bindings.}
Throughout the simulation, \texttt{bindings} is a dictionary
$\{ v \mapsto [i_1, i_2, \ldots] \}$ mapping each pattern
variable $v$ to the (sorted) list of partition row indices
assigned to it during the current match attempt.  At the
conclusion of a successful match these lists provide the concrete
row sets over which measure expressions and navigation functions
are evaluated.

Rows belonging to \emph{exclusion} variables (those wrapped in
\texttt{\{- \ldots -\}}) participate in condition evaluation and
state transitions, but the implementation omits them from the
\texttt{ALL ROWS PER MATCH} output and from the measure-evaluation
context.

% ................................
\subsection{After Match Skip}
\label{subsec:meth-exec-skip}
% ................................

After recording a match the engine must decide where to start the
next match attempt, controlling whether matches can overlap.
SQL:2016 defines four \texttt{AFTER MATCH SKIP} strategies;
Table~\ref{tab:skip-modes} summarizes their semantics and the
corresponding next-start-position computation.

\begin{table}[htbp]
  \centering
  \caption{The four \texttt{AFTER MATCH SKIP} modes and the
    resulting next scan position}
  \label{tab:skip-modes}
  \begin{tabular}{lp{9cm}}
    \toprule
    \textbf{Mode} & \textbf{Next-start computation} \\
    \midrule
    \textsc{Past Last Row} (default) &
      $i \leftarrow \mathit{best.end} + 1$. No overlap is possible;
      the scan resumes immediately after the last matched row. \\[4pt]
    \textsc{To Next Row} &
      $i \leftarrow \mathit{best.start} + 1$. Overlapping matches
      are produced because only a single row is skipped regardless
      of match length. \\[4pt]
    \textsc{To First} $V$ &
      $i \leftarrow \min(\mathit{bindings}[V])$. The scan
      resumes at the \emph{first} row assigned to variable $V$.
      Requires $V$ to appear in the pattern; enforced at compile time. \\[4pt]
    \textsc{To Last} $V$ &
      $i \leftarrow \max(\mathit{bindings}[V])$. The scan
      resumes at the \emph{last} row assigned to variable $V$.
      May overlap if $V$'s last row precedes the match end. \\
    \bottomrule
  \end{tabular}
\end{table}

\noindent
To prevent infinite loops, the engine enforces a monotonic
progress invariant: the next scan position is always strictly
greater than the current start position.  For
\textsc{To Next Row} the engine additionally detects cycles in
the sequence of recent start positions and breaks out if the same
position is revisited.

% ................................
\subsection{Output Modes}
\label{subsec:meth-exec-output}
% ................................

SQL row-pattern recognition, including Trino's implementation,
exposes two principal output-row multiplicity modes, each admitting
further sub-options~\cite{trino2022}.  The current engine follows
these output multiplicities and applies the effective measure
contexts summarized in Table~\ref{tab:rows-per-match}.  Trino
documents \textsc{Running} as the default row-pattern expression
semantics; this implementation additionally applies function-specific
defaults for undecorated measures in \texttt{ALL ROWS PER MATCH}
mode to match the behavior exercised by the regression suite.

\begin{table}[htbp]
  \centering
  \caption{Output modes and their row-count and measure-semantics
    behavior}
  \label{tab:rows-per-match}
  \begin{tabular}{llp{7cm}}
    \toprule
    \textbf{Mode} & \textbf{Rows out} & \textbf{Measure evaluation context} \\
    \midrule
    \textsc{One Row Per Match}
      & 1 per match
      & \textsc{Final}: measures see the complete, finalized
        variable bindings.  Navigation functions
        (\texttt{FIRST}, \texttt{LAST}) operate over the full
        matched span. \\[6pt]
    \textsc{All Rows Per Match}
      & 1 per matched row
      & Implementation default per function: \texttt{LAST} uses
        \textsc{Running}; \texttt{FIRST}/\texttt{PREV}/\texttt{NEXT}
        use \textsc{Final}; \texttt{COUNT}, \texttt{SUM}, and
        \texttt{ARRAY\_AGG} use \textsc{Running}; other aggregate
        functions default to \textsc{Final}.
        Explicit \texttt{RUNNING}/\texttt{FINAL} keywords override
        the default. \\[6pt]
    \textsc{All Rows \ldots With Unmatched}
      & 1 per every input row
      & Matched rows carry measure values; unmatched rows emit
        \texttt{NULL} for all measure columns. \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{One Row Per Match.}
For each recorded match the engine constructs a single output
dictionary.  The \texttt{PARTITION BY} and \texttt{ORDER BY}
column values are copied from the first matched row; measure
expressions are evaluated once with \textsc{Final} semantics
(all variable bindings are visible).  The resulting dictionary is
appended to the results list and later assembled into the output
DataFrame.

\paragraph{All Rows Per Match.}
For each recorded match the engine iterates over the non-excluded
rows belonging to the match span (from \texttt{best.start} to
\texttt{best.end} inclusive).  For each output row $r_j$ in this
filtered span the engine:
\begin{enumerate}
  \item assigns $r_j$ a \texttt{CLASSIFIER} column value equal to
    the name of the pattern variable to which row $j$ was assigned;
  \item evaluates each measure expression with the semantics
    determined by Table~\ref{tab:rows-per-match}: \textsc{Running}
    semantics limits variable bindings to rows $\leq j$,
    while \textsc{Final} semantics always uses the complete
    bindings; and
  \item appends the resulting row dictionary to the results list.
\end{enumerate}
Rows assigned to an exclusion variable are omitted from the
\texttt{ALL ROWS PER MATCH} output in the current implementation.

% ................................
\subsection{Measure Evaluation}
\label{subsec:meth-exec-measures}
% ................................

Measure expressions are evaluated by the
\texttt{MeasureEvaluator} class in
\texttt{src/matcher/measure\_evaluator.py}.  Each declared
measure is compiled into a callable similar to the condition
callables described in Section~\ref{subsec:meth-exec-classify}.
The evaluator dispatches on the expression type:

\begin{description}
  \item[Navigation functions.]
    \texttt{FIRST($v$.$c$)}, \texttt{LAST($v$.$c$)},
    \texttt{PREV($c$, $k$)}, and \texttt{NEXT($c$, $k$)}
    resolve to direct index lookups into the partition row list:
    \begin{itemize}
      \item $\texttt{FIRST}(v.c) \to
        P[\mathit{bindings}[v][0]][c]$
      \item $\texttt{LAST}(v.c) \to
        P[\mathit{bindings}[v][-1]][c]$
      \item $\texttt{PREV}(c,\,k) \to
        P[j - k][c]$ (where $j$ is the current output row index)
      \item $\texttt{NEXT}(c,\,k) \to
        P[j + k][c]$
    \end{itemize}
    All accesses are bounds-checked: out-of-partition references
    return \texttt{NULL} per SQL:2016.

  \item[Aggregate functions.]
    \texttt{COUNT}, \texttt{SUM}, and \texttt{AVG} are implemented
    as running or final aggregates over the row index sets stored
    in \texttt{bindings}.  Under \textsc{Running} semantics the
    aggregate is computed over
    $\{ i \in \mathit{bindings}[v] \mid i \leq j \}$;
    under \textsc{Final} semantics it uses all rows in
    $\mathit{bindings}[v]$.

  \item[Pattern-variable column references.]
    An expression of the form $v.c$ (without a function wrapper)
    resolves to the column $c$ of the \emph{first} row assigned to
    $v$.  In \textsc{Running} mode the reference is restricted to
    assigned rows up to the current output position.

  \item[Arithmetic and logical expressions.]
    Compound expressions (e.g.\ \texttt{LAST(A.price) -
    FIRST(A.price)}) are evaluated by recursively evaluating
    sub-expressions and applying the indicated operator.
\end{description}

\paragraph{RUNNING vs.\ FINAL semantics.}
The distinction between the two semantics is central to
\texttt{ALL ROWS PER MATCH} output.  Under \textsc{Running}
semantics a measure evaluated for output row $j$ sees only rows
$r_0, \ldots, r_j$—the match is \emph{in progress} and future
rows are unknown.  Under \textsc{Final} semantics the full
match is always available regardless of which row is being output.

Consider a \texttt{SUM(A.price)} measure over a match covering
rows 3, 5, and 7.  In \textsc{Running} mode the output for each
row is a cumulative partial sum:
\begin{center}
\begin{tabular}{cl}
  Output row & \textsc{Running} \texttt{SUM} \\
  \hline
  3 & $P[3].\mathit{price}$ \\
  5 & $P[3].\mathit{price} + P[5].\mathit{price}$ \\
  7 & $P[3].\mathit{price} + P[5].\mathit{price} + P[7].\mathit{price}$ \\
\end{tabular}
\end{center}
whereas \textsc{Final} mode emits the same total on all three
output rows.

% ................................
\subsection{Worked End-to-End Example}
\label{subsec:meth-exec-example}
% ................................

To illustrate the full execution pipeline consider the following
query applied to a stock-price DataFrame with columns
\texttt{symbol}, \texttt{date}, and \texttt{price}:

\begin{lstlisting}[language=SQL, caption={Example query: detecting
  a V-shape price pattern}, label={lst:exec-example-query}]
SELECT *
FROM prices
MATCH_RECOGNIZE (
  PARTITION BY symbol
  ORDER BY date
  MEASURES
    FIRST(A.price) AS start_price,
    LAST(B.price)  AS bottom_price,
    LAST(C.price)  AS end_price
  ONE ROW PER MATCH
  AFTER MATCH SKIP PAST LAST ROW
  PATTERN (A+ B+ C+)
  DEFINE
    A AS PREV(A.price, 1) IS NULL OR A.price >= PREV(A.price, 1),
    B AS B.price <  PREV(B.price, 1),
    C AS C.price >  PREV(C.price, 1)
)
\end{lstlisting}

\noindent
Suppose the partition for \texttt{symbol = 'AAPL'} contains
five rows sorted by \texttt{date} with prices
$100, 102, 98, 95, 103$.  Execution proceeds as follows.

\vspace{0.5em}
\noindent
\textbf{Step 1 — Partition and Order.}
\texttt{groupby('symbol')} produces a single-symbol sub-frame;
\texttt{sort\_values('date')} confirms the price sequence
$[100, 102, 98, 95, 103]$.

\vspace{0.5em}
\noindent
\textbf{Step 2 — DFA construction.}
The pattern \texttt{A+ B+ C+} compiles to an NFA with 16 states in
the current implementation and then to a DFA with 4 states via
subset construction (Section~\ref{subsec:meth-dfa}).

\vspace{0.5em}
\noindent
\textbf{Step 3 — Simulation from $i=0$.}

\begin{itemize}[leftmargin=2em]
  \item Row 0 ($p=100$): \texttt{PREV(A.price)} is undefined at the
    first row, so the explicit \texttt{IS NULL} guard in $A$'s
    predicate is satisfied.  Transition $q_0 \xrightarrow{A} q_1$ fires;
    $\mathit{bindings}[A] = [0]$.  State $q_1$ is not accepting.
  \item Row 1 ($p=102 \geq 100$): \texttt{A} predicate satisfied;
    self-loop $q_1 \xrightarrow{A} q_1$.  $\mathit{bindings}[A] = [0,1]$.
  \item Row 2 ($p=98 < 102$): \texttt{A} fails, \texttt{B} passes
    ($p < \mathit{prev}$).  Transition $q_1 \xrightarrow{B} q_2$.
    $\mathit{bindings}[B] = [2]$.
  \item Row 3 ($p=95 < 98$): \texttt{B} satisfied; self-loop
    $q_2 \xrightarrow{B} q_2$.  $\mathit{bindings}[B] = [2,3]$.
  \item Row 4 ($p=103 > 95$): \texttt{C} satisfied; transition
    $q_2 \xrightarrow{C} q_3$.  $q_3$ is accepting.
    $\mathit{best} = (\mathit{start}{=}0,\; \mathit{end}{=}4,\;
    \{A{:}[0,1],\; B{:}[2,3],\; C{:}[4]\})$.
\end{itemize}

\vspace{0.5em}
\noindent
\textbf{Step 4 — After Match Skip.}
Mode is \textsc{Past Last Row}: next scan starts at $i = 5$, which
equals $|P|$, so the simulation terminates.

\vspace{0.5em}
\noindent
\textbf{Step 5 — Measure evaluation (ONE ROW PER MATCH, FINAL).}
\begin{align*}
  \texttt{start\_price}  &= \texttt{FIRST}(A.\mathit{price})
    = P[0][\mathit{price}] = 100 \\
  \texttt{bottom\_price} &= \texttt{LAST}(B.\mathit{price})
    = P[3][\mathit{price}] = 95 \\
  \texttt{end\_price}    &= \texttt{LAST}(C.\mathit{price})
    = P[4][\mathit{price}] = 103
\end{align*}

\vspace{0.5em}
\noindent
\textbf{Step 6 — Output assembly.}
A single output row is emitted:
\texttt{(symbol='AAPL', start\_price=100, bottom\_price=95,
end\_price=103)}.

This example demonstrates how all execution components--partition
ordering, DFA simulation with greedy quantifiers, variable binding
accumulation, after-match skip, and measure evaluation—interact to
produce a single output row for the supported SQL:2016 row-pattern
subset from a five-row price sequence.


% ----------------------------------------------------------------
\section{Implementation Components}
\label{sec:meth-impl}
% ----------------------------------------------------------------

The implementation provides a \texttt{MATCH\_RECOGNIZE} engine for Pandas
DataFrames following the supported SQL:2016 row-pattern semantics for
sequential data.  The implementation has four core components: an
ANTLR4-based SQL parser that transforms \texttt{MATCH\_RECOGNIZE}
queries into abstract syntax trees; a finite-automata engine that
converts pattern expressions into optimized NFAs and DFAs using
Thompson's construction; a pattern-matching executor that processes
DataFrames with support for partitioning and ordering; and a measure
evaluator that handles navigation functions (\texttt{PREV},
\texttt{NEXT}, \texttt{FIRST}, \texttt{LAST}) and aggregations with
both running and final semantics.  The system supports advanced
pattern features including anchors, quantifiers, alternation,
permuted patterns, and exclusion syntax.

% ................................
\subsection{ANTLR4 Grammar and AST Construction}
\label{subsec:impl-antlr4}
% ................................

The system uses the Trino SQL grammar from the ANTLR community
grammars project and extracts its \texttt{MATCH\_RECOGNIZE}
productions into project-specific AST nodes.  This ANTLR4 grammar
covers the pattern constructs used by the implementation, including
nested patterns, alternation, quantifiers, grouping, and
exclusions~\cite{antlr_trino_sql}.  The parser's output is a
detailed parse tree that captures the syntactic structure of each
SQL query~\cite{parr2013definitive,tarazona2021antlr}.

% ................................
\subsection{Finite Automata}
\label{subsec:impl-automata}
% ................................

\paragraph{Pattern Tokenization.}
Pattern expressions are tokenized using a single-pass tokenization
routine that supports anchors, grouping,
quantifiers, and nested constructs.

\paragraph{NFA Construction.}
Thompson's construction algorithm is used to translate pattern
tokens into a Non-deterministic Finite Automaton (NFA).  Each
state captures transitions, variable bindings, anchors, and
metadata for advanced features.  Quantifiers (\texttt{*},
\texttt{+}, \texttt{?}, \texttt{\{$n$,$m$\}}) are directly
mapped to sub-automata~\cite{georgiou2025streaming}.

\paragraph{DFA Construction and Optimization.}
The NFA is systematically converted into a Deterministic Finite
Automaton (DFA) using subset construction, reducing the number of
active states that must be tracked during execution.  Optimization
steps include removal of unreachable states, compatible-state
merging, and transition grouping.  The
resulting DFA supports efficient row-by-row evaluation while
preserving pattern semantics, priorities, and custom conditions
for the supported pattern semantics~\cite{e104-d_3_381,10.1145/986537.986588,georgiou2025streaming}.

% ................................
\subsection{Execution}
\label{subsec:impl-execution}
% ................................

The matching engine applies the DFA sequentially over each partition
of the DataFrame with specialized handling for anchors, skip modes,
and edge cases.  The executor coordinates the entire
pattern-matching process, integrating parsing, automata
construction, match identification, and result formatting.

\paragraph{Condition and Measure Evaluation.}
A dedicated evaluation subsystem translates SQL pattern conditions
and measures into controlled Python callables and evaluator routines
that support navigation functions and aggregations with SQL-style
\texttt{NULL} handling for both running and final calculations.

% ----------------------------------------------------------------
\section{Performance Optimization Techniques}
\label{sec:bg-performance}
% ----------------------------------------------------------------

% ................................
\subsection{Parallel Partition Processing}
\label{subsec:bg-parallel}
% ................................

Because \texttt{MATCH\_RECOGNIZE} evaluates each
\texttt{PARTITION BY} group independently, partitions provide a
natural unit of parallel work.  The implementation includes a
\texttt{ParallelExecutionManager} in
\texttt{src/utils/performance\_optimizer.py}, which estimates whether
partition-level work is large enough to justify parallel execution.
The current top-level \texttt{match\_recognize} path uses this
manager as a parallel-aware scheduling layer but still processes the
partition work sequentially after ordering each partition.  The
design therefore preserves the partition independence needed for
future parallel dispatch without changing match semantics in the
current execution path.

% ................................
\subsection{LRU Pattern Caching}
\label{subsec:bg-caching}
% ................................

Pattern compilation---tokenization $\to$ NFA construction $\to$ DFA
conversion $\to$ optimization---is computationally expensive,
especially for patterns with quantifiers and \texttt{PERMUTE}
constructs.  The implementation maintains a thread-safe
\emph{Least Recently Used} (LRU) cache that maps a hash of the
pattern string and \texttt{DEFINE} predicates to the compiled
pattern (the full NFA and DFA pipeline result).
Key configuration parameters are listed in
Table~\ref{tab:bg-cache-config}.

\begin{table}[htbp]
  \centering
  \caption{Pattern cache configuration defaults
    (\texttt{src/config/production\_config.py})}
  \label{tab:bg-cache-config}
  \begin{tabular}{lrl}
    \toprule
    \textbf{Parameter}         & \textbf{Default} & \textbf{Description} \\
    \midrule
    \texttt{cache\_size\_limit}        & 50\,000 entries  & Maximum cached patterns \\
    \texttt{cache\_memory\_limit\_mb}  & 2\,048 MB        & Maximum cache memory \\
    \texttt{cache\_ttl\_seconds}       & 3\,600 s         & Entry time-to-live \\
    \texttt{cache\_clear\_threshold\_mb} & 1\,600 MB      & Proactive eviction threshold \\
    \bottomrule
  \end{tabular}
\end{table}

% ................................
\subsection{Memory Management}
\label{subsec:bg-memory}
% ................................

Large-scale matching over millions of rows requires careful memory
management.  The implementation treats partitions as independent
execution units and iterates over them separately during matching.
In the current executor, however, partition objects are materialized
from the Pandas \texttt{groupby()} result before they are processed,
so the memory model is partition-oriented rather than fully streaming.
Memory usage is monitored via \texttt{psutil}; garbage collection is
triggered proactively when consumption approaches user-configured
limits (\texttt{cache\_clear\_threshold\_mb}).  An object-pool layer
(\texttt{src/utils/memory\_management.py}) reuses frequently allocated
objects across successive match operations, reducing allocator
pressure on long-running queries.

% ................................
\subsection{Enhanced Condition Evaluation}
\label{subsec:bg-vectorised}
% ................................

For larger partitions, the engine reduces repeated Python
function-call overhead through smart condition preprocessing.  The
main path in \texttt{EnhancedMatcher.\_smart\_condition\_preprocessing()}
enlarges the condition-evaluation cache when the partition contains
more than $5\,000$ rows and pre-warms common condition paths on a
sample when the partition contains more than $50\,000$ rows.  The
codebase also contains
\texttt{EnhancedMatcher.\_vectorize\_condition\_evaluation()}, which
can materialize a Boolean condition matrix, but this is not the
primary path used by the current matcher.

Table~\ref{tab:perf-summary} summarizes the performance
optimizations and the conditions under which each is applied.

\begin{table}[htbp]
  \centering
  \caption{Performance optimizations and their activation
    conditions}
  \label{tab:perf-summary}
  \begin{tabular}{p{3.8cm}p{4.2cm}p{4.4cm}}
    \toprule
    \textbf{Technique} & \textbf{Activation condition}
      & \textbf{Primary benefit} \\
    \midrule
    LRU pattern cache
      & Always on
      & Eliminates redundant NFA/DFA construction across repeated
        queries \\[4pt]
    Parallel-aware partition scheduling
      & Multiple partitions, sufficient rows, and acceptable memory
        pressure
      & Preserves a scheduling boundary for future concurrent
        partition execution \\[4pt]
    Enhanced condition cache
      & Partition size $> 5\,000$ rows
      & Increases cache capacity for repeated predicate evaluation \\[4pt]
    Condition pre-warming
      & Partition size $> 50\,000$ rows
      & Warms common predicate paths on a sample of rows before the
        full scan \\[4pt]
    Partition-oriented processing
      & Always on
      & Keeps match state partition-local, although Pandas group
        objects may still be materialized before processing \\[4pt]
    Proactive GC / memory guard
      & Process RSS $\geq$ \texttt{cache\_clear\_threshold\_mb}
      & Prevents OOM errors on very large dataset scans \\
    \bottomrule
  \end{tabular}
\end{table}

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:impl-summary}
% ----------------------------------------------------------------

This chapter has presented the implementation and optimization of the
\texttt{MATCH\_RECOGNIZE} engine.  It explained how row patterns are
tokenized, compiled into NFAs, converted into DFAs, and executed over
ordered Pandas partitions.  It also described the implementation of
after-match skip behavior, output modes, measure evaluation, and the
runtime components that connect the parser and logical plan to the
matching engine.

The chapter also described the optimization layer: LRU-style pattern
caching to avoid repeated compilation, parallel-aware partition
scheduling, partition-oriented memory management, and enhanced condition
caching/pre-warming to reduce repeated predicate overhead.  These
optimizations are orthogonal to the core matching
semantics and can be evaluated independently in the experimental
chapter.

Chapter~\ref{chap:evaluation} evaluates these design and
implementation decisions empirically, measuring correctness,
throughput, memory behavior, and performance across the selected
pattern families and dataset sizes.
