% ============================================================
%  ch6_experimental_evaluation.tex
%  Chapter 6: Experimental Evaluation
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% !TEX root = Thesis.tex
\chapter{Experimental Evaluation}
\label{chap:evaluation}

This chapter evaluates the proposed
\texttt{MATCH\_\allowbreak{}RECOGNIZE}
implementation across five dimensions: output correctness validated
against Trino and Oracle on selected reference queries, coverage of
the supported SQL:2016
row-pattern subset, row-pattern type coverage, performance and
scalability, and developer usability.
Section~\ref{sec:eval-methodology}
describes the evaluation methodology, including the hardware
environment and the metrics used.
Section~\ref{sec:eval-dataset} presents the datasets and test
patterns.
Section~\ref{sec:eval-correctness} validates the concept example by
example on the small datasets.
Section~\ref{sec:eval-coverage} reports feature-coverage,
pattern-type-coverage, usability, and complexity-analysis results.
Section~\ref{sec:eval-performance} presents the empirical and
cross-system analysis, including the comparison with existing tools
and the performance, throughput, and memory results.

% ----------------------------------------------------------------
\section{Evaluation Methodology}
\label{sec:eval-methodology}
% ----------------------------------------------------------------

This chapter evaluates the proposed
\texttt{MATCH\_\allowbreak{}RECOGNIZE} implementation through five
complementary criteria---correctness, SQL:2016 feature coverage,
pattern-type coverage, performance and scalability, and usability.
Section~\ref{subsec:eval-metrics} defines each criterion together
with the way it is measured, and
Section~\ref{subsec:eval-environment} describes the hardware and
software environment.

% ................................
\subsection{Evaluation Criteria and Metrics}
\label{subsec:eval-metrics}
% ................................

Each test case is assessed on the five complementary criteria
introduced above, defined here together with the way each is
measured.

\paragraph{Correctness.}
The result DataFrame produced by the implementation is compared
against two independent SQL references---Trino~473 and
Oracle~XE~21c---using an exact row-level equality check (same rows,
same column values, and same order where order is deterministic).
This dimension uses the two small synthetic tables---the
\texttt{orders} table (Table~\ref{tab:orders}) for the dip example
(Table~\ref{tab:sql-match-recognize-examples}) and the
\texttt{stock\_price} table (Table~\ref{tab:stock_price}) for the
stock-price case (Section~\ref{subsec:eval-stock-price})---as well as
the Amazon benchmark data, whose 30 pattern--size results are checked
against Trino and Oracle (Section~\ref{subsec:eval-cross-system}).
The hand-written Pandas baseline is not used as a correctness oracle;
it appears only in the usability dimension
(Section~\ref{subsec:eval-usability}) to contrast the procedural
Pandas code with the single declarative
\texttt{MATCH\_\allowbreak{}RECOGNIZE} statement.

\paragraph{SQL:2016 feature coverage.}
Support for the key \texttt{MATCH\_\allowbreak{}RECOGNIZE}
clauses---\texttt{PARTITION~BY}, \texttt{ORDER~BY}, \texttt{PATTERN},
\texttt{DEFINE}, \texttt{MEASURES}, \texttt{SUBSET}, the output modes
(\texttt{ONE~ROW~PER~MATCH} / \texttt{ALL~ROWS~PER~MATCH}),
after-match skip policies, and navigation
functions~\cite{iso2016,ISO2021,monierashraf2025rowmatch}---is
verified against the implemented subset.  Each clause and sub-feature
is marked as \emph{supported}, \emph{partially supported}, or
\emph{unsupported} based on whether the implementation supports the
feature and whether the exercised behavior matches the reference
tests.  The clauses are exercised by the reference examples above and
the Trino-aligned test suite, and the resulting statuses are
summarized in Table~\ref{tab:feature_coverage}.

\paragraph{Pattern-type coverage.}
Test cases are grouped by row-pattern type---simple sequences,
quantified repetitions
(\texttt{*}, \texttt{+}, \texttt{?}, \texttt{\{n,m\}}),
ascending/descending trends, recovery patterns, alternation,
exclusion blocks (\texttt{\{-\ldots-\}}), anchored patterns, and
\texttt{PERMUTE} constructs.  A pattern type is considered covered if
at least one test case exercises it and produces correct output.  The
types are drawn from the illustrative examples and the five benchmark
pattern families, and each is mapped to its test artifact in
Table~\ref{tab:pattern_type_coverage}.

\paragraph{Performance and scalability.}
These measurements use the large real-world Amazon UK Products dataset
(Section~\ref{subsec:eval-real-dataset}) across five pattern families
and six dataset sizes.
Execution time is recorded as \emph{client-observed end-to-end query
latency} in wall-clock seconds for each pattern--size run: it
includes result transfer and client-side DataFrame construction, and
is therefore not pure engine execution time.  Every cell of the benchmark matrix is executed
five times as warmups and then twenty times as measured runs; the
reported time is the arithmetic \emph{mean} of the twenty measured
runs, and its sample \emph{standard deviation} (together with the
minimum and maximum) is recorded alongside it to document run-to-run
spread, so a cell time can be read as mean~$\pm$~std.  Every reported
time is thus a warmed-run latency (measured after five warmups)
rather than a cold-start time.  Five warmup runs let each engine reach
steady state before measurement---important for the JVM-based Trino,
whose first cold run is otherwise several times slower than its warmed
speed---and the reported standard deviation shows how stable the
warmed runs are.
Throughput for a cell is its dataset size divided by this mean
time, and \emph{average throughput} for a system is the arithmetic
mean of its 30 per-cell throughput values.  Summary tables report
averages of these means across pattern families or dataset sizes;
all totals and averages are computed from the full-precision
measurements and only the displayed values are rounded, so a column
of rounded entries need not sum exactly to a rounded total.  The
scaling behavior is characterized both descriptively and by an
ordinary-least-squares regression of mean execution time against
row count across the six dataset sizes, whose coefficient of
determination ($R^2$) is reported per pattern family.  Five warmup
and twenty measured runs per cell give a stable, fully-warmed
estimate for every engine.

\paragraph{Memory.}
Two distinct memory metrics are recorded for every run, on the same
Amazon benchmark cells used for the timing measurements.
\emph{Query memory} is the peak additional memory attributable to
executing the measured query, excluding the already-loaded data and
the idle engine; it is taken from each system's own memory
accounting (process RSS peak delta for the Pandas engine, the
per-query \texttt{peakMemoryBytes} counter reported by Trino, and
the \texttt{session pga memory max} high-water delta of a fresh
session in Oracle).  These are \emph{system-native memory
indicators} of the same concept rather than one instrument: each
engine counts memory at a different level (operating-system pages,
reserved user memory, allocated work areas), so the values support
order-of-magnitude, per-system trend comparisons, not exact
cross-system equality.  \emph{Footprint memory} is the practical
deployment footprint of having the system running at all: the
absolute process RSS peak for the Pandas engine, and for Trino and
Oracle the peak container memory sampled while the measured query
ran, on containers that keep running across dataset sizes within a
pass.  Footprint memory is reported separately and is not a
per-query or algorithm-level metric.
Each run yields one peak value for each metric; the single value
reported per cell in the memory tables is the \emph{mean of the twenty
per-run peaks}, matching the mean used for execution time (the minimum
and maximum are not shown for memory).

\paragraph{Usability (developer effort).}
The number of non-blank, non-comment lines in the equivalent
idiomatic Pandas solution is recorded and compared with the
single-statement \texttt{MATCH\_\allowbreak{}RECOGNIZE} query.  This
dimension uses the three worked examples of
Section~\ref{sec:eval-correctness}: for each, the declarative
statement is compared line-for-line with its hand-written Pandas
baseline, and the counts are reported in
Table~\ref{tab:loc_comparison} and discussed in
Section~\ref{subsec:eval-usability}.
Hand-written Pandas solutions typically require substantially more
code (e.g.\ chained \texttt{groupby}/\texttt{shift}/joins) than a
single declarative statement, so reducing lines of code is a core
goal of the row-pattern-recognition
approach~\cite{iso2016,ISO2021,melton2017sqlrpr,Python_UDF}; the
comparison provides a concrete measure of developer-effort reduction.

% ................................
\subsection{Environment and Configuration}
\label{subsec:eval-environment}
% ................................

The experiments were performed on a workstation running Ubuntu
24.04.4 LTS with an Intel Core i7-11800H processor.  For the
cross-system benchmark, every system is restricted to one CPU core
and a 32\,GB memory budget; the enforcement mechanisms are detailed
in Section~\ref{subsec:eval-complexity}.

The software environment comprised:
\begin{itemize}
  \item \textbf{Baseline SQL engines:} Trino~473 and Oracle~XE~21c
    (both run as Docker containers).
  \item \textbf{Runtimes:} Java~11.0.25 LTS and Python~3.12.7.
  \item \textbf{Core Python libraries:} \texttt{pandas}~2.3.3,
    \texttt{numpy}~2.1.3,
    \texttt{antlr4-python3-\allowbreak{}runtime}~4.13.2,
    \texttt{antlr4-\allowbreak{}tools}~0.2.1,
    and \texttt{pytest}~8.3.4.
\end{itemize}

The \texttt{MATCH\_\allowbreak{}RECOGNIZE} implementation uses ANTLR4 for
parsing Trino SQL, \texttt{pandas} for DataFrame management, and
custom AST/NFA/DFA modules for pattern detection (see
Chapter~\ref{chap:methodology}).

Unless stated otherwise, all experiments use this configuration.  For
the correctness comparisons in particular, equivalent input data are
loaded into Trino, Oracle, and a Pandas DataFrame using matching
schemas, and the result sets are then compared by row order and
measure values after normalizing case-only column-name differences
between the SQL and Pandas outputs.

% ----------------------------------------------------------------
\section{Dataset and Test Patterns}
\label{sec:eval-dataset}
% ----------------------------------------------------------------

% ................................
\subsection{Datasets}
\label{subsec:eval-dataset-orders}
% ................................

This section describes, one by one, the three datasets used in the
evaluation: two small synthetic tables that can be verified by hand
(\texttt{orders} and \texttt{stock\_price}) and the large real-world
Amazon UK Products dataset, which supplies realistic data volumes (up
to 2\,million rows) for the timing and memory measurements.  The
concept is then validated example by example, dataset by dataset, in
Section~\ref{sec:eval-correctness}, and the large-scale performance
benchmark on the Amazon dataset follows in
Section~\ref{sec:eval-performance}.

\paragraph{Dataset 1: the \texttt{orders} table.}
The \texttt{orders} table supports the first worked example.  It
comprises three attributes~\cite{trino2021row}: \texttt{customer\_id}
(\texttt{VARCHAR}), which identifies the customer (two distinct
values, \texttt{cust\_1} and \texttt{cust\_2}); \texttt{order\_date}
(\texttt{DATE}), the date the order was placed (ranging from
2020-05-11 to 2020-05-18); and \texttt{price} (\texttt{INTEGER}), the
purchase amount on that date.  Each customer appears multiple times in
chronological order---\texttt{cust\_1} spans five dates and
\texttt{cust\_2} spans three---which enables focused validation of
partitioning, ordering, and pattern detection (e.g.\ sequences of
rising or falling prices per customer).  Table~\ref{tab:orders} shows
the eight records used.

\begin{table}[H]
  \centering
  \caption{Sample \texttt{orders} dataset
    (\texttt{customer\_id}, \texttt{order\_date}, \texttt{price})}
  \label{tab:orders}
  \begin{tabularx}{\textwidth}{|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|}
    \hline
    \textbf{customer\_id} & \textbf{order\_date} & \textbf{price} \\
    \hline
    cust\_1 & 2020-05-11 & 100 \\
    cust\_1 & 2020-05-12 & 200 \\
    cust\_2 & 2020-05-13 &   8 \\
    cust\_1 & 2020-05-14 & 100 \\
    cust\_2 & 2020-05-15 &   4 \\
    cust\_1 & 2020-05-16 &  50 \\
    cust\_1 & 2020-05-17 & 100 \\
    cust\_2 & 2020-05-18 &   6 \\
    \hline
  \end{tabularx}
\end{table}

\paragraph{Dataset 2: the \texttt{stock\_price} table.}
The \texttt{stock\_price} table supports the second worked example.
It represents a history of daily stock prices for market analysis and
includes three columns: \texttt{company} (the company name),
\texttt{price\_date} (the trading date), and \texttt{price} (the
stock's value on that day)~\cite{petkovic2022rpr}.  Two companies,
\texttt{A} and \texttt{B}, are recorded over ten trading days,
yielding twenty rows that support per-company trend detection.
Table~\ref{tab:stock_price} shows the data.

\begin{table}[H]
  \centering
  \caption[Stock price table]{The tabular form of the \texttt{stock\_price} table}
  \label{tab:stock_price}
  \begin{tabular}{@{}lll@{}}
    \toprule
    \texttt{Company} & \texttt{Price\_date} & \texttt{Price} \\
    \midrule
    A & 2020-10-01 & 50 \\
    B & 2020-10-01 & 89 \\
    A & 2020-10-02 & 36 \\
    B & 2020-10-02 & 24 \\
    A & 2020-10-03 & 39 \\
    B & 2020-10-03 & 37 \\
    A & 2020-10-04 & 42 \\
    B & 2020-10-04 & 63 \\
    A & 2020-10-05 & 30 \\
    B & 2020-10-05 & 65 \\
    A & 2020-10-06 & 47 \\
    B & 2020-10-06 & 56 \\
    A & 2020-10-07 & 71 \\
    B & 2020-10-07 & 50 \\
    A & 2020-10-08 & 80 \\
    B & 2020-10-08 & 54 \\
    A & 2020-10-09 & 75 \\
    B & 2020-10-09 & 30 \\
    A & 2020-10-10 & 63 \\
    B & 2020-10-10 & 32 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Dataset 3: the Amazon UK Products dataset.}
The Amazon UK Products Dataset is a large e-commerce dataset
containing information about products listed on Amazon's UK
marketplace~\cite{asaniczka2023amazon}, commonly used for data
analysis, machine learning, and benchmarking.  It is provided as a
single CSV file with roughly 2.2\,million product records.  Each
record includes: \texttt{asin} (Amazon Standard Identification Number;
unique product ID); \texttt{title} (product name); \texttt{imgUrl}
(primary product image URL); \texttt{productURL} (canonical product
page URL); \texttt{stars} (average customer rating, 1--5,
floating-point); \texttt{reviews} (number of customer reviews,
non-negative integer); \texttt{price} (listed price in GBP, numeric);
\texttt{isBestSeller} (bestseller indicator, boolean);
\texttt{boughtInLastMonth} (approximate purchases in the last 30 days,
integer); and \texttt{categoryName} (top-level category label, e.g.\
Electronics, Books, Clothing).  This dataset is representative for
price-trend analysis, brand/feature-popularity studies, pattern
recognition, and recommender-system benchmarking.  It supplies the
data volume for the performance and scalability benchmark
(Section~\ref{subsec:eval-real-dataset}); the derived
\texttt{category} column and the construction of the six dataset
sizes are described with the benchmark patterns in
Section~\ref{subsec:eval-pattern-defs}.

% ----------------------------------------------------------------
\section{Validation of Concept}
\label{sec:eval-correctness}
% ----------------------------------------------------------------

This section validates the implementation example by example and
dataset by dataset.  Each worked example runs the same
\texttt{MATCH\_\allowbreak{}RECOGNIZE} query through the proposed engine
and the two SQL references and checks the evaluation criteria of
Section~\ref{subsec:eval-metrics}: output correctness against
Trino~473 and Oracle~XE~21c, the clauses and pattern type the query
exercises, and developer effort against a
hand-written Pandas baseline.  Beyond these three examples, the engine's
output is checked by the project's regression suite---369 test cases
across 34 files in the \texttt{tests/} directory of the
repository~\cite{monierashraf2025rowmatch}, exercising the supported
feature and pattern set against Trino-aligned expected results---and,
live against both SQL references, by the 30 pattern--size queries of
the cross-system benchmark (Section~\ref{subsec:eval-cross-system}),
where the proposed engine, Trino, and Oracle produce identical
results throughout.

% ................................
\subsection{Example 1: Dip Pattern on the \texttt{orders} Dataset}
\label{subsec:eval-dip}
% ................................

The first example detects a `dip'---three consecutive orders with
successively lower prices---on the \texttt{orders} dataset
(Dataset~1, Table~\ref{tab:orders}).  The same detection is expressed
three ways---as a Trino SQL query (Listing~\ref{lst:orders-trino}), as
a call to the proposed engine (Listing~\ref{lst:orders-mr}), and as a
hand-written Python/pandas UDF (Listing~\ref{lst:orders-udf})---and
Table~\ref{tab:sql-match-recognize-examples} confirms that all three,
together with Oracle, return the identical match.

\begin{lstlisting}[basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Trino SQL for the dip pattern]Trino SQL: create and load the \texttt{orders} table, then detect the dip pattern with \texttt{MATCH\_\allowbreak{}RECOGNIZE}.},
  label={lst:orders-trino}]
CREATE TABLE memory.default.orders (
    customer_id VARCHAR,
    order_date  DATE,
    price       INTEGER
);

INSERT INTO memory.default.orders VALUES
    ('cust_1', DATE '2020-05-11', 100),
    ('cust_1', DATE '2020-05-12', 200),
    ('cust_2', DATE '2020-05-13',   8),
    ('cust_1', DATE '2020-05-14', 100),
    ('cust_2', DATE '2020-05-15',   4),
    ('cust_1', DATE '2020-05-16',  50),
    ('cust_1', DATE '2020-05-17', 100),
    ('cust_2', DATE '2020-05-18',   6);

SELECT customer_id, start_date, end_date, bottom_price
FROM memory.default.orders
MATCH_RECOGNIZE (
  PARTITION BY customer_id
  ORDER BY order_date
  MEASURES
    A.order_date AS start_date,
    C.order_date AS end_date,
    C.price      AS bottom_price
  PATTERN (A B C)
  DEFINE
    B AS B.price < A.price,
    C AS C.price < B.price
);
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Proposed MATCH\_RECOGNIZE on orders]Proposed engine: the same query as a \texttt{MATCH\_\allowbreak{}RECOGNIZE} string over an \texttt{orders} DataFrame.},
  label={lst:orders-mr}]
import pandas as pd
from src.executor.match_recognize import match_recognize

orders = pd.DataFrame({
    "customer_id": ["cust_1","cust_1","cust_2","cust_1",
                    "cust_2","cust_1","cust_1","cust_2"],
    "order_date":  ["2020-05-11","2020-05-12","2020-05-13","2020-05-14",
                    "2020-05-15","2020-05-16","2020-05-17","2020-05-18"],
    "price":       [100, 200, 8, 100, 4, 50, 100, 6],
})

query = """
SELECT customer_id, start_date, end_date, bottom_price
FROM orders
MATCH_RECOGNIZE (
  PARTITION BY customer_id
  ORDER BY order_date
  MEASURES
    A.order_date AS start_date,
    C.order_date AS end_date,
    C.price      AS bottom_price
  PATTERN (A B C)
  DEFINE
    B AS B.price < A.price,
    C AS C.price < B.price
); """
result = match_recognize(query, orders)
print(result)
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Equivalent dip detection in pandas]Python UDF: the same `dip' detection as hand-written Python/pandas.},
  label={lst:orders-udf}]
import pandas as pd

def detect_dip(df, cust_col="customer_id",
               date_col="order_date", price_col="price"):
    g = df.copy()
    g[date_col] = pd.to_datetime(g[date_col])
    g = g.sort_values([cust_col, date_col]).reset_index(drop=True)
    rows = []
    for cust, grp in g.groupby(cust_col, sort=False):
        grp = grp.reset_index(drop=True)
        i = 0
        while i + 2 < len(grp):
            a, b, c = (grp.loc[i, price_col],
                       grp.loc[i + 1, price_col],
                       grp.loc[i + 2, price_col])
            if b < a and c < b:            # A B C: successively lower prices
                rows.append({
                    "customer_id":  cust,
                    "start_date":   grp.loc[i,     date_col].date(),
                    "end_date":     grp.loc[i + 2, date_col].date(),
                    "bottom_price": c,
                })
                i += 3                     # AFTER MATCH SKIP PAST LAST ROW
            else:
                i += 1
    return pd.DataFrame(rows)

result = detect_dip(orders)
print(result)
\end{lstlisting}

For the \emph{correctness} criterion, all four implementations return
the same single match, shown in
Table~\ref{tab:sql-match-recognize-examples}; the query also exercises
the \texttt{PARTITION~BY}, \texttt{ORDER~BY}, \texttt{PATTERN},
\texttt{DEFINE}, and \texttt{MEASURES} clauses and covers the
simple-sequence pattern type.

\begin{table}[H]
\centering
\caption{Dip-pattern result on the \texttt{orders} dataset: the Trino
  SQL query, the proposed engine, the Python UDF, and Oracle all
  return the identical match~\cite{monierashraf2025rowmatch}}
\label{tab:sql-match-recognize-examples}
{\small
\begin{tabular}{lllr}
\toprule
\textbf{Implementation} & \texttt{customer\_id} & \texttt{start\_date} $\rightarrow$ \texttt{end\_date} & \texttt{bottom\_price} \\
\midrule
Trino~473 (SQL)             & cust\_1 & 2020-05-12 $\rightarrow$ 2020-05-16 & 50 \\
Oracle XE~21c (SQL)         & cust\_1 & 2020-05-12 $\rightarrow$ 2020-05-16 & 50 \\
Proposed engine            & cust\_1 & 2020-05-12 $\rightarrow$ 2020-05-16 & 50 \\
Python UDF                 & cust\_1 & 2020-05-12 $\rightarrow$ 2020-05-16 & 50 \\
\bottomrule
\end{tabular}
}
\end{table}

% ................................
\subsection{Example 2: Stock-Price Analysis on the \texttt{stock\_price} Dataset}
\label{subsec:eval-stock-price}
% ................................

The second example uses the \texttt{stock\_price} table (Dataset~2,
Table~\ref{tab:stock_price}) to detect bullish momentum: a three-row
window in which the second and third rows increase relative to their
immediate predecessors.  The query partitions by company
(\texttt{PARTITION BY Company}), orders chronologically
(\texttt{ORDER BY Price\_date}), applies \texttt{PATTERN (A B C)} with
\texttt{DEFINE} predicates on \texttt{B} and \texttt{C} (variable
\texttt{A} uses the implicit always-true behaviour), and reports the
starting price, ending price, and match number in \texttt{MEASURES}.
As in Example~1, the same detection is expressed as a Trino SQL query
(Listing~\ref{lst:stock-trino}), as a call to the proposed engine
(Listing~\ref{lst:stock-mr}), and as a hand-written Python/pandas UDF
(Listing~\ref{lst:stock-pandas}); Table~\ref{tab:stock-results}
confirms that all three, together with Oracle, return the identical
three matches.  Additional use cases are available in the project
repository~\cite{monierashraf2025rowmatch}.

\begin{lstlisting}[basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Trino SQL for the stock-price pattern]Trino SQL: create and load the \texttt{stock\_price} table, then detect the increasing-window pattern with \texttt{MATCH\_\allowbreak{}RECOGNIZE}.},
  label={lst:stock-trino}]
CREATE TABLE memory.default.stock_price (
    Company    VARCHAR,
    Price_date DATE,
    Price      INTEGER
);

INSERT INTO memory.default.stock_price (Company, Price_date, Price) VALUES
    ('A', DATE '2020-10-01', 50), ('B', DATE '2020-10-01', 89),
    ('A', DATE '2020-10-02', 36), ('B', DATE '2020-10-02', 24),
    ('A', DATE '2020-10-03', 39), ('B', DATE '2020-10-03', 37),
    ('A', DATE '2020-10-04', 42), ('B', DATE '2020-10-04', 63),
    ('A', DATE '2020-10-05', 30), ('B', DATE '2020-10-05', 65),
    ('A', DATE '2020-10-06', 47), ('B', DATE '2020-10-06', 56),
    ('A', DATE '2020-10-07', 71), ('B', DATE '2020-10-07', 50),
    ('A', DATE '2020-10-08', 80), ('B', DATE '2020-10-08', 54),
    ('A', DATE '2020-10-09', 75), ('B', DATE '2020-10-09', 30),
    ('A', DATE '2020-10-10', 63), ('B', DATE '2020-10-10', 32);

SELECT *
FROM memory.default.stock_price
MATCH_RECOGNIZE (
  PARTITION BY Company
  ORDER BY Price_date
  MEASURES
    FIRST(A.Price) AS start_price,
    LAST(C.Price)  AS end_price,
    MATCH_NUMBER() AS match_id
  PATTERN (A B C)
  DEFINE
    B AS B.Price > PREV(B.Price),
    C AS C.Price > PREV(C.Price)
);
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Proposed MATCH\_RECOGNIZE on stock\_price]Proposed engine: the same query as a \texttt{MATCH\_\allowbreak{}RECOGNIZE} string over a \texttt{stock\_price} DataFrame.},
  label={lst:stock-mr}]
import pandas as pd
from src.executor.match_recognize import match_recognize

data = {
    "Company": ["A","B","A","B","A","B","A","B","A","B",
                "A","B","A","B","A","B","A","B","A","B"],
    "Price_date": ["2020-10-01","2020-10-01","2020-10-02","2020-10-02",
                   "2020-10-03","2020-10-03","2020-10-04","2020-10-04",
                   "2020-10-05","2020-10-05","2020-10-06","2020-10-06",
                   "2020-10-07","2020-10-07","2020-10-08","2020-10-08",
                   "2020-10-09","2020-10-09","2020-10-10","2020-10-10"],
    "Price": [50,89,36,24,39,37,42,63,30,65,
              47,56,71,50,80,54,75,30,63,32],
}
df = pd.DataFrame(data)
df["Price_date"] = pd.to_datetime(df["Price_date"])

query = """
SELECT *
FROM stock_price
MATCH_RECOGNIZE (
  PARTITION BY Company
  ORDER BY Price_date
  MEASURES
    FIRST(A.Price) AS start_price,
    LAST(C.Price)  AS end_price,
    MATCH_NUMBER() AS match_id
  PATTERN (A B C)
  DEFINE
    B AS B.Price > PREV(B.Price),
    C AS C.Price > PREV(C.Price)
); """
result = match_recognize(query, df)
print(result)
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Equivalent logic in pandas]Python UDF: the same increasing-window detection as hand-written Python/pandas.},
  label={lst:stock-pandas}]
import pandas as pd

def detect_two_consecutive_increases(
    df: pd.DataFrame,
    company_col: str = "Company",
    date_col:    str = "Price_date",
    price_col:   str = "Price",
) -> pd.DataFrame:
    gdf = df.copy()
    gdf[date_col] = pd.to_datetime(gdf[date_col])
    gdf = gdf.sort_values([company_col, date_col]).reset_index(drop=True)
    out_rows = []
    for company, g in gdf.groupby(company_col, sort=False):
        g = g.reset_index(drop=True)
        inc = g[price_col].diff() > 0
        start_mask = (inc.shift(-1).fillna(False)
                      & inc.shift(-2).fillna(False))
        candidates = [int(i) for i, ok in enumerate(start_mask)
                      if ok and i + 2 < len(g)]
        match_id, next_start_pos = 0, 0
        for i in candidates:
            if i < next_start_pos:
                continue
            match_id += 1
            out_rows.append({
                "company":     company,
                "start_price": g.loc[i, price_col],
                "end_price":   g.loc[i + 2, price_col],
                "match_id":    match_id,
            })
            next_start_pos = i + 3   # AFTER MATCH SKIP PAST LAST ROW
    return pd.DataFrame(out_rows)

result = detect_two_consecutive_increases(df)
print(result)
\end{lstlisting}

For the \emph{correctness} criterion, all four implementations return
the same three matches, shown in Table~\ref{tab:stock-results}.  This
provides evidence that partitioning, ordering, implicit
\texttt{DEFINE} behaviour, navigation with \texttt{PREV},
\texttt{MATCH\_NUMBER()}, and measure materialisation are handled
correctly for this supported pattern family.

\begin{table}[H]
\centering
\caption{Stock-price result on the \texttt{stock\_price} dataset: the
  Trino SQL query, the proposed engine, the Python UDF, and Oracle all
  return these three identical matches~\cite{petkovic2022rpr,monierashraf2025oracle}}
\label{tab:stock-results}
{\small
\begin{tabular}{rlrrr}
\toprule
\textbf{Match} & \texttt{company} & \texttt{start\_price} & \texttt{end\_price} & \texttt{match\_id} \\
\midrule
1 & A & 36 & 42 & 1 \\
2 & A & 30 & 71 & 2 \\
3 & B & 24 & 63 & 1 \\
\bottomrule
\end{tabular}
}
\end{table}

% ................................
\subsection{Example 3: V-Shape Recovery on the \texttt{stock\_price} Dataset}
\label{subsec:eval-vshape}
% ................................

The third example reuses the \texttt{stock\_price} table (Dataset~2,
Table~\ref{tab:stock_price}) but detects a different, quantified
pattern: a \emph{V-shape recovery}---a strictly falling run followed
by a strictly rising run, i.e.\ a local price minimum and its
rebound.  Unlike the fixed three-row patterns of Examples~1 and~2,
this uses the \texttt{+} quantifier twice: \texttt{PATTERN (A B+ C+)}
matches a start row \texttt{A}, one or more falling rows \texttt{B},
and one or more rising rows \texttt{C}, with the fall and rise
expressed through \texttt{PREV} navigation in \texttt{DEFINE}.  The
measures report the starting price, the bottom (last falling row), and
the recovered price (last rising row).  As before, the same detection
is expressed as a Trino SQL query (Listing~\ref{lst:vshape-trino}), as
a call to the proposed engine (Listing~\ref{lst:vshape-mr}), and as a
hand-written Python/pandas UDF (Listing~\ref{lst:vshape-udf}), and
Table~\ref{tab:vshape-results} confirms that all three, together with
Oracle, return the identical matches.

\begin{lstlisting}[basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Trino SQL for the V-shape pattern]Trino SQL: detect the V-shape recovery on the \texttt{stock\_price} table (already created and loaded in Listing~\ref{lst:stock-trino}).},
  label={lst:vshape-trino}]
SELECT *
FROM memory.default.stock_price
MATCH_RECOGNIZE (
  PARTITION BY Company
  ORDER BY Price_date
  MEASURES
    FIRST(A.Price) AS start_price,
    LAST(B.Price)  AS bottom_price,
    LAST(C.Price)  AS end_price,
    MATCH_NUMBER() AS match_id
  ONE ROW PER MATCH
  PATTERN (A B+ C+)
  DEFINE
    B AS B.Price < PREV(B.Price),
    C AS C.Price > PREV(C.Price)
);
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Proposed MATCH\_RECOGNIZE V-shape]Proposed engine: the same V-shape query over the \texttt{stock\_price} DataFrame \texttt{df} of Listing~\ref{lst:stock-mr}.},
  label={lst:vshape-mr}]
from src.executor.match_recognize import match_recognize

query = """
SELECT *
FROM stock_price
MATCH_RECOGNIZE (
  PARTITION BY Company
  ORDER BY Price_date
  MEASURES
    FIRST(A.Price) AS start_price,
    LAST(B.Price)  AS bottom_price,
    LAST(C.Price)  AS end_price,
    MATCH_NUMBER() AS match_id
  ONE ROW PER MATCH
  PATTERN (A B+ C+)
  DEFINE
    B AS B.Price < PREV(B.Price),
    C AS C.Price > PREV(C.Price)
); """
result = match_recognize(query, df)
print(result)
\end{lstlisting}

\begin{lstlisting}[language=Python,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  caption={[Equivalent V-shape detection in pandas]Python UDF: the same V-shape detection as hand-written Python/pandas.},
  label={lst:vshape-udf}]
import pandas as pd

def detect_v_shapes(df, key="Company",
                    ts="Price_date", val="Price"):
    g = df.copy()
    g[ts] = pd.to_datetime(g[ts])
    g = g.sort_values([key, ts]).reset_index(drop=True)
    out = []
    for company, grp in g.groupby(key, sort=False):
        v = grp[val].to_numpy()
        n, i, match_id = len(v), 0, 0
        while i < n - 2:
            j = i
            while j + 1 < n and v[j + 1] < v[j]:   # B+: strictly falling
                j += 1
            if j == i:                              # need at least one fall
                i += 1
                continue
            k = j
            while k + 1 < n and v[k + 1] > v[k]:    # C+: strictly rising
                k += 1
            if k == j:                              # need at least one rise
                i += 1
                continue
            match_id += 1
            out.append({
                "company":      company,
                "start_price":  int(v[i]),
                "bottom_price": int(v[j]),
                "end_price":    int(v[k]),
                "match_id":     match_id,
            })
            i = k + 1                               # AFTER MATCH SKIP PAST LAST ROW
    return pd.DataFrame(out)

result = detect_v_shapes(df)
print(result)
\end{lstlisting}

For the \emph{correctness} criterion, all four implementations return
the same three matches, shown in Table~\ref{tab:vshape-results}.
Beyond the clauses of the earlier examples, this query additionally
exercises the \texttt{+} quantifier (one-or-more) on two pattern
variables, confirming that the engine's greedy consecutive-quantifier
handling agrees with the reference engines.

\begin{table}[H]
\centering
\caption{V-shape result on the \texttt{stock\_price} dataset: the
  Trino SQL query, the proposed engine, the Python UDF, and Oracle all
  return these three identical matches~\cite{petkovic2022rpr,monierashraf2025oracle}}
\label{tab:vshape-results}
{\small
\begin{tabular}{rlrrr}
\toprule
\textbf{Match} & \texttt{company} & \texttt{start\_price} & \texttt{bottom\_price} & \texttt{end\_price} \\
\midrule
1 & A & 50 & 36 & 42 \\
2 & B & 89 & 24 & 65 \\
3 & B & 56 & 50 & 54 \\
\bottomrule
\end{tabular}
}
\end{table}

% ................................
\subsection{Usability: Lines-of-Code Comparison}
\label{subsec:eval-usability}
% ................................

The three worked examples of Section~\ref{sec:eval-correctness} each
express the same detection three ways---as a Trino SQL query, as a
call to the proposed engine, and as a hand-written Python/pandas UDF
(Listings~\ref{lst:orders-udf}, \ref{lst:stock-pandas},
and~\ref{lst:vshape-udf}).  This places the declarative
\texttt{MATCH\_\allowbreak{}RECOGNIZE} form side by side with
the two realistic alternatives an analyst actually faces:
writing the detection \emph{by hand in Python}, or \emph{sending the
data to an external SQL engine} such as Trino or Oracle.

\paragraph{Value versus hand-written Python.}
Table~\ref{tab:loc_comparison} counts non-blank, non-comment lines for
the three examples.  In each case the single declarative statement is
about half the code of the equivalent UDF or less---14 versus 27 lines
for the dip example, 14 versus 33 for the stock-price example, and 16
versus 35 for the quantified V-shape example---and the gap widens as
the pattern grows more complex.

\begin{table}[H]
\centering
\caption{Lines-of-code comparison for the three worked examples
  (non-blank, non-comment lines)}
\label{tab:loc_comparison}
{\small
\begin{tabular}{lllr}
\toprule
\textbf{Example} & \textbf{Implementation} & \textbf{Notes} & \textbf{LOC} \\
\midrule
Dip (\texttt{orders}) & \texttt{MATCH\_RECOGNIZE} statement & single declarative query & 14 \\
Dip (\texttt{orders}) & Hand-written \texttt{pandas} UDF & loop / index / skip bookkeeping & 27 \\
\midrule
Stock (\texttt{stock\_price}) & \texttt{MATCH\_RECOGNIZE} statement & single declarative query & 14 \\
Stock (\texttt{stock\_price}) & Hand-written \texttt{pandas} UDF & \texttt{groupby}/\texttt{shift}/index bookkeeping & 33 \\
\midrule
V-shape (\texttt{stock\_price}) & \texttt{MATCH\_RECOGNIZE} statement & quantified declarative query & 16 \\
V-shape (\texttt{stock\_price}) & Hand-written \texttt{pandas} UDF & nested falling/rising run scan & 35 \\
\bottomrule
\end{tabular}
}
\end{table}

Beyond the raw line count, the declarative form is easier to
\emph{write}, \emph{read}, and \emph{edit}:
\begin{itemize}
  \item \textbf{Intent, not mechanics.} The query states \emph{what}
    to find---a partition key, an ordering, a \texttt{PATTERN}, and
    per-variable \texttt{DEFINE} conditions---while the engine decides
    \emph{how} to find it.  Each UDF must instead spell out the
    grouping, sorting, windowing, comparison, and output construction
    by hand.
  \item \textbf{No manual index or boundary bookkeeping.} The UDFs
    track positions (\texttt{i}, \texttt{i+1}, \texttt{i+2}), guard
    array bounds, and advance the cursor by hand to emulate
    \texttt{AFTER MATCH SKIP PAST LAST ROW}; a single off-by-one error
    silently changes the result.  The statement encodes the same skip
    policy in one clause.
  \item \textbf{Editable and reusable.} Changing the pattern---say, a
    four-row dip, a different comparison, or an added measure---is a
    small edit to \texttt{PATTERN}, \texttt{DEFINE}, or
    \texttt{MEASURES}, whereas the UDF must be re-derived (new
    offsets, new loop structure) for each variant.  The declarative
    form therefore generalises across the pattern families of
    Section~\ref{subsec:eval-pattern-defs} without rewriting
    procedural code.
  \item \textbf{Fewer places to go wrong.} Because partitioning,
    ordering, skip behaviour, and measures live in one statement
    rather than in scattered procedural steps, the hand-written UDFs
    were more prone to errors during development~\cite{Python_UDF}.
\end{itemize}

\paragraph{Value versus an external SQL engine.}
\texttt{MATCH\_\allowbreak{}RECOGNIZE} is already available in mature
SQL engines such as Trino and Oracle, and both examples confirm that
the proposed engine returns results identical to them.  Its value is
therefore not a new capability but \emph{where} that capability runs:
the proposed engine executes the same declarative query
\emph{in-process, directly on an in-memory \texttt{pandas}
DataFrame}, whereas Trino and Oracle require the data to be moved into
a database first.  For a DataFrame-centred or notebook workflow this
avoids standing up and connecting to a database server; declaring a
schema and loading the rows (the \texttt{CREATE TABLE}/\texttt{INSERT}
round-trip visible in the Trino listings of both examples); and
serialising the results back out again.  It also keeps the operational
footprint small---the engine peaks below 1\,GB as a single Python
process, against roughly 2.4\,GB for the Oracle container and about
23\,GB for the Trino container under the same workload
(Section~\ref{subsec:eval-memory})---while preserving the same
SQL:2016 syntax, so a query remains portable back to those engines
when a full database is warranted.  The trade-off is scope: the
proposed engine supports the subset of
Section~\ref{subsec:eval-feature-coverage}, not the complete SQL
surface, mature optimiser, or distributed execution of Trino and
Oracle.

As noted in Chapter~\ref{chap:conclusions}, this is a code-level
comparison, not a controlled user study.

% ----------------------------------------------------------------
\section{Coverage and Usability Results}
\label{sec:eval-coverage}
% ----------------------------------------------------------------

This section reports the SQL:2016 feature-coverage and
pattern-type-coverage dimensions defined in
Section~\ref{sec:eval-methodology}; the usability dimension was
covered with the worked examples in
Section~\ref{subsec:eval-usability}, and the theoretical complexity
analysis follows in Section~\ref{subsec:eval-complexity}.

% ................................
\subsection{SQL Feature Coverage}
\label{subsec:eval-feature-coverage}
% ................................

Table~\ref{tab:feature_coverage} summarizes the support status of
the main \texttt{MATCH\_\allowbreak{}RECOGNIZE} clauses in the current
implementation.  The statuses reflect the implemented subset
described in Chapters~\ref{chap:methodology}
and~\ref{chap:implementation} as exercised by the Trino-aligned
reference tests; \emph{partially supported} means that commonly used
forms are implemented and verified while less common variants remain
outside the validated scope.

\begin{table}[H]
\centering
\caption{Support status of \texttt{MATCH\_RECOGNIZE} features in the
  proposed engine}
\label{tab:feature_coverage}
{\small
\begin{tabular}{lll}
\toprule
\textbf{Feature} & \textbf{Status} & \textbf{Exercised by} \\
\midrule
\texttt{PARTITION BY} & supported & reference queries, tests \\
\texttt{ORDER BY} & supported & all benchmark queries \\
Quantifiers \texttt{* + ? \{n,m\}} & supported & benchmark families \\
Alternation \texttt{(A|B)} & supported & benchmark families \\
Grouping / nesting & supported & \texttt{complex\_nested} \\
Exclusion \texttt{\{- \ldots -\}} & supported & test suite \\
Anchors \texttt{\^{}}, \texttt{\$} & supported & test suite \\
\texttt{PERMUTE} & supported (complexity guards) & test suite \\
\texttt{DEFINE} predicates & supported & all benchmark queries \\
Navigation \texttt{PREV/NEXT/FIRST/LAST} & supported & stock case, tests \\
\texttt{MEASURES} core aggregates & supported & benchmark queries \\
\texttt{RUNNING}/\texttt{FINAL} semantics & partially supported & test suite \\
General SQL expressions & partially supported & test suite \\
\texttt{SUBSET} & supported & test suite \\
\texttt{ONE ROW PER MATCH} & supported & all benchmark queries \\
\texttt{ALL ROWS PER MATCH} & supported & test suite \\
\texttt{AFTER MATCH SKIP} modes & supported & test suite \\
\texttt{MATCH\_NUMBER()}, \texttt{CLASSIFIER()} & supported & stock case, tests \\
\bottomrule
\end{tabular}
}
\end{table}

% ................................
\subsection{Pattern-Type Coverage}
\label{subsec:eval-pattern-coverage}
% ................................

Table~\ref{tab:pattern_type_coverage} maps each row-pattern type
from Section~\ref{subsec:eval-metrics} to the test artifact that
exercises it.  A type counts as covered when at least one test case
exercises it and produces output identical to the reference systems.

\begin{table}[H]
\centering
\caption{Pattern-type coverage}
\label{tab:pattern_type_coverage}
{\small
\begin{tabular}{lll}
\toprule
\textbf{Pattern type} & \textbf{Exercised by} & \textbf{Covered} \\
\midrule
Simple sequence & \texttt{simple\_sequence} family; dip example & yes \\
Quantified repetition & \texttt{quantified} family & yes \\
Ascending/descending trend & stock-price case (Example 2) & yes \\
Recovery (V-shape) & V-shape case (Example 3); test suite & yes \\
Alternation & \texttt{alternation} family & yes \\
Exclusion & Trino-aligned test suite & yes \\
Anchors & Trino-aligned test suite & yes \\
\texttt{PERMUTE} & Trino-aligned test suite & yes \\
\bottomrule
\end{tabular}
}
\end{table}

% ................................................................
\subsection{Complexity Analysis}
\label{subsec:eval-complexity}
% ................................................................

This subsection explains complexity in two complementary ways:
\emph{theoretical analysis} and \emph{empirical analysis}.  The
theoretical analysis describes how the algorithm should grow as the
input grows and is developed here; the empirical analysis reports what
was actually measured in the experiments and is presented in
Section~\ref{sec:eval-performance}.  These two views are related, but
they are not the same: theory predicts scaling, while experiments
measure the current implementation on a specific machine and dataset.

\begin{table}[H]
\centering
\caption{Theoretical and empirical views of complexity analysis}
\label{tab:complexity_analysis}
\begin{tabularx}{\textwidth}{
  >{\raggedright\arraybackslash}p{0.46\textwidth}
  >{\raggedright\arraybackslash}X}
\toprule
\textbf{Theoretical Analysis} & \textbf{Empirical Analysis} \\
\midrule
Mathematical analysis. &
Experimental measurement. \\
\midrule
Independent of hardware and software environment. &
Depends on the hardware, operating system, Python runtime,
\texttt{pandas}, and memory behavior. \\
\midrule
Uses asymptotic notation such as Big-O, Big-$\Theta$, and
Big-$\Omega$. &
Uses actual execution time, throughput, and memory usage. \\
\midrule
Predicts scalability as input size grows. &
Measures real-world performance for the evaluated benchmark. \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Theoretical analysis.}
As implemented, the execution path has five main steps:

\begin{enumerate}
  \item parse the SQL query and extract the
    \texttt{MATCH\_\allowbreak{}RECOGNIZE} clauses;
  \item compile the row pattern into an internal automaton
    (NFA, then DFA);
  \item partition the input and order each partition;
  \item vectorize the \texttt{DEFINE} predicates over the ordered
    rows and derive the set of positions that can start a match;
  \item scan the candidate start positions, run the automaton, and
    build the output matches and measures.
\end{enumerate}

Let $n$ be the number of input rows, $p$ the number of partitions,
$n_i$ the number of rows in partition $i$, $d$ the number of
\texttt{DEFINE} variables, $k$ the number of matches produced, and
$\ell$ the average number of rows returned per match.  Thus,
$\sum_{i=1}^{p} n_i = n$.  For the five benchmark patterns used in
this chapter the pattern is fixed while the dataset size grows from
100\,K to 2.22\,M rows.

Steps~1 and~2 depend only on the query, not on $n$.  Automaton
compilation exhibits combinatorial worst-case growth in the pattern
size---exponential in general, and factorial for \texttt{PERMUTE},
which enumerates all permutations of its arguments---so the
implementation applies complexity guards and conservative fallbacks;
for the fixed benchmark patterns this cost is a small constant paid
once.

Step~3 is the ordering step, and the implementation treats it
conditionally rather than always sorting.  Each partition is first
tested for monotonicity in the \texttt{ORDER~BY} columns
($O(n_i)$); \texttt{pandas} \texttt{sort\_values} is invoked
\emph{only} when the partition is not already ordered, at a cost of
\[
  O\!\left(\sum_{i=1}^{p} n_i \log n_i\right) \le O(n \log n).
\]
When every partition is already ordered the sort is skipped and the
step reduces to the $O(n)$ monotonicity check.  The benchmark queries
order by a sequential \texttt{seq\_id} whose values are already
monotonic, so this run takes the $O(n)$ path---which is precisely why
the measured scaling is approximately linear rather than
$n \log n$.  Both branches are exercised and confirmed empirically in
Section~\ref{subsec:eval-ordering}, which reruns the benchmark on
shuffled input to force the $O(n\log n)$ branch and recovers exactly
the predicted super-linear ordering cost.

Step~4 evaluates the row-local \texttt{DEFINE} predicates as
vectorized column operations, producing per-variable Boolean arrays
and run-length tables in $O(nd)$ time and letting the scan skip
positions that cannot begin a match; $d$ is a small constant for the
benchmark patterns.  For patterns whose variables are all
vectorizable, the engine additionally precomputes per-DFA-state
decision tables (byte arrays holding the winning transition per
row), at a cost of $O(n \cdot |Q_D|)$ where the DFA state count
$|Q_D|$ is a query constant.

Step~5 visits the candidate start positions.  Each match attempt
runs the compiled matcher over the rows it consumes: the DFA walk
reduces to $O(1)$ byte-array indexing per row consumed, and the
linear-pattern search over quantifier bounds memoizes failed
(token, position) states across all start positions of the scan, so
each such state is explored at most once per pass ($O(n \cdot t)$
in total for the $t$ pattern tokens, a query constant).  Variable
assignments are recorded as run segments and expanded only for
accepted matches.  Under the default
\texttt{AFTER MATCH SKIP PAST LAST ROW} used by the benchmark,
accepted matches are non-overlapping and the scan advances past
each match, so the total scan and output work is
\[
  O(n + k\ell),
\]
where $O(n)$ covers the position scan and $O(k\ell)$ the construction
of matched rows and measures (the fused \texttt{ONE ROW PER MATCH}
driver emits each output row directly from the match's run
segments, and running-aggregate arithmetic in measures is computed
by an $O(\ell)$ prefix pass per match rather than an $O(\ell^2)$
per-row re-evaluation).  Combining the steps, the conservative
end-to-end bound is
\[
  O(n \log n + nd + k\ell),
\]
which on already-ordered input such as the benchmark data---where the
sort is skipped and the ordering step is the $O(n)$ monotonicity
check---reduces to $O(n + nd + k\ell)$, and collapses further to
$O(n + k\ell)$ (linear in $n$) for fixed $d$.  This
bound is not universal: overlapping skip modes (e.g.\ \texttt{SKIP TO
NEXT ROW}) and heavily ambiguous patterns can revisit rows and raise
the per-position cost above the linear regime, and queries whose
\texttt{DEFINE} predicates depend on the evolving match state
(aggregates or \texttt{CLASSIFIER()} in \texttt{DEFINE}) are routed
to an exact preference-ordered backtracking search whose worst case
is exponential in the match length and is therefore capped by an
explicit step budget (Section~\ref{subsec:meth-exec-backtracking}).

For daily workloads, where new rows often arrive out of order, the
most practical future improvement is incremental ordering.  Historical
data can be stored in sorted form; when a new daily batch arrives,
only the new batch is sorted and then merged with the already ordered
history.  This does not remove ordering completely, but it avoids
sorting the full history again for every run.

The extra space used by the matcher comes mainly from the partition
and ordering structures, the compiled automaton and vectorized
predicate matrix, and the output rows:
\[
  \text{extra space} \approx O(n + \text{pattern state} + k\ell).
\]
The input DataFrame itself also uses memory proportional to the number
of rows and columns.  Because the benchmark uses no
\texttt{PARTITION~BY} clause, it forms a single partition and runs on
one core; the engine's parallel-partition path is engaged only when
more than one partition is present and is therefore not exercised
here.

% ----------------------------------------------------------------
\section{Empirical and Cross-System Analysis}
\label{sec:eval-performance}
\label{subsec:eval-real-dataset}
% ----------------------------------------------------------------

Where the complexity analysis of Section~\ref{subsec:eval-complexity}
predicts scaling by asymptotic argument, this section reports what
was actually \emph{measured}: the empirical behaviour of the proposed
engine and its cross-system comparison with Trino~473 and
Oracle~XE~21c.  These are experiments rather than proofs---they
characterise the current implementation on a specific machine,
dataset, and resource budget, not bounds that hold for all inputs.

Performance was evaluated on the large real-world Amazon UK Products
dataset (Dataset~3, Section~\ref{subsec:eval-dataset-orders}) using
five pattern families and six input sizes (100\,K, 200\,K,
400\,K, 800\,K, 1.6\,M, and 2.22\,M rows, the last being the full
dataset), giving 30 pattern--size cells per system; the derived
\texttt{category} column and the construction of the six sizes are
detailed with the benchmark patterns in
Section~\ref{subsec:eval-pattern-defs}.  Each cell is reported as the
mean of twenty measured runs after five warmups, together with its
standard deviation.  All 90 cross-system cells completed successfully.
Over its 30 cells the proposed engine achieved an average throughput
of approximately 5\,043\,941~rows/s, and the sum of its 30 mean cell
times was about 5.96\,s (this figure aggregates the reported means,
not the total duration of all repetitions).  The results show
predictable, approximately linear scaling over the evaluated range
($R^2 \geq 0.98$ for the proposed engine's time-versus-size fit on
each of its five pattern families): simple sequences and bounded
quantifiers are the fastest families, while complex nested patterns
incur higher transition and condition-evaluation overhead.

The remainder of this section reports the empirical results across all
evaluation metrics: Section~\ref{subsec:eval-pattern-defs} defines the
five benchmark patterns; Section~\ref{subsec:eval-cross-system}
describes the resource policy together with the per-cell
execution-time and match-count results;
Section~\ref{subsec:eval-throughput} reports throughput;
Section~\ref{subsec:eval-memory} reports the two memory metrics; and
Section~\ref{subsec:eval-key-findings} summarises the key findings.

\paragraph{Positioning relative to existing systems.}
The evaluated engine is not intended to replace full SQL engines such
as Oracle or Trino, nor streaming platforms such as Flink; its design
goal is narrower---to bring a practical subset of SQL row-pattern
matching into in-memory Python analytics over \texttt{pandas}
DataFrames.  Compared with Trino and Oracle, it avoids loading data
into a database server before running
\texttt{MATCH\_\allowbreak{}RECOGNIZE}, which is convenient for
notebook and DataFrame-centred analysis, but it supports only the
subset described in Chapters~\ref{chap:methodology}
and~\ref{chap:implementation}; Trino and Oracle remain more complete
SQL systems with mature optimizers, distributed execution, and
broader standard coverage.  Compared with hand-written Pandas logic,
the SQL form is more declarative and reusable: partitioning,
ordering, pattern variables, skip behaviour, and measures are
expressed in one query rather than in custom procedural code for each
pattern.

\paragraph{Applicability and reference validation.}
The implementation applies to several sequential-analysis settings,
such as log analysis, time-series processing, and financial sequence
analysis, where recurring row patterns are important.  Across all 60
cross-system cells the proposed engine's output was compared
row-by-row against the two SQL references, Trino~473 and
Oracle~XE~21c, and matched them in every case; the hand-written
Python-UDF baseline (Section~\ref{subsec:eval-usability}) is reported
only as a usability contrast---it required substantially more code and
was more prone to errors during development~\cite{Python_UDF}---not as
a correctness oracle.

% ................................................................
\subsection{Pattern Definitions}
\label{subsec:eval-pattern-defs}
% ................................................................

{\small
\begin{longtable}{llp{0.55\textwidth}}
\caption{SQL Pattern Definitions and Detection Goals}
\label{tab:pattern_definitions} \\
\toprule
\textbf{Pattern Name} & \textbf{SQL Pattern} & \textbf{Description \& Goal} \\
\midrule
\endfirsthead
\toprule
\textbf{Pattern Name} & \textbf{SQL Pattern} & \textbf{Description \& Goal} \\
\midrule
\endhead
\bottomrule
\endfoot

simple\_sequence & \texttt{A+ B+} &
\textbf{Goal:} Detect transitions from budget to low-mid price
listings.
\newline
\textbf{Use:} Identify runs where budget products
(price~$\le$~\pounds10) are immediately followed by low-mid products
(\pounds10--25) in the listing order, useful for locating price-band
boundaries and clustered cheap listings. \\
\midrule

alternation & \texttt{A (B|C)+ D} &
\textbf{Goal:} Detect budget-to-upper-mid transitions through mixed
middle bands.
\newline
\textbf{Use:} Find sequences that start with a budget product,
continue with a mix of low-mid and mid products (\pounds10--50), and
end with an upper-mid product (\pounds50--100).  Exercises
alternation inside a quantified group. \\
\midrule

quantified & \texttt{A\{1,5\} B* C+} &
\textbf{Goal:} Detect bounded budget runs followed by mid-band
products.
\newline
\textbf{Use:} Find sequences with 1--5 budget products, optionally
followed by low-mid products, then one or more mid products
(\pounds25--50).  Exercises bounded, optional, and unbounded
quantifiers in one pattern. \\
\midrule

optional\_pattern & \texttt{A+ B? C*} &
\textbf{Goal:} Flexible budget-run detection with optional band
transitions.
\newline
\textbf{Use:} Broad matching of budget runs with an optional low-mid
product and optional mid products.  High recall; produces the
largest match sets in the benchmark. \\
\midrule

complex\_nested & \texttt{(A|B)+ (C\{1,3\} D*)+} &
\textbf{Goal:} Nested grouping of low-price runs followed by
mid/upper-mid groups.
\newline
\textbf{Use:} Detect runs of low-priced products
(price~$\le$~\pounds25) followed by one or more groups of 1--3 mid
products with optional upper-mid products.  Exercises nested
alternation combined with bounded repetition; the most expensive
pattern family. \\
\end{longtable}
}

\noindent
All patterns operate on a derived \texttt{category} column computed
from the product \texttt{price} with fixed thresholds:
$A$~= budget (price~$\le$~\pounds10),
$B$~= low-mid (\pounds10~$<$~price~$\le$~\pounds25),
$C$~= mid (\pounds25~$<$~price~$\le$~\pounds50),
$D$~= upper-mid (\pounds50~$<$~price~$\le$~\pounds100), and
$E$~= premium (price~$>$~\pounds100).
Each row also carries a sequential \texttt{seq\_id} that preserves
the original row order of the source CSV.  The six dataset sizes
(100\,K, 200\,K, 400\,K, 800\,K, 1.6\,M, and 2{,}222{,}742 rows) are
nested prefixes of the same file: the size-$n$ dataset is the first
$n$ rows of the source CSV (no random sampling, so no seed is
involved; the largest size is the entire 2{,}222{,}742-row dataset),
and \texttt{seq\_id} is assigned as $0,1,\dots,n-1$ in that order.  The benchmark queries use \texttt{ORDER~BY seq\_id} and
no \texttt{PARTITION~BY} clause, so each dataset forms a single
logical partition and the patterns match runs of consecutive rows in
this fixed order.  The sequence is
therefore a \emph{controlled benchmark sequence} derived from the
CSV listing order of the product catalog; it is not a temporal event
sequence, and no claim is made that adjacent rows are related in
time.  What matters for the benchmark is that all three systems
process the identical row sequence.
Table~\ref{tab:pattern_definitions} provides detailed
specifications of all five SQL patterns used in the evaluation,
including their formal syntax and intended detection goals.  Each
pattern represents a different complexity level and use case, ranging
from simple sequential matching to complex nested structures with
quantifiers and alternation operators.

\paragraph{Benchmark setup and resource policy.}
The empirical analysis measures the implemented systems directly.  In
this chapter, the final benchmark consists of 90 successful
benchmark cells: three systems, five pattern families, and six
dataset sizes from 100\,K to 2.22\,M rows, where each cell is the mean
of twenty measured executions after five warmups (1{,}800 measured
executions in total).  The proposed Pandas engine, Trino~473, and
Oracle~XE~21c were evaluated on the same prepared CSV subsets (nested
prefixes of the source dataset) and the same
\texttt{MATCH\_\allowbreak{}RECOGNIZE} query bodies.

The resource policy was enforced as follows.  The three systems ran
strictly sequentially, one at a time, with the other database
containers stopped.  The Pandas engine was pinned to CPU core~0 via
\texttt{sched\_setaffinity}, with the numerical-library
thread pools (\texttt{OMP}, \texttt{OpenBLAS}, \texttt{MKL},
\texttt{NumExpr}) limited to one thread; the database containers
were pinned to the same core with
\texttt{docker update -{}-cpuset-cpus=0} combined with a one-CPU
quota (\texttt{-{}-cpus=1}).  Every system was given the same
nominal 32\,GB memory budget, enforced by the mechanism native to
each deployment: an address-space \texttt{rlimit} for the Python
process, and a container memory limit with \texttt{-{}-memory-swap}
equal to the limit (no extra swap) for the databases.  These
mechanisms are equivalent in budget but not identical in enforcement
semantics.  The 32\,GB figure is therefore an
\emph{engine/container} memory budget: it caps the Pandas engine
process and each database container.  The budget is below the
workstation's 64\,GB of installed RAM, so it is an actually enforced
cap.  For the database runs, the separate Python client that issues
the query and materializes the result DataFrame is pinned to core~0
via \texttt{sched\_setaffinity} (the same call that pins the Pandas
engine), applied once to the benchmark process at start-up; container
pinning alone would not constrain this host-side client, and the
client's own memory is not itself placed under the 32\,GB
\texttt{rlimit}.  Client and database engine therefore time-share the
single core~0, and, correspondingly, the container footprint values
cover only the database process and exclude this client, whose
result-construction memory is counted in neither database memory
metric.
Reported times are client-observed wall-clock
durations of query execution including result retrieval, measured
tightly around the query call so that sampling overhead was
minimized.  The timer
stops only after the client has fully materialized the result set
(fetching all rows and constructing the client-side DataFrame), and
every system returns the same normalized result columns and row
counts for each cell (Table~\ref{tab:match_counts}), so result
retrieval is included in the timing on equal terms.  Database
loading time is excluded from the reported query times; the
measurements report query execution after the data has been loaded
into the target system.

The theoretical model explains the expected growth, while the empirical
results show the current implementation's observed behavior.  The
Pandas engine shows stable throughput across the evaluated range,
which is consistent with a scan-dominated execution path for these
fixed-size patterns.  In this run it records the lowest average
execution time of the three systems on every one of the 30
pattern--size cells: averaged per family it is roughly
6--13$\times$ faster than Trino and 2.7--5.5$\times$ faster than
Oracle, with the largest gap on \texttt{complex\_nested}.  Memory is reported using the two metrics defined in
Section~\ref{subsec:eval-metrics}: the engine-reported
system-native \emph{query-memory indicators}, and the
\emph{operational footprint}, which documents the cost of the
resident process or database container and is reported separately.

% ................................................................
\subsection{Cross-System Comparison Under a Common Resource Budget}
\label{subsec:eval-cross-system}
% ................................................................

The benchmark artifacts were saved externally for reproducibility in
\texttt{Performance/cross\_system\_matrix}: one result file per system
(\texttt{pandas\_results.csv}, \texttt{trino\_results.csv}, and
\texttt{oracle\_results.csv}), the shared input datasets, the saved SQL
query files, and combined summary files.  The common query body for
each pattern is stored once and the per-system query files differ only
in the source table reference required by each system.

Because runtime and memory depend directly on how many matches each
pattern produces, the match count for every pattern--size combination
is reported in Table~\ref{tab:match_counts}
(Appendix~\ref{app:benchmark-tables}).  All queries use
\texttt{ONE ROW PER MATCH}, so the match count equals the number of
returned result rows; the counts are identical across the three
systems (this is exactly what the correctness comparison verifies).
\texttt{optional\_pattern} produces the largest output sets and
\texttt{alternation} the smallest, which explains much of the
per-pattern variation in the time and memory tables that follow.

% Final run (each cell = 5 warmup + 20 measured; 1 core, 32GB; sizes 100K-2.22M; mean +/- sample std)
% pandas: total=5.96, avg=0.20, avgthr=5{,}043{,}941, minthr=1{,}791{,}628, maxthr=8{,}055{,}323, maxqmem=173.81, maxfmem=665.63, medCV=8.8%
% Trino~473: total=57.88, avg=1.93, avgthr=609{,}122, minthr=153{,}963, maxthr=1{,}009{,}725, maxqmem=41.50, maxfmem=23{,}225.34, medCV=5.9%
% Oracle XE~21c: total=28.21, avg=0.94, avgthr=1{,}258{,}336, minthr=359{,}113, maxthr=1{,}802{,}506, maxqmem=58.19, maxfmem=2{,}523.6, medCV=2.1%

Table~\ref{tab:pattern_summary} (Appendix~\ref{app:benchmark-tables})
summarizes each pattern family across
all six dataset sizes.  The proposed engine records the lowest
average time on all five pattern families in this run, against both
database systems.  The largest gap is on \texttt{complex\_nested},
the family where Trino shows by far its largest relative cost
(5.04\,s versus 0.40\,s on average, about 12.6$\times$).  This
indicates that both implementation strategy and pattern structure
affect the measured runtime.

Table~\ref{tab:size_summary} (Appendix~\ref{app:benchmark-tables})
groups the same measurements by dataset
size.  The proposed engine records the lowest average time at every
size, and its throughput remains high (per-size averages of roughly
3.6\,M--5.6\,M rows/s) as the input increases.  Relative to a
fully-warmed Trino, the engine's per-size averages are lower at all
six sizes, by a factor of approximately 7.3--10.6$\times$.
Relative to Oracle, the engine's per-size averages are also lower
across the evaluated range, by a factor of approximately
3.1--5.2$\times$, using the
per-size averages of Table~\ref{tab:size_summary}.

{\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{llrrr}
\caption{Execution time (s), mean $\pm$ sample standard deviation
  over the twenty measured runs, by pattern, dataset size, and system}
\label{tab:execution_times} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endfirsthead
\caption*{Table~\thetable{} (continued): Execution time (s, mean $\pm$ std) by pattern, size, and system} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endhead
\bottomrule
\endfoot
simple\_sequence & 100{,}000 & 0.02 $\pm$ 0.01 & 0.28 $\pm$ 0.06 & 0.22 $\pm$ 0.18 \\
simple\_sequence & 200{,}000 & 0.03 $\pm$ 0.00 & 0.32 $\pm$ 0.12 & 0.14 $\pm$ 0.01 \\
simple\_sequence & 400{,}000 & 0.05 $\pm$ 0.00 & 0.67 $\pm$ 0.16 & 0.34 $\pm$ 0.01 \\
simple\_sequence & 800{,}000 & 0.11 $\pm$ 0.02 & 1.20 $\pm$ 0.15 & 0.65 $\pm$ 0.01 \\
simple\_sequence & 1{,}600{,}000 & 0.22 $\pm$ 0.03 & 2.16 $\pm$ 0.12 & 1.26 $\pm$ 0.01 \\
simple\_sequence & 2{,}222{,}742 & 0.30 $\pm$ 0.01 & 2.95 $\pm$ 0.13 & 1.77 $\pm$ 0.01 \\
\midrule
alternation & 100{,}000 & 0.03 $\pm$ 0.00 & 0.30 $\pm$ 0.29 & 0.06 $\pm$ 0.00 \\
alternation & 200{,}000 & 0.05 $\pm$ 0.00 & 0.32 $\pm$ 0.10 & 0.11 $\pm$ 0.00 \\
alternation & 400{,}000 & 0.09 $\pm$ 0.01 & 0.55 $\pm$ 0.13 & 0.22 $\pm$ 0.00 \\
alternation & 800{,}000 & 0.16 $\pm$ 0.01 & 0.92 $\pm$ 0.03 & 0.44 $\pm$ 0.01 \\
alternation & 1{,}600{,}000 & 0.31 $\pm$ 0.01 & 1.82 $\pm$ 0.07 & 0.89 $\pm$ 0.01 \\
alternation & 2{,}222{,}742 & 0.44 $\pm$ 0.02 & 2.49 $\pm$ 0.07 & 1.24 $\pm$ 0.01 \\
\midrule
quantified & 100{,}000 & 0.02 $\pm$ 0.00 & 0.24 $\pm$ 0.02 & 0.09 $\pm$ 0.00 \\
quantified & 200{,}000 & 0.04 $\pm$ 0.01 & 0.36 $\pm$ 0.09 & 0.16 $\pm$ 0.00 \\
quantified & 400{,}000 & 0.06 $\pm$ 0.01 & 0.61 $\pm$ 0.04 & 0.33 $\pm$ 0.01 \\
quantified & 800{,}000 & 0.10 $\pm$ 0.01 & 1.23 $\pm$ 0.04 & 0.72 $\pm$ 0.01 \\
quantified & 1{,}600{,}000 & 0.21 $\pm$ 0.03 & 2.47 $\pm$ 0.15 & 1.40 $\pm$ 0.02 \\
quantified & 2{,}222{,}742 & 0.31 $\pm$ 0.04 & 3.21 $\pm$ 0.08 & 1.95 $\pm$ 0.03 \\
\midrule
optional\_pattern & 100{,}000 & 0.02 $\pm$ 0.00 & 0.20 $\pm$ 0.03 & 0.07 $\pm$ 0.00 \\
optional\_pattern & 200{,}000 & 0.04 $\pm$ 0.00 & 0.29 $\pm$ 0.07 & 0.13 $\pm$ 0.01 \\
optional\_pattern & 400{,}000 & 0.07 $\pm$ 0.01 & 0.46 $\pm$ 0.02 & 0.25 $\pm$ 0.01 \\
optional\_pattern & 800{,}000 & 0.13 $\pm$ 0.01 & 0.81 $\pm$ 0.03 & 0.47 $\pm$ 0.02 \\
optional\_pattern & 1{,}600{,}000 & 0.29 $\pm$ 0.04 & 1.59 $\pm$ 0.04 & 0.92 $\pm$ 0.03 \\
optional\_pattern & 2{,}222{,}742 & 0.46 $\pm$ 0.05 & 2.20 $\pm$ 0.04 & 1.29 $\pm$ 0.04 \\
\midrule
complex\_nested & 100{,}000 & 0.06 $\pm$ 0.00 & 0.29 $\pm$ 0.02 & 0.11 $\pm$ 0.01 \\
complex\_nested & 200{,}000 & 0.10 $\pm$ 0.01 & 0.54 $\pm$ 0.04 & 0.24 $\pm$ 0.01 \\
complex\_nested & 400{,}000 & 0.18 $\pm$ 0.02 & 1.25 $\pm$ 0.10 & 0.57 $\pm$ 0.04 \\
complex\_nested & 800{,}000 & 0.37 $\pm$ 0.01 & 4.80 $\pm$ 0.11 & 2.14 $\pm$ 0.02 \\
complex\_nested & 1{,}600{,}000 & 0.70 $\pm$ 0.01 & 10.39 $\pm$ 0.16 & 4.46 $\pm$ 0.09 \\
complex\_nested & 2{,}222{,}742 & 0.99 $\pm$ 0.01 & 12.96 $\pm$ 0.25 & 5.59 $\pm$ 0.08 \\
\end{longtable}
}

Table~\ref{tab:execution_times} reports the detailed execution-time
matrix.  The proposed engine grows predictably with input size, and
the relative cost of each pattern depends on the matcher path selected
for that pattern.  For Trino and Oracle, \texttt{complex\_nested}
remains the most expensive pattern family.

% ................................................................
\subsection{Throughput by Pattern, Size, and System}
\label{subsec:eval-throughput}
% ................................................................

{\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{llrrr}
\caption{Throughput (rows/s) by pattern, dataset size, and system}
\label{tab:throughput} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endfirsthead
\caption*{Table~\thetable{} (continued): Throughput (rows/s) by pattern, size, and system} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endhead
\bottomrule
\endfoot
simple\_sequence & 100{,}000 & 4{,}557{,}285 & 357{,}076 & 455{,}079 \\
simple\_sequence & 200{,}000 & 7{,}066{,}948 & 624{,}322 & 1{,}442{,}657 \\
simple\_sequence & 400{,}000 & 7{,}864{,}039 & 599{,}042 & 1{,}186{,}792 \\
simple\_sequence & 800{,}000 & 7{,}051{,}074 & 667{,}059 & 1{,}234{,}424 \\
simple\_sequence & 1{,}600{,}000 & 7{,}345{,}525 & 740{,}875 & 1{,}271{,}565 \\
simple\_sequence & 2{,}222{,}742 & 7{,}517{,}413 & 752{,}286 & 1{,}258{,}944 \\
\midrule
alternation & 100{,}000 & 3{,}311{,}689 & 338{,}621 & 1{,}641{,}751 \\
alternation & 200{,}000 & 4{,}198{,}875 & 634{,}194 & 1{,}783{,}969 \\
alternation & 400{,}000 & 4{,}441{,}633 & 722{,}472 & 1{,}791{,}942 \\
alternation & 800{,}000 & 4{,}901{,}958 & 866{,}742 & 1{,}802{,}506 \\
alternation & 1{,}600{,}000 & 5{,}106{,}703 & 881{,}308 & 1{,}792{,}047 \\
alternation & 2{,}222{,}742 & 5{,}023{,}855 & 893{,}397 & 1{,}794{,}710 \\
\midrule
quantified & 100{,}000 & 4{,}442{,}142 & 417{,}568 & 1{,}137{,}354 \\
quantified & 200{,}000 & 5{,}002{,}090 & 551{,}188 & 1{,}252{,}330 \\
quantified & 400{,}000 & 6{,}350{,}487 & 651{,}060 & 1{,}209{,}887 \\
quantified & 800{,}000 & 8{,}055{,}323 & 649{,}590 & 1{,}116{,}448 \\
quantified & 1{,}600{,}000 & 7{,}582{,}348 & 647{,}215 & 1{,}140{,}903 \\
quantified & 2{,}222{,}742 & 7{,}243{,}840 & 691{,}696 & 1{,}138{,}626 \\
\midrule
optional\_pattern & 100{,}000 & 4{,}141{,}043 & 494{,}680 & 1{,}390{,}356 \\
optional\_pattern & 200{,}000 & 5{,}527{,}191 & 696{,}430 & 1{,}571{,}544 \\
optional\_pattern & 400{,}000 & 5{,}417{,}165 & 868{,}398 & 1{,}615{,}209 \\
optional\_pattern & 800{,}000 & 6{,}035{,}844 & 984{,}378 & 1{,}702{,}193 \\
optional\_pattern & 1{,}600{,}000 & 5{,}555{,}254 & 1{,}009{,}358 & 1{,}738{,}158 \\
optional\_pattern & 2{,}222{,}742 & 4{,}881{,}440 & 1{,}009{,}725 & 1{,}728{,}719 \\
\midrule
complex\_nested & 100{,}000 & 1{,}791{,}628 & 343{,}097 & 876{,}493 \\
complex\_nested & 200{,}000 & 2{,}025{,}051 & 368{,}396 & 838{,}858 \\
complex\_nested & 400{,}000 & 2{,}188{,}285 & 321{,}233 & 705{,}413 \\
complex\_nested & 800{,}000 & 2{,}161{,}881 & 166{,}814 & 374{,}437 \\
complex\_nested & 1{,}600{,}000 & 2{,}274{,}674 & 153{,}963 & 359{,}113 \\
complex\_nested & 2{,}222{,}742 & 2{,}255{,}543 & 171{,}492 & 397{,}651 \\
\end{longtable}
}

Table~\ref{tab:throughput} reports throughput for every pattern--size
combination.  Throughput is computed as dataset size divided by
measured execution time, so higher values indicate faster evaluation.

% ................................................................
\subsection{Memory Metrics by Pattern, Size, and System}
\label{subsec:eval-memory}
% ................................................................

Memory is reported using the two metrics defined in
Section~\ref{subsec:eval-metrics}.  \emph{Query memory}
(Table~\ref{tab:memory_cache}) is the peak additional memory
attributable to executing the measured query, taken from each
system's own memory accounting: process RSS peak delta for the
Pandas engine, the per-query \texttt{peakMemoryBytes} counter for
Trino, and the session PGA high-water delta of a fresh session for
Oracle.  All three exclude the stored table data and the idle
engine.  They are, however, \emph{system-native indicators} rather
than one instrument: RSS counts operating-system pages, Trino counts
reserved user memory, and Oracle counts allocated work-area memory.
The values should therefore be read as order-of-magnitude and
per-system trend evidence, not as directly comparable measurements
of identical quantities.
\emph{Footprint memory} (Table~\ref{tab:memory_footprint}) is the
absolute process RSS peak for the Pandas engine and the Docker
container peak for Trino and Oracle; it documents the practical
deployment cost of the resident engine and is \emph{not} a
per-query or algorithm-level metric.

{\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{llrrr}
\caption{Query memory (MB) by pattern, dataset size, and system
  (engine-reported peak memory of the measured query, mean of the
  twenty per-run peaks)}
\label{tab:memory_cache} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endfirsthead
\caption*{Table~\thetable{} (continued): Query memory (MB) by pattern, size, and system} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endhead
\bottomrule
\endfoot
simple\_sequence & 100{,}000 & 1.34 & 1.24 & 1.24 \\
simple\_sequence & 200{,}000 & 2.14 & 2.80 & 3.75 \\
simple\_sequence & 400{,}000 & 3.18 & 6.53 & 9.06 \\
simple\_sequence & 800{,}000 & 6.52 & 13.61 & 19.69 \\
simple\_sequence & 1{,}600{,}000 & 8.87 & 28.70 & 40.69 \\
simple\_sequence & 2{,}222{,}742 & 160.03 & 41.48 & 58.19 \\
\midrule
alternation & 100{,}000 & 0.05 & 1.24 & 1.19 \\
alternation & 200{,}000 & 0.00 & 2.62 & 3.75 \\
alternation & 400{,}000 & 0.00 & 5.77 & 9.06 \\
alternation & 800{,}000 & 0.00 & 12.86 & 19.69 \\
alternation & 1{,}600{,}000 & 0.90 & 28.63 & 40.69 \\
alternation & 2{,}222{,}742 & 0.99 & 41.42 & 58.19 \\
\midrule
quantified & 100{,}000 & 0.53 & 1.46 & 1.19 \\
quantified & 200{,}000 & 2.64 & 3.08 & 3.75 \\
quantified & 400{,}000 & 4.10 & 7.01 & 9.06 \\
quantified & 800{,}000 & 5.19 & 13.83 & 19.69 \\
quantified & 1{,}600{,}000 & 4.75 & 28.76 & 40.69 \\
quantified & 2{,}222{,}742 & 164.31 & 41.42 & 58.19 \\
\midrule
optional\_pattern & 100{,}000 & 0.96 & 1.34 & 1.19 \\
optional\_pattern & 200{,}000 & 4.45 & 2.85 & 3.75 \\
optional\_pattern & 400{,}000 & 2.87 & 5.90 & 9.06 \\
optional\_pattern & 800{,}000 & 3.32 & 10.95 & 19.69 \\
optional\_pattern & 1{,}600{,}000 & 35.68 & 28.62 & 40.69 \\
optional\_pattern & 2{,}222{,}742 & 173.81 & 41.40 & 58.19 \\
\midrule
complex\_nested & 100{,}000 & 2.09 & 1.18 & 1.19 \\
complex\_nested & 200{,}000 & 1.17 & 2.79 & 3.75 \\
complex\_nested & 400{,}000 & 4.64 & 7.89 & 9.06 \\
complex\_nested & 800{,}000 & 8.00 & 13.63 & 19.69 \\
complex\_nested & 1{,}600{,}000 & 16.75 & 28.75 & 40.69 \\
complex\_nested & 2{,}222{,}742 & 42.01 & 41.50 & 58.19 \\
\end{longtable}
}

Table~\ref{tab:memory_cache} shows that the average query-memory
indicators generally increase with input size, as expected for a
per-query working set.  For the most memory-intensive pattern
families at the largest input (the full 2.22\,M rows), the three
indicators fall between roughly 41 and 174\,MB; this range describes
only those worst-case cells, not the benchmark as a whole, where most
readings are far smaller.  At 2.22\,M rows the Pandas engine's
vectorized execution path trades additional working memory for its
lower execution times on three families---its largest indicator is
on \texttt{optional\_pattern} (173.81\,MB), followed by
\texttt{quantified} (164.31\,MB) and \texttt{simple\_sequence}
(160.03\,MB), each above Oracle's 58.19\,MB and Trino's roughly
41.5\,MB---whereas on the other two families its readings stay at or
below the databases' (\texttt{complex\_nested} at 42.01\,MB and
\texttt{alternation} at the measurement floor, against Trino's
41.4--41.5\,MB and Oracle's 58.19\,MB).

Values of 0.00 in Table~\ref{tab:memory_cache} do not mean that a
query executed without memory; they mean that the query's working
set stayed below the resolution of the corresponding counter.
Oracle pre-allocates an initial block of PGA to every new session at
connect time, so a small query whose \texttt{MATCH\_\allowbreak{}RECOGNIZE}
work areas fit inside that initial allocation never raises the
session's PGA high-water mark, and the measured delta is zero.
Likewise, a Pandas run that fits into pages already resident in the
Python process produces no RSS growth.  Readings close to this floor
are also quantized to the runtime's allocation granularity (Oracle,
for example, extends session PGA in small fixed-size chunks), so
near-floor values should not be compared against each other.  At larger input sizes the database counters
grow steadily with the input, while the RSS-based Pandas readings
fluctuate more (values can rise, fall, or return to zero between
neighboring cells) because RSS deltas depend on page reuse and
garbage-collection timing inside the Python process; the overall
upward trend beyond the measurement floor is what the metric is
intended to show.

{\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{longtable}{llrrr}
\caption{Footprint memory (MB) by pattern, dataset size, and system
  (process RSS peak for Pandas; peak container memory sampled during
  the measured run for Trino/Oracle, whose containers keep running
  across dataset sizes within a pass; each value is the mean of the
  twenty per-run peaks)}
\label{tab:memory_footprint} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endfirsthead
\caption*{Table~\thetable{} (continued): Footprint memory (MB) by pattern, size, and system} \\
\toprule
\textbf{Pattern} & \textbf{Size (rows)} & \textbf{Pandas} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
\endhead
\bottomrule
\endfoot
simple\_sequence & 100{,}000 & 155.95 & 8{,}320.72 & 2{,}523.6 \\
simple\_sequence & 200{,}000 & 182.35 & 9{,}180.11 & 2{,}224.13 \\
simple\_sequence & 400{,}000 & 227.79 & 10{,}453.5 & 2{,}228.84 \\
simple\_sequence & 800{,}000 & 323.32 & 12{,}917.25 & 2{,}219.88 \\
simple\_sequence & 1{,}600{,}000 & 507.72 & 16{,}923.65 & 2{,}225.82 \\
simple\_sequence & 2{,}222{,}742 & 645.45 & 23{,}130.11 & 2{,}245.99 \\
\midrule
alternation & 100{,}000 & 149.14 & 8{,}368.18 & 2{,}307.12 \\
alternation & 200{,}000 & 182.07 & 9{,}231.36 & 2{,}224.54 \\
alternation & 400{,}000 & 226.23 & 10{,}626.56 & 2{,}229.2 \\
alternation & 800{,}000 & 318.72 & 13{,}210.62 & 2{,}219.93 \\
alternation & 1{,}600{,}000 & 448.96 & 17{,}058.82 & 2{,}225.51 \\
alternation & 2{,}222{,}742 & 445.44 & 23{,}203.84 & 2{,}240.36 \\
\midrule
quantified & 100{,}000 & 156.47 & 8{,}651.16 & 2{,}187.78 \\
quantified & 200{,}000 & 184.05 & 9{,}309.9 & 2{,}266.47 \\
quantified & 400{,}000 & 229.84 & 10{,}836.99 & 2{,}229.25 \\
quantified & 800{,}000 & 323.14 & 13{,}230.08 & 2{,}219.88 \\
quantified & 1{,}600{,}000 & 503.56 & 17{,}062.4 & 2{,}225.36 \\
quantified & 2{,}222{,}742 & 658.42 & 23{,}212.54 & 2{,}304.41 \\
\midrule
optional\_pattern & 100{,}000 & 156.01 & 8{,}656.79 & 2{,}187.72 \\
optional\_pattern & 200{,}000 & 184.98 & 9{,}410.05 & 2{,}243.23 \\
optional\_pattern & 400{,}000 & 232.13 & 10{,}865.15 & 2{,}220.34 \\
optional\_pattern & 800{,}000 & 324.95 & 13{,}240.32 & 2{,}219.67 \\
optional\_pattern & 1{,}600{,}000 & 500.74 & 17{,}070.08 & 2{,}225.25 \\
optional\_pattern & 2{,}222{,}742 & 665.63 & 23{,}214.08 & 2{,}254.34 \\
\midrule
complex\_nested & 100{,}000 & 152.95 & 8{,}660.99 & 2{,}201.4 \\
complex\_nested & 200{,}000 & 184.92 & 9{,}582.13 & 2{,}230.89 \\
complex\_nested & 400{,}000 & 230.45 & 10{,}875.9 & 2{,}218.85 \\
complex\_nested & 800{,}000 & 318.66 & 13{,}250.56 & 2{,}239.49 \\
complex\_nested & 1{,}600{,}000 & 481.85 & 17{,}080.32 & 2{,}281.88 \\
complex\_nested & 2{,}222{,}742 & 501.71 & 23{,}225.34 & 2{,}298.27 \\
\end{longtable}
}

Table~\ref{tab:memory_footprint} shows the practical deployment
footprints.  The boundaries differ by construction: the Pandas value
covers one Python process including the loaded input DataFrame,
whereas the database values cover entire Docker containers.  The
Pandas engine peaks at about 0.67\,GB at the full 2.22\,M rows.  Trino's
container grows to about 23\,GB: the benchmark table is held by its
memory connector inside the JVM heap, the table is dropped and
recreated before each dataset size is loaded, and the container
keeps running across all sizes within the system's benchmark pass
(it is restarted only between systems).  Each database value is the
peak container memory sampled during that measured run, not a
monotonic high-water counter: because the JVM rarely returns freed
heap to the operating system, the readings carry allocations
retained from earlier sizes, yet they can also fluctuate downward
between neighboring cells because the container-level usage metric
excludes reclaimable page cache that the kernel may free at any
time.  Oracle's container stays near 2.4--2.5\,GB.  The two kinds
of values therefore have different reset boundaries: the Pandas
readings are per-run process peaks, while the database readings are
per-run samples of a container whose memory state resets only when
the container is restarted at the start of the pass, and they must
not be interpreted as independent per-size footprints.  Restarting each container before every dataset
size would yield independent per-size footprint measurements and is
the recommended refinement for future benchmark runs; it was not
applied here to keep the pass duration tractable.  These values
document what it costs to have each system running as deployed here;
per-query comparisons should use Table~\ref{tab:memory_cache}.

% ................................................................
\subsection{Key Findings}
\label{subsec:eval-key-findings}
% ................................................................

\begin{table}[H]
\centering
\caption{Overall benchmark test statistics by system.  Footprint
  values have different reset boundaries: per-run process peaks for
  the Pandas engine, per-run peak samples of containers that persist
  across dataset sizes for Trino/Oracle
  (Section~\ref{subsec:eval-memory}).}
\label{tab:overall_stats}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrrrrr}
\toprule
\textbf{System} & \textbf{Tests} & \textbf{Sum of Mean} & \textbf{Avg Time} & \textbf{Avg Thr.} & \textbf{Thr. Range} & \textbf{Max Query} & \textbf{Max Footpr.} \\
& & \textbf{Times (s)} & \textbf{(s)} & \textbf{(rows/s)} & \textbf{(rows/s)} & \textbf{Mem.\ (MB)} & \textbf{Mem.\ (MB)} \\
\midrule
Proposed Pandas engine & 30 & 5.96 & 0.20 & 5{,}043{,}941 & 1{,}791{,}628--8{,}055{,}323 & 173.81 & 665.63 \\
Trino~473 & 30 & 57.88 & 1.93 & 609{,}122 & 153{,}963--1{,}009{,}725 & 41.50 & 23{,}225.34 \\
Oracle XE~21c & 30 & 28.21 & 0.94 & 1{,}258{,}336 & 359{,}113--1{,}802{,}506 & 58.19 & 2{,}523.6 \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\centering
\caption{Cross-system benchmark summary under 1 CPU core and a
  32\,GB memory limit.  Footprint values have different reset
  boundaries: per-run process peaks for the Pandas engine, per-run
  peak samples of containers that persist across dataset sizes for
  Trino/Oracle.}
\label{tab:cross_system_summary}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrrrr}
\toprule
\textbf{System} & \textbf{Tests} & \textbf{Avg Time} & \textbf{Avg Throughput} & \textbf{Max Query} & \textbf{Max Footpr.} & \textbf{Correctness} \\
& & \textbf{(s)} & \textbf{(rows/s)} & \textbf{Mem.\ (MB)} & \textbf{Mem.\ (MB)} & \\
\midrule
Proposed Pandas engine & 30 & 0.20 & 5{,}043{,}941 & 173.81 & 665.63 & proposed output \\
Trino~473 & 30 & 1.93 & 609{,}122 & 41.50 & 23{,}225.34 & 30/30 identical \\
Oracle XE~21c & 30 & 0.94 & 1{,}258{,}336 & 58.19 & 2{,}523.6 & 30/30 identical \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[H]
\centering
\caption{Average execution time by pattern under 1 CPU core and a 32\,GB memory limit}
\label{tab:avg_time_by_pattern}
\begin{tabular}{lrrr}
\toprule
\textbf{Pattern} & \textbf{Pandas (s)} & \textbf{Trino~473 (s)} & \textbf{Oracle XE~21c (s)} \\
\midrule
simple\_sequence & 0.12 & 1.26 & 0.73 \\
alternation & 0.18 & 1.07 & 0.50 \\
quantified & 0.12 & 1.36 & 0.77 \\
optional\_pattern & 0.17 & 0.92 & 0.52 \\
complex\_nested & 0.40 & 5.04 & 2.18 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Average execution time by dataset size under 1 CPU core and a 32\,GB memory limit}
\label{tab:avg_time_by_size}
\begin{tabular}{rrrr}
\toprule
\textbf{Size (rows)} & \textbf{Pandas (s)} & \textbf{Trino~473 (s)} & \textbf{Oracle XE~21c (s)} \\
\midrule
100{,}000 & 0.03 & 0.26 & 0.11 \\
200{,}000 & 0.05 & 0.37 & 0.16 \\
400{,}000 & 0.09 & 0.71 & 0.34 \\
800{,}000 & 0.18 & 1.79 & 0.88 \\
1{,}600{,}000 & 0.35 & 3.68 & 1.79 \\
2{,}222{,}742 & 0.50 & 4.76 & 2.37 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Average query memory by dataset size under 1 CPU core and a 32\,GB memory limit}
\label{tab:avg_memory_by_size}
\begin{tabular}{rrrr}
\toprule
\textbf{Size (rows)} & \textbf{Pandas (MB)} & \textbf{Trino~473 (MB)} & \textbf{Oracle XE~21c (MB)} \\
\midrule
100{,}000 & 0.99 & 1.29 & 1.20 \\
200{,}000 & 2.08 & 2.83 & 3.75 \\
400{,}000 & 2.96 & 6.62 & 9.06 \\
800{,}000 & 4.60 & 12.98 & 19.69 \\
1{,}600{,}000 & 13.39 & 28.69 & 40.69 \\
2{,}222{,}742 & 108.23 & 41.44 & 58.19 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Average footprint memory by dataset size under 1 CPU core and a 32\,GB memory limit}
\label{tab:avg_footprint_by_size}
\begin{tabular}{rrrr}
\toprule
\textbf{Size (rows)} & \textbf{Pandas (MB)} & \textbf{Trino~473 (MB)} & \textbf{Oracle XE~21c (MB)} \\
\midrule
100{,}000 & 154.11 & 8{,}531.57 & 2{,}281.52 \\
200{,}000 & 183.68 & 9{,}342.71 & 2{,}237.85 \\
400{,}000 & 229.29 & 10{,}731.62 & 2{,}225.3 \\
800{,}000 & 321.76 & 13{,}169.77 & 2{,}223.77 \\
1{,}600{,}000 & 488.57 & 17{,}039.05 & 2{,}236.76 \\
2{,}222{,}742 & 583.33 & 23{,}197.18 & 2{,}268.67 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Cross-system correctness by pattern (6 dataset sizes each)}
\label{tab:cross_system_correctness}
\begin{tabular}{lccc}
\toprule
\textbf{Pattern} & \textbf{Pandas Engine} & \textbf{Trino~473} & \textbf{Oracle XE~21c} \\
\midrule
simple\_sequence & proposed output & 6/6 identical & 6/6 identical \\
alternation & proposed output & 6/6 identical & 6/6 identical \\
quantified & proposed output & 6/6 identical & 6/6 identical \\
optional\_pattern & proposed output & 6/6 identical & 6/6 identical \\
complex\_nested & proposed output & 6/6 identical & 6/6 identical \\
\bottomrule
\end{tabular}
\end{table}


Tables~\ref{tab:overall_stats}--\ref{tab:cross_system_correctness}
summarize the main findings.  All 90 benchmark cells completed
successfully, and Trino and Oracle produced output identical to the
proposed engine's normalized output for all 30 pattern--size
combinations each---all three systems produced identical result
sets.  The proposed Pandas engine had the lowest average
execution time in this benchmark (0.20\,s), ahead of Oracle
(0.94\,s) and a fully-warmed Trino (1.93\,s): it recorded the
lowest time on all 30 pattern--size cells, faster than Trino by
roughly 6--13$\times$ per family and faster than Oracle by roughly
2.7--5.5$\times$, with the widest margins on \texttt{complex\_nested}
(where Trino averages 5.04\,s against the engine's 0.40\,s).  The
system-native
query-memory indicators of all three systems fall in the same order
of magnitude, reaching at most 41--174\,MB for the most
memory-intensive patterns at the full 2.22\,M rows---where the
engine's largest indicators (\texttt{optional\_pattern} at
173.81\,MB and \texttt{quantified} at 164.31\,MB)
exceed Oracle's 58.19\,MB and Trino's roughly 41.5\,MB, while on
\texttt{alternation} and \texttt{complex\_nested} its readings stay
at or below the databases'---and the
operational
footprints differ substantially: the Pandas engine peaks at about
0.67\,GB as a single Python process, the Oracle container stays near
2.5\,GB (about 3.8$\times$ the Pandas peak), and the Trino
container reaches about 23\,GB under its memory-connector setup.
Across the twenty measured runs per cell, run-to-run variability is
low: the median \emph{coefficient of variation} (the standard
deviation as a fraction of the mean) is about 9\% for the Pandas
engine, 6\% for Trino, and 2\% for Oracle; the five warmup runs
remove the cold-start outliers that a single warmup had left behind
for the JVM-based Trino.  These
results are not presented as a claim that one system is universally
better; they show the trade-off observed under the selected
workload, data sizes, and resource configuration.

\paragraph{Why the proposed engine is faster on this workload.}
The measured advantage is a direct consequence of specialization,
not of a generally superior execution model, and it should be read
together with the scope limits stated at the start of this section.
Four factors explain it.  \emph{First}, the engine is a
single-purpose row-pattern matcher: it carries none of the
general-purpose query-planning, catalog, transaction, logging, or
distributed-coordination machinery that Trino and Oracle must run on
every query, so it has almost no fixed per-query overhead to amortize
at these input sizes.  \emph{Second}, execution is entirely
in-process over \texttt{pandas}/NumPy arrays: there is no client
protocol, no result serialization, and no cross-process data movement
between a client and a server, all of which the database paths pay
even when the table is already loaded.  \emph{Third}, the
implementation compiles each query to a narrow fast path tailored to
the evaluated workload---\texttt{ONE ROW PER MATCH}, a single logical
partition, and row-local \texttt{DEFINE} predicates.  These predicates
are evaluated as vectorized column operations (Chapter~\ref{chap:implementation}),
category comparisons are reduced to integer-code checks, the
deterministic automaton is walked from precompiled per-state decision
tables that skip whole constant-mask runs in one step, and the output
columns and running measures are assembled with grouped NumPy
reductions rather than one Python object per matched row.  Each of
these turns a per-row Python operation into an amortized array
operation, which is where the bulk of the speedup arises.
\emph{Fourth}, the benchmark queries order by an already-sorted key
and use no \texttt{PARTITION~BY}, so the engine takes its
linear-scan path and never sorts, whereas the database engines still
route the query through their full operator pipelines.  The
trade-off is equally clear: this speed holds only for the supported
feature subset and single-node in-memory setting, and it costs more
per-query working memory on the sequence-style families
(Table~\ref{tab:memory_cache}); Trino and Oracle remain the
appropriate choice for large-scale, distributed, transactional, or
broader-SQL workloads that fall outside this engine's scope.

Table~\ref{tab:relative_comparison} condenses that trade-off into a
single relative view, expressing the proposed engine's average
across the 30 benchmark cells as a percentage difference from each
database engine.  Under this workload the engine's average execution
time is about 90\% below a fully-warmed Trino's and about 79\% below
Oracle's, with correspondingly higher average throughput
(about 8.3$\times$ Trino's and 4.0$\times$ Oracle's).
Its mean per-query working set is about 41\% above Trino's
counter and essentially equal to Oracle's---the price of the
vectorized fast paths on the sequence-style families---while its
operational footprint is far smaller---about $42\times$ smaller than
the Trino container and about $6.9\times$ smaller than the Oracle
container---because it runs inside a single Python process with no
resident database server.  In short, under this workload the engine
records the lowest average execution time of the three systems while
retaining a much lighter deployment footprint, at the cost of a
per-query memory indicator comparable to the databases' rather than
below them.

\begin{table}[H]
\centering
\caption{Proposed engine relative to each SQL engine, averaged over
  the 30 benchmark cells (a positive value means the proposed engine
  is higher on that metric, a negative value lower).  Lower is better
  for execution time and both memory metrics; higher is better for
  throughput.}
\label{tab:relative_comparison}
\begin{tabular}{lrr}
\toprule
\textbf{Metric} & \textbf{vs.\ Trino~473} & \textbf{vs.\ Oracle XE~21c} \\
\midrule
Execution time & $-90\%$ & $-79\%$ \\
Throughput & $+728\%$ & $+301\%$ \\
Query memory & $+41\%$ & $-0\%$ \\
Operational footprint & $-98\%$ & $-85\%$ \\
\bottomrule
\end{tabular}
\end{table}

% ................................................................
\subsection{Input Ordering: Best-Case and Worst-Case Sorting}
\label{subsec:eval-ordering}
% ................................................................

The benchmark reported so far orders by a sequential \texttt{seq\_id}
whose values are already monotonic in the input, so the proposed
engine takes the \emph{ordered-input} branch of
Section~\ref{subsec:eval-complexity}: a single $O(n)$ monotonicity
check lets it skip \texttt{sort\_values} entirely.  That is realistic
for append-ordered DataFrames, but it exercises only the $O(n)$ side
of the ordering step and gives the engine a shortcut the two database
systems never take, since a SQL table has no guaranteed row order and
\texttt{ORDER~BY} therefore sorts on every query.  To measure the
other side---the $O(n\log n)$ branch the complexity analysis
predicted but the ordered benchmark never triggered---the identical
matrix was rerun with the physical row order of every input
\emph{shuffled} (deterministically, per size), which forces a genuine
sort in all three systems.  Because the logical order after sorting is
unchanged, the match results and counts are byte-identical across the
two scenarios (verified for all 90 cells), so any time difference is
attributable to sorting alone.  Scenario~A (ordered) is thus the
best case and Scenario~B (shuffled) the fair worst case for the
ordering step, and the two bracket the engine's realistic
performance.  Table~\ref{tab:ordering_scenarios} reports the average
per system.

\begin{table}[H]
\centering
\caption{Average execution time under ordered (Scenario~A) and
  shuffled (Scenario~B) input, over all 30 pattern--size cells per
  system.  Scenario~B forces a genuine sort in every system; match
  counts are identical between the two scenarios.}
\label{tab:ordering_scenarios}
\begin{tabular}{lrrr}
\toprule
\textbf{System} & \textbf{A ordered (s)} & \textbf{B shuffled (s)}
  & \textbf{Slowdown} \\
\midrule
Proposed Pandas engine & 0.20 & 0.27 & $1.34\times$ \\
Trino~473              & 1.93 & 2.18 & $1.13\times$ \\
Oracle XE~21c          & 0.94 & 1.09 & $1.16\times$ \\
\bottomrule
\end{tabular}
\end{table}

Three observations follow.  \emph{First}, the proposed engine incurs
the largest relative slowdown ($1.34\times$) because it is the only
system that skips sorting in the ordered case; the penalty is
therefore the full cost of the sort it now performs.  It is widest on
the sequence-style families, where matching is so cheap that the sort
dominates (\texttt{simple\_sequence} $1.56\times$,
\texttt{quantified} $1.48\times$), and narrowest on
\texttt{complex\_nested} ($1.20\times$), where the automaton walk
dominates and the sort is a smaller share of the total.
\emph{Second}, the databases slow down far less ($1.13$--$1.16\times$)
because they sort in \emph{both} scenarios; they benefit only from the
fact that sorting already-ordered rows is cheaper than sorting
shuffled ones (fewer comparisons and moves, better cache locality),
not from skipping the sort.  \emph{Third}, and most important, the
ranking is unchanged: at the full 2.22\,M rows the engine remains
about $4.2\times$ faster than Oracle and $8.1\times$ faster than Trino
under shuffled input (against $4.8\times$ and $9.6\times$ under
ordered input).  The engine's advantage is therefore not an artifact
of the ordered-input shortcut---it survives when every system must
sort.

These measurements also validate the complexity model of
Section~\ref{subsec:eval-complexity} directly.  The incremental cost
of Scenario~B over Scenario~A isolates the ordering step, and it
behaves exactly as an $O(n\log n)$ sort should: at the larger sizes it
becomes essentially \emph{pattern-independent} (its spread across the
five families narrows to a coefficient of variation of about
13\%\ at 800\,K rows and above), confirming that it is the
per-query sort rather than any matching work; and its per-row cost
\emph{rises} with input size, from about 53 to 83\,ns per row between
100\,K and 2.22\,M rows, which is the signature of super-linear
growth---a purely linear step would hold that figure constant.
Normalizing instead by $n\log_2 n$ removes the trend (the figure stays
within a narrow $3.2$--$3.9$ band), and an $n\log n$ fit of the
incremental cost ($R^2 = 0.995$) is closer than a linear one
($R^2 = 0.993$).  Meanwhile the ordered scenario itself remains
approximately linear over the range ($R^2 \geq 0.99$ for a linear
time-versus-size fit).  In short, the empirical results match the
assumed complexity on both sides: linear scaling when the input is
already ordered and the sort is skipped, and an added $O(n\log n)$
term when a genuine sort is required.

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:eval-summary}
% ----------------------------------------------------------------

This chapter evaluated the proposed \texttt{MATCH\_\allowbreak{}RECOGNIZE}
implementation through correctness checks, feature-coverage tests,
complexity analysis, performance measurements, memory analysis, and
comparison with existing systems.  The experiments used five pattern
families across six dataset sizes (up to the full 2{,}222{,}742-row
dataset), with every cell reported as the mean of twenty measured
runs after five warmups (with its standard deviation), producing 30
successful benchmark
cells for the proposed engine and a 90-cell cross-system comparison
against Trino~473 and Oracle~XE~21c under the same nominal resource
budget.

The results show that the implementation preserves the evaluated
semantics for the benchmark pattern families---Trino and Oracle
produced output identical to the engine's normalized output in all
30 pattern--size combinations each (all three systems returned
identical result sets)---while reducing
developer effort compared with hand-written Pandas logic.  Under the
selected workload and resource configuration, the proposed engine
sustained stable throughput of roughly 5.0\,M~rows/s with
approximately linear scaling, the lowest average execution time of
the three systems in this benchmark (0.20\,s, versus 0.94\,s for
Oracle and 1.93\,s for a fully-warmed Trino),
query-memory indicator values in the same order of magnitude as the
database engines, and a substantially smaller operational footprint,
since it runs inside a single Python process without a resident
database server.  The next chapter summarizes the thesis
contributions, discusses limitations, and outlines future work.
