% ============================================================
%  ch7_conclusion_future.tex
%  Chapter 7: Conclusions and Future Work
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% !TEX root = Thesis.tex
\chapter{Conclusions and Future Work}
\label{chap:conclusions}

This chapter concludes the thesis by interpreting the main findings,
stating the remaining limitations, and outlining future research
directions.  Section~\ref{sec:discussion} discusses the practical
meaning of the evaluation results.  Section~\ref{sec:conc-limitations}
summarizes the current limitations.  Section~\ref{sec:conc-validation}
maps the chapter's claims to the evidence used to support them.
Section~\ref{sec:conclusion} states the overall conclusion, and
Section~\ref{sec:conc-future} presents future work.

% ----------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}
% ----------------------------------------------------------------

The results show that a practical subset of SQL row-pattern
recognition can be moved into \texttt{pandas} without abandoning the
declarative structure of \texttt{MATCH\_\allowbreak{}RECOGNIZE}.  The correctness
experiments compared selected reference queries against Trino~473,
Oracle, and hand-written \texttt{pandas} baselines.  For the reported
reference cases, the proposed engine produced the same rows and
measure values as the SQL baselines, including partition-local
ordering, implicit \texttt{DEFINE} behavior, navigation with
\texttt{PREV}, \texttt{MATCH\_NUMBER()}, and measure materialization.
These results do not imply broad equivalence with every feature of
commercial SQL systems; instead, they provide evidence that the
implemented subset preserves the evaluated semantics.

The comparison with idiomatic \texttt{pandas} is also important.
Equivalent procedural implementations require explicit
\texttt{groupby}, \texttt{shift}/\texttt{diff}, candidate filtering,
index bookkeeping, and manual handling of skip behavior.  The SQL
form expresses the same intent in one declarative statement.  This
reduces developer effort, makes the pattern definition easier to
inspect, and supports reuse across notebook-based analysis and SQL
workflows, provided that the query stays within the supported syntax
and semantic subset.

The performance evaluation further supports the practical viability of
the approach for medium-scale in-memory workloads.  Chapter~\ref{chap:evaluation}
reported 30 successful pattern--size benchmark cells (each the mean
of twenty measured runs after five warmups, with its standard
deviation) using five pattern families and six dataset
sizes from 100\,K to the full 2{,}222{,}742-row dataset.  Across these
runs, the implementation achieved an average throughput of
4\,976\,348~rows/s, with throughput remaining stable as input size
increased.  The main variation came
from pattern structure: simple sequences and bounded quantifiers
were the fastest families, while \texttt{complex\_nested}, which
carries the highest transition and condition-evaluation overhead,
was the slowest.  In the cross-system
comparison under
the same nominal resource budget, the engine recorded the lowest
average execution time of the three systems (0.20\,s, versus
1.04\,s for Oracle and 1.98\,s for a fully-warmed Trino): it was
faster than both databases on every one of the 30 pattern--size
cells, by roughly 5.5--14.5$\times$ over Trino and 2.8--6.9$\times$
over Oracle per family.  This advantage follows from the engine's
narrow specialization---no general-purpose query-planning or
distributed-execution overhead, fully in-process execution over
\texttt{pandas}/NumPy arrays, and compiled fast paths that turn
per-row work into vectorized array operations---and holds only for
the supported feature subset and single-node in-memory setting.
The memory measurements
were also bounded for the evaluated range, with the engine's
incremental peak query memory remaining below 300\,MB across the 30
reported scenarios---the highest per-query working memory of the
three systems, the explicit price of the vectorized fast path.

Architecturally, the system benefits from a separation between parsing,
semantic validation, automata construction, matching, and result
materialization.  The implementation uses an ANTLR4 parser based on
the Trino grammar, normalized internal representations, NFA/DFA
construction, a \texttt{pandas}-oriented execution layer, and measure
evaluation.  This modular structure makes the implementation easier to
validate and easier to extend.  It also limits the effect of future
changes: improving pattern compilation, adding a backend, or extending
measure evaluation can be done without redesigning the entire engine.

The optimization layer contributes to this result but should be
interpreted precisely.  The implementation includes LRU-style pattern
caching, bounded memory-management mechanisms, condition-cache
pre-warming, and a parallel-aware scheduling abstraction for partition
work.  The current thesis evaluation, however, should be read as an
evaluation of the implemented \texttt{pandas} execution path rather
than as evidence for a distributed or broadly parallel SQL engine.
This distinction is important because SQL engines such as Trino and
Oracle provide mature optimizers, broader dialect coverage, and
execution infrastructure that remain outside the scope of the present
prototype.

The central finding is therefore scoped but meaningful: the thesis
demonstrates that SQL row-pattern semantics can be compiled into a
finite-automata-based engine and executed directly over DataFrames,
while producing correct results for the evaluated reference scenarios
and predictable performance over the measured input range.

% ----------------------------------------------------------------
\section{Validation of Chapter Claims}
\label{sec:conc-validation}
% ----------------------------------------------------------------

Table~\ref{tab:conc-claim-validation} summarizes the main claims made
in this chapter, the evidence supporting each claim, and the boundary
conditions under which the claim should be interpreted.  This mapping
is intended to keep the conclusion scientifically scoped: each claim is
linked either to the correctness comparisons, the implementation
description, or the measured benchmark results.

{\small
\begin{longtable}{@{}p{0.18\textwidth}p{0.27\textwidth}p{0.25\textwidth}p{0.20\textwidth}@{}}
\caption{Validation mapping between thesis claims, supporting
evidence, and scope boundaries}
\label{tab:conc-claim-validation} \\
\toprule
\textbf{Claim area} & \textbf{Supporting evidence} &
\textbf{Validated conclusion} & \textbf{Scope boundary} \\
\midrule
\endfirsthead
\caption*{Table~\thetable{} (continued): Validation mapping between
thesis claims, supporting evidence, and scope boundaries} \\
\toprule
\textbf{Claim area} & \textbf{Supporting evidence} &
\textbf{Validated conclusion} & \textbf{Scope boundary} \\
\midrule
\endhead
\bottomrule
\endfoot
Correctness &
Selected Trino~473, Oracle, and hand-written \texttt{pandas}
comparisons in Chapter~\ref{chap:evaluation}. &
The reported reference queries produce the same rows and measure
values across the compared systems. &
The evidence covers selected supported scenarios, not every SQL:2016
or vendor-specific case. \\
\midrule
Feature support &
Implemented parser, semantic model, NFA/DFA compilation, output
modes, skip policies, navigation, subsets, anchors, exclusion blocks,
and \texttt{PERMUTE} support described in
Chapters~\ref{chap:methodology} and~\ref{chap:implementation}. &
The system supports a practical row-pattern subset suitable for
DataFrame-centered analysis. &
Unsupported syntax, uncommon edge cases, and cost-based optimizer behavior
remain outside the thesis scope. \\
\midrule
Performance &
The 30-cell Amazon UK Products benchmark over five pattern families
and six dataset sizes from 100\,K to the full 2.22\,M rows, reported
as means of twenty measured runs after five warmups (with standard
deviation, after per-metric $1.5\times$IQR outlier filtering). &
Execution time scales predictably over the measured range
($R^2 \geq 0.99$ for the proposed engine on each of its five pattern
families), with average throughput of
4\,976\,348~rows/s and the lowest average execution time of the
three systems (0.20\,s, versus 1.04\,s for Oracle and 1.98\,s for a
fully-warmed Trino) under the same nominal budget. &
The benchmark does not establish performance for very large,
distributed, or highly adversarial workloads. \\
\midrule
Memory use &
Incremental-peak and footprint measurements for all 30 benchmark
scenarios under one shared cgroup-v2 instrument, including the
engine's maximum observed incremental peak of 290.11\,MB. &
Memory use remains bounded for the evaluated workloads. &
The measurements are empirical and do not define a formal upper bound
for all possible patterns or datasets. \\
\midrule
Developer effort &
Comparison between declarative \texttt{MATCH\_\allowbreak{}RECOGNIZE} queries and
hand-written \texttt{pandas} baselines using grouping, shifting, and
manual skip handling. &
The SQL form reduces procedural code and improves readability for the
evaluated pattern tasks. &
The claim is based on code-level comparison, not a controlled user
study. \\
\midrule
Architecture &
The modular parser--AST--automata--matcher--measure-evaluator pipeline
described in Chapter~\ref{chap:implementation}. &
The design separates concerns and creates extension points for future
backends and optimizations. &
The current implementation remains a \texttt{pandas}-centered
prototype, not a mature distributed SQL optimizer. \\
\end{longtable}
}

The validation evidence also has three important threats to validity.
First, \emph{internal validity} depends on the correctness of the
reference baselines and on careful normalization of case-only column
differences between SQL and \texttt{pandas} outputs.  Second,
\emph{construct validity} depends on whether the selected metrics
capture the intended qualities: row-level equality captures output
correctness for the tested cases, throughput captures execution speed,
and line-count comparison captures only one aspect of developer
effort.  Third, \emph{external validity} is limited by the use of one
workstation, one main real-world dataset, and a bounded range of input
sizes, and by the fact that the Trino and Oracle baselines were run at
their default container configuration rather than expert-tuned, so a
differently tuned deployment could narrow the measured performance
gap.  These threats do not invalidate the reported results, but they
define the conditions under which the results should be generalized.

% ----------------------------------------------------------------
\section{Limitations}
\label{sec:conc-limitations}
% ----------------------------------------------------------------

Several limitations remain and should be considered when interpreting
the results.

\textbf{Supported subset rather than exhaustive SQL coverage.}
The implementation targets a substantial but bounded subset of
\texttt{MATCH\_\allowbreak{}RECOGNIZE}.  It covers core clauses such as
\texttt{PARTITION BY}, \texttt{ORDER BY}, \texttt{PATTERN},
\texttt{DEFINE}, selected \texttt{MEASURES}, supported skip modes,
output modes, navigation functions, subsets, exclusion blocks,
anchors, and \texttt{PERMUTE}.  It does not claim exhaustive SQL:2016
coverage or feature parity with Trino, Oracle, or other production SQL
engines.  Vendor-specific extensions and uncommon edge cases remain
outside the validated scope.

\textbf{Complex pattern and quantifier interactions.}
The automata-based design handles ordinary sequences, alternation,
grouping, bounded repetitions, and common quantified patterns.
However, deeply nested expressions with multiple greedy or unbounded
quantifiers can increase the number of states and transitions
substantially.  Constructs such as \texttt{PERMUTE} are especially
expensive because their direct semantics involve factorial expansion,
so the implementation relies on complexity guards and conservative
fallback behavior for large cases.

\textbf{Measure and aggregate expression scope.}
The engine supports selected aggregate and scalar expressions used by
the implemented measure evaluator, including commonly used functions
such as \texttt{SUM}, \texttt{COUNT}, \texttt{MIN}, \texttt{MAX},
\texttt{AVG}, statistical aggregates, collection/string aggregates,
conditional aggregates, and supported \texttt{RUNNING}/\texttt{FINAL}
forms.  Nevertheless, arbitrary user-defined aggregate functions and
all possible SQL expression combinations remain outside the current
scope.
This is a natural limitation of an in-memory prototype that implements
its own expression evaluator rather than reusing a general SQL
execution engine.

\textbf{Performance boundaries.}
The benchmark demonstrates predictable behavior from 100\,K to the
full 2.22\,M rows on the evaluated workstation.  Larger datasets, wider schemas,
more partitions, and highly ambiguous patterns may expose bottlenecks
from Python execution, \texttt{pandas} group processing, row-context
construction, and predicate evaluation.  Native SQL engines and
compiled columnar systems can still provide higher throughput for
large-scale workloads.

\textbf{Memory behavior.}
Memory usage remained bounded in the reported benchmark, and the
engine-reported query memory grows with input size as expected.
Readings near the measurement floor are, however, quantized to the
allocation granularity of each runtime (memory-pool chunks for the
databases, resident pages for Python), and small values can also be
affected by garbage-collection timing and result-set
materialization.  Therefore, the reported memory results should be
interpreted as empirical measurements for the tested workloads rather
than as a formal upper bound for all possible patterns.

\textbf{Optimizer integration.}
The engine operates independently of database query optimizers.  It
does not currently perform cost-based plan selection, predicate
pushdown into external storage, join reordering, or distributed
resource scheduling.  As a result, some plan-level optimizations
available in database systems are not available in the current
implementation.

% ----------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
% ----------------------------------------------------------------

This thesis presented a SQL-in-\texttt{pandas} implementation for
executing a supported subset of \texttt{MATCH\_\allowbreak{}RECOGNIZE} over
DataFrames.  The work contributes a parsing and validation pipeline
based on Trino SQL syntax, an internal representation of row-pattern
queries, finite-automata-based pattern compilation, a
\texttt{pandas}-based matcher, measure evaluation, and optimization
support through caching and memory-aware execution mechanisms.

The evaluation provides three main conclusions.  First, the selected
correctness scenarios matched the reference outputs from Trino,
Oracle, and hand-written \texttt{pandas} baselines.  Second, the
declarative SQL formulation substantially reduces the amount of
procedural Python code needed to express sequential row-pattern
analysis.  Third, the benchmark over the Amazon UK Products Dataset
shows stable performance across 30 pattern--size cells from 100\,K to
the full 2.22\,M rows, with an average throughput of
4\,976\,348~rows/s, the lowest average execution time of the three
systems under the same nominal resource budget (0.20\,s, versus
1.04\,s for Oracle and 1.98\,s for a fully-warmed Trino), and
bounded memory use in the measured range.

The proposed approach is therefore best understood as a bridge between
SQL row-pattern recognition and DataFrame-centered analytics.  It does
not replace mature SQL engines, but it makes a useful subset of their
row-pattern functionality available inside Python workflows where data
are already represented as \texttt{pandas} DataFrames.

% ----------------------------------------------------------------
\section{Future Work}
\label{sec:conc-future}
% ----------------------------------------------------------------

Future work can extend the implementation along five complementary
directions.

\textbf{Broader SQL and dialect coverage.}
The most direct extension is to increase coverage of SQL:2016
\texttt{MATCH\_\allowbreak{}RECOGNIZE} semantics and selected vendor behavior.
This includes additional navigation combinations, broader
expression support, richer aggregate handling, more extensive
\texttt{RUNNING}/\texttt{FINAL} semantics, and a larger conformance
suite based on Trino, Oracle, and standard-derived examples.

\textbf{Parallel and distributed execution.}
Independent partitions provide a natural unit of parallel work.
Future versions can connect the existing parallel-aware scheduling
abstraction to real concurrent execution, while preserving deterministic
partition ordering and skip semantics.  A further step is distributed
execution over frameworks such as Dask or Spark, which would require
boundary-aware state propagation for patterns whose matches may span
partition chunks.

\textbf{Backend abstraction and columnar acceleration.}
The current execution path is centered on \texttt{pandas}.  A backend
abstraction could target Polars, Apache Arrow, or other columnar
engines.  This would allow more predicate evaluation to run in
compiled vectorized kernels and could reduce Python-level row
iteration.  Additional acceleration through Cython, Numba, or
specialized compiled predicates is also a promising direction, but it
requires careful validation to preserve SQL-style null handling,
navigation semantics, and deterministic output.

\textbf{Optimization and diagnostics.}
The automata pipeline can be improved through stronger state pruning,
guard pushdown, more precise selectivity estimates, and better
handling of high-ambiguity quantified patterns.  Diagnostic tooling is
also important: explain plans, NFA/DFA visualizations, match traces,
and per-pattern performance counters would make it easier to debug
incorrect patterns and understand performance behavior.

\textbf{Usability and integration.}
Future work can improve the user-facing workflow by adding clearer
error messages, notebook-friendly visualizations, reusable test
fixtures, and integration with existing DataFrame query interfaces.
These additions would make the engine easier to adopt in exploratory
data analysis and teaching settings.

% ----------------------------------------------------------------
\section{Chapter Summary}
\label{sec:conc-summary}
% ----------------------------------------------------------------

This chapter summarized the thesis findings and placed them in a
scientifically scoped context.  The implementation was shown to
support a useful subset of \texttt{MATCH\_\allowbreak{}RECOGNIZE} over
\texttt{pandas}, with correctness evidence from selected reference
queries and performance evidence from a 40-run benchmark over
medium-scale DataFrames.

The chapter also identified the main limitations: bounded SQL
coverage, complex-pattern growth, expression-evaluation scope,
large-scale performance constraints, memory variability, and limited
optimizer integration.  The proposed future work targets these issues
through broader conformance testing, parallel and distributed
execution, backend abstraction, optimization improvements, and better
developer tooling.
