SQL:2016 introduced \texttt{MATCH\_RECOGNIZE} to support row
pattern recognition, allowing data analysts to detect ordered sequences of
rows that match according to a user-defined pattern. This feature is useful in
many domains such as financial monitoring, climate analysis, security logs, sensor analytics,
and e-commerce, where the significant information frequently constitutes a sequence of events rather than an isolated record.  Several database
systems, including Oracle, Snowflake, Trino, and Apache Flink, now
provide row-pattern functionality, but Pandas does not offer a native
equivalent for in-memory Python analytics.  As a result, analysts must
often reproduce sequential patterns with shifted columns, manual loops,
and custom state management, which makes the code longer, less
readable, and more prone to errors.  This thesis investigates how core
SQL:2016 \texttt{MATCH\_RECOGNIZE}-style behavior can be implemented
directly over Pandas DataFrames.  The proposed engine parses
SQL-based \texttt{MATCH\_RECOGNIZE} queries from a string, constructs
a structured query representation, validates the supported
row-pattern clauses, translates pattern expressions into finite
automata, and executes the resulting matcher over ordered DataFrame
partitions.  The
implementation supports the evaluated Feature~R010-style scope,
including \texttt{PARTITION BY}, \texttt{ORDER BY}, \texttt{PATTERN},
\texttt{DEFINE}, \texttt{MEASURES},skip policies, output modes, 
quantifiers, alternation, grouping, anchors, exclusions,
\texttt{PERMUTE}, and selected navigation and aggregate functions.
It also uses controlled predicate evaluation instead of executing raw
Python expressions.  The system was evaluated through targeted
correctness tests, comparisons with Trino and Oracle reference
outputs, and performance experiments on samples from the Amazon UK
Products~2023 catalog.  The results show that the proposed approach
preserves the evaluated row-pattern semantics while reducing developer
effort compared with hand-written Pandas implementations.  The thesis
demonstrates that bringing \texttt{MATCH\_RECOGNIZE}-style functionality
to Pandas can bridge SQL and Python workflows, making sequential 
data analysis more readable, portable, and productive.

\vspace{\baselineskip}
\noindent\textbf{Keywords:}  \texttt{MATCH\_RECOGNIZE}; Row Pattern Recognition;
Pandas; SQL:2016; Finite Automata; Pattern Matching.
