% ============================================================
%  ch2_background_concepts.tex
%  Chapter 2: Background and Concepts
%  Master's Thesis -- SQL MATCH_RECOGNIZE on Pandas DataFrames
% ============================================================
% ================================================================
%  CHAPTER 2 -- BACKGROUND AND CONCEPTS
% ================================================================
\chapter{Background and Concepts}
\label{chap:background}

This chapter establishes the theoretical and practical foundations
required to understand the design and implementation of the
\texttt{MATCH\_RECOGNIZE} engine for Pandas DataFrames.
Section~\ref{sec:bg-pattern-matching} introduces the problem of
sequential pattern detection in relational data.
Section~\ref{sec:bg-match-recognize} examines the SQL:2016
\texttt{MATCH\_RECOGNIZE} clause in detail.
Section~\ref{sec:bg-performance} in Chapter~4 describes performance
optimization techniques. Chapter~3 surveys related work and identifies
the research gap addressed by this thesis.

% ----------------------------------------------------------------
\section{Pattern Matching in Sequential Data Analysis}
\label{sec:bg-pattern-matching}
% ----------------------------------------------------------------

Sequential pattern detection represents a fundamental challenge in
data analysis, where the objective is to identify ordered sequences
of events or states within large datasets.  In contrast to
conventional queries that match rows individually against one or
more predicates, pattern matching takes preceding and subsequent
rows into consideration and returns groups of rows when they follow
a specified pattern.  This makes it possible to retrieve specific
shapes---such as V-, W- or U-shapes---within the data, making
pattern matching predestined for all kinds of processes driven by
sequences of events: detecting unusual behaviour in security
applications, analysing stock price development, sensor data
analysis, fraud detection, and generally complex event
processing~\cite{OracleMR}.

Unlike traditional relational operations that focus on individual
rows or aggregated values, pattern matching examines the temporal
or logical ordering of data points to extract meaningful sequences.
This capability is essential across many domains including financial
market analysis, fraud detection, IoT sensor monitoring, and
behavioural analytics.

Historically, relational database systems excelled at set-based
operations---selection, projection, joins, and aggregation---but
lacked expressive mechanisms for sequence analysis.  Analysts
requiring pattern detection were forced to either export data to
specialised stream-processing systems, implement complex procedural
logic using cursors and temporary tables, or employ application-layer
processing with significant performance overhead.  These limitations
created a gap between the widespread availability of sequential data
and the tools available to analyse it effectively.  The introduction
of row pattern recognition in SQL:2016 addresses this deficiency by
bringing sequence analysis capabilities directly into the query
language~\cite{ISO2016SQL}.

% ................................
\subsection{Real-World Applications}
\label{subsec:bg-applications}
% ................................

\paragraph{Financial Market Analysis.}
Detecting V-shaped price patterns---price decline followed by
recovery---requires examining ordered sequences of price movements.
For example, identifying stocks whose price troughs are successively
higher (a ``higher-low'' pattern) signals a potential upward trend.
The canonical example used throughout this thesis is the detection
of such V-patterns in customer order data~\cite{OracleMR}.

\paragraph{Fraud Detection.}
Credit card fraud often manifests as specific transaction sequences,
such as multiple small test charges followed by a large purchase, or
geographically impossible transactions occurring within unrealistic
timeframes.  Expressing these rules declaratively in SQL is more
maintainable than equivalent procedural logic.

\paragraph{IoT and Sensor Data.}
Manufacturing systems monitor sensor readings to detect equipment
degradation patterns, such as a gradual temperature rise followed by
a sudden drop, which may indicate imminent failure.  Pattern queries
over time-series sensor streams are a natural fit for
\texttt{MATCH\_RECOGNIZE}.

\paragraph{Clickstream and Funnel Analysis.}
E-commerce platforms analyse user navigation sequences to identify
conversion paths or abandonment patterns.  A classic funnel query
detects the sequence
``view~$\rightarrow$ add-to-cart~$\rightarrow$ remove~$\rightarrow$
exit,'' which signals purchase hesitation.  With
\texttt{MATCH\_RECOGNIZE}, the analyst can also quantify \emph{how
many} add-to-cart events occurred before removal, extract the exact
timestamps of each step, and compute the total dwell time in a single
declarative query---tasks that would otherwise require multiple
correlated subqueries or procedural session-stitching logic.

\paragraph{Network Security and Intrusion Detection.}
Cyber-attacks rarely manifest as a single anomalous event; they
unfold as ordered sequences of actions.  A typical multi-stage
intrusion follows the pattern: port scanning~$\rightarrow$ repeated
failed authentication attempts~$\rightarrow$ a single successful
login~$\rightarrow$ privilege escalation.
\texttt{MATCH\_RECOGNIZE} can express this entire kill-chain as one
pattern query over an event log, flagging the offending session and
extracting the first and last timestamps, the targeted port, and the
account compromised---all without exporting the log to a separate
SIEM tool~\cite{petkovic2022fraud}.

\paragraph{Healthcare and Patient Monitoring.}
In clinical data streams, early warning systems must detect
physiological deterioration before a critical event.  A typical
sepsis-onset pattern consists of rising heart rate rows, followed by
a drop in blood pressure rows, followed by a spike in white-blood-cell
count.  Expressing this over an ordered time-series of vital signs
with \texttt{MATCH\_RECOGNIZE} allows hospitals to produce real-time
alerts from the same SQL infrastructure used for reporting, without
duplicating data into a separate stream-processing system.

\paragraph{Supply Chain and Logistics.}
Operational disruptions in supply chains often emerge as a cascade:
a single late shipment~$\rightarrow$ repeated partial deliveries~$\rightarrow$
a stockout event.  \texttt{MATCH\_RECOGNIZE} applied to an order
fulfilment log can identify such escalation sequences per supplier or
warehouse and measure the time between first delay and eventual
stockout, enabling proactive mitigation rather than retrospective
reporting.

\paragraph{Financial Compliance and Trade Surveillance.}
Regulatory frameworks such as MiFID~II require exchanges to detect
potential market manipulation patterns, for example
``layering,'' where a trader places a large resting order
(to move the price), then executes a trade on the opposite side, then
cancels the resting order.  This three-step sequence---place, trade,
cancel---with precise timing constraints is naturally expressed as a
\texttt{MATCH\_RECOGNIZE} pattern, turning a complex compliance rule
directly into an auditable SQL query.


% ----------------------------------------------------------------
\section{The SQL:2016 \texttt{MATCH\_RECOGNIZE} Clause}
\label{sec:bg-match-recognize}
% ----------------------------------------------------------------

Nearly a decade ago, the SQL:2016 standard introduced
\texttt{MATCH\_RECOGNIZE}~\cite{ISO2016SQL,iso2016}, a construct
which enabled row pattern matching in SQL.  Row pattern matching is
a technique which allows finding sequences of rows based on a
pattern defined by regular expressions.  Using a syntax similar to
regular expressions, it is possible to filter data sets processed
by pattern specifications to recognise complex sequences of
events~\cite{winand2017sqlnew}.  This process facilitates the
recognition of patterns in data series that would otherwise require
the use of additional SQL clauses, such as the \texttt{WINDOW}
clause.  Consequently, the \texttt{MATCH\_RECOGNIZE} clause does
not introduce new data-filtering power but rather serves as a
shortcut to simplify special areas of application that can
otherwise be circumvented by combining other clauses.

Up to now, \texttt{MATCH\_RECOGNIZE} is not supported by many
database management systems, including
PostgreSQL~\cite{howe2020matchrecognize} and
DuckDB~\cite{findling2025bringing}.  This thesis contributes to
the overarching research topic of making pattern matching available
to DBMSs that do not natively support it, so that a wider range of
users can benefit from its power and expressiveness.  The clause
brings capabilities previously available only in specialised
stream-processing systems into standard relational databases.

The SQL standard distinguishes between different feature levels for
row pattern matching, which can be used to classify
implementations~\cite{winand_match_recognize}.
\textbf{Feature~R010} provides the most basic support and serves
as a basis for further features; it includes row pattern
recognition inside the \texttt{FROM} clause and basic aggregate
functions (\texttt{MIN}, \texttt{MAX}, \texttt{SUM},
\texttt{COUNT}, \texttt{AVG}) inside the \texttt{MEASURES}
clause.  \textbf{Feature~R020} extends the areas of application:
\texttt{MATCH\_RECOGNIZE} can be used within the \texttt{WINDOW}
and \texttt{OVER} clauses to describe windows more precisely.
\textbf{Feature~R030} allows full aggregate support inside the
\texttt{MATCH\_RECOGNIZE}
clause~\cite{ISO2021,michels2018new,melton2017sqlrpr}.

% ................................
\subsection{Clause Structure and Components}
\label{subsec:bg-clause-structure}
% ................................

The clause operates on a rowset (table or derived table) and
consists of eight interconnected sub-clauses, as shown in
Listing~\ref{lst:mr-syntax}.

\begin{lstlisting}[
    language     = SQL,
    caption      = {General syntax of the \texttt{MATCH\_RECOGNIZE} clause},
    label        = {lst:mr-syntax}
]
SELECT <columns>
FROM   <table>
MATCH_RECOGNIZE (
    PARTITION BY <partition_columns>   -- 1. Data organisation
    ORDER BY     <ordering_columns>    -- 2. Row sequencing
    MEASURES     <expressions>         -- 3. Output computations
    <rows_per_match>                   -- 4. Result granularity
    AFTER MATCH SKIP <option>          -- 5. Overlap handling
    PATTERN ( <pattern_regex> )        -- 6. Sequence specification
    SUBSET   <union_variable_defs>     -- 7. Union variables (optional)
    DEFINE  <variable_definitions>     -- 8. Variable predicates
) AS <alias>
\end{lstlisting}

\begin{enumerate}

\item \textbf{PARTITION BY.}
  Divides the input rowset into independent partitions, analogous to
  window-function partitioning.  Pattern matching runs independently
  within each partition, enabling both parallel processing and
  domain-specific grouping (e.g.\ per customer or per sensor).

\item \textbf{ORDER BY.}
  Establishes the logical ordering of rows within each partition.
  The pattern-matching algorithm processes rows strictly according
  to this ordering.  Multiple sort keys with \texttt{ASC}/
  \texttt{DESC} modifiers are supported.

\item \textbf{MEASURES.}
  Defines computed expressions that extract values from matched row
  sequences using pattern variables and navigation functions
  (\texttt{FIRST}, \texttt{LAST}, \texttt{PREV}, \texttt{NEXT}).
  Each \texttt{MEASURE} expression becomes a column in the output.

\item \textbf{Output options.}
  \begin{itemize}
    \item \texttt{ONE ROW PER MATCH}: one summary row per complete
      match.
    \item \texttt{ALL ROWS PER MATCH}: one row for every row
      participating in a match.
    \item \texttt{ALL ROWS PER MATCH WITH UNMATCHED ROWS}: includes
      non-matching rows in the result.
  \end{itemize}

\item \textbf{AFTER MATCH SKIP.}
  Controls how the engine resumes after finding a match:
  \begin{itemize}
    \item \texttt{SKIP TO NEXT ROW} -- resume from the row after
      match start (maximum overlaps).
    \item \texttt{SKIP PAST LAST ROW} -- resume from the row after
      match end (non-overlapping matches).
    \item \texttt{SKIP TO FIRST} \textit{var} -- resume from the
      first row classified as \textit{var}.
    \item \texttt{SKIP TO LAST} \textit{var} -- resume from the last
      row classified as \textit{var}.
  \end{itemize}

\item \textbf{PATTERN.}
  Specifies the sequence using regular-expression-like syntax with
  pattern variables as terminals:
  \begin{itemize}
    \item Concatenation: \texttt{A~B~C} (A followed by B followed by C).
    \item Alternation: \texttt{A~|~B} (either A or B).
    \item Quantifiers: \texttt{A*} (zero or more), \texttt{A+}
      (one or more), \texttt{A?} (zero or one), \texttt{A\{n\}}
      (exactly $n$), \texttt{A\{n,m\}} (between $n$ and $m$).
    \item Grouping: \texttt{(A~B)+} (one or more AB pairs).
    \item Anchors: \texttt{\^{}A~B~C\$} (pattern must span the
      entire partition).
    \item Exclusions: \texttt{\{-~A+~B+~-\}} (rows \emph{not}
      matching the enclosed sub-pattern).
    \item Permutations: \texttt{PERMUTE(A,~B,~C)} (any ordering
      of A, B, and C).
  \end{itemize}

\item \textbf{SUBSET.}
  An optional clause for defining union variables, each of which
  consists of a set of variables defined in the \texttt{PATTERN}
  clause.  Union variables can be used to match rows to one of
  several different patterns, and may appear inside
  \texttt{MEASURES} and \texttt{DEFINE} expressions.

\item \textbf{DEFINE.}
  Assigns Boolean predicates to pattern variables.  The initial
  step in execution involves implementing a filter on the ordered
  data using this clause, which specifies a list of named Boolean
  expressions (\emph{steps}) that either enable or deny the
  inclusion of a row within the matched sequence based on a
  predetermined criterion.  Each row is evaluated against these
  predicates to determine its variable classification.  Predicates
  may reference current-row columns, navigation functions,
  aggregations over matched sub-sequences, and Boolean/comparison
  operators~\cite{winand_match_recognize}.

\end{enumerate}

The complete \texttt{MATCH\_RECOGNIZE} clause is embedded in a
\texttt{SELECT} query and can therefore be further processed by
additional specifications.  Possible extensions include applying
groupings or aggregate functions to the outer
\texttt{SELECT}~\cite{winand_match_recognize}.

% ................................
\subsection{Illustrative Example: Stock Price V-Pattern}
\label{subsec:bg-example}
% ................................

Listing~\ref{lst:mr-example} shows a complete \texttt{MATCH\_RECOGNIZE}
query that detects V-shaped price patterns in customer order data.

\begin{lstlisting}[
    language = SQL,
    caption  = {Detecting V-shaped price patterns with \texttt{MATCH\_RECOGNIZE}},
    label    = {lst:mr-example}
]
SELECT customer_id,
       start_price, bottom_price, final_price,
       start_date,  final_date
FROM   orders
MATCH_RECOGNIZE (
    PARTITION BY customer_id
    ORDER BY     order_date
    MEASURES
        START.price          AS start_price,
        LAST(DOWN.price)     AS bottom_price,
        LAST(UP.price)       AS final_price,
        START.order_date     AS start_date,
        LAST(UP.order_date)  AS final_date
    ONE ROW PER MATCH
    AFTER MATCH SKIP PAST LAST ROW
    PATTERN (START DOWN+ UP+)
    DEFINE
        DOWN AS price < PREV(price),
        UP   AS price > PREV(price)
);
\end{lstlisting}

The engine partitions rows by \texttt{customer\_id}, sorts by
\texttt{order\_date}, and seeks sequences matching
\texttt{START~DOWN+~UP+}: one \texttt{START} row (any row, as it
carries no \texttt{DEFINE} predicate), one or more \texttt{DOWN}
rows (each with a strictly lower price than its predecessor), and
one or more \texttt{UP} rows (each with a strictly higher price
than its predecessor).  For every match found, the \texttt{MEASURES}
sub-clause extracts the starting price and date from the
\texttt{START} row, the lowest price from the last \texttt{DOWN}
row, and the final price from the last \texttt{UP} row.

% ................................
\subsection{Comparison of Existing Implementations}
\label{subsec:bg-existing}
% ................................

Several database and stream-processing systems have implemented
\texttt{MATCH\_RECOGNIZE}.  The SQL standard defines three feature
profiles for classifying those implementations~\cite{ISO2021,michels2018new}:
\textbf{Feature~R010} covers row pattern recognition in the
\texttt{FROM} clause together with basic aggregate functions;
\textbf{Feature~R020} additionally permits \texttt{MATCH\_RECOGNIZE}
inside \texttt{WINDOW}/\texttt{OVER} clauses; and
\textbf{Feature~R030} enables full aggregate support within the
clause.  In practice, most systems have adopted only Feature~R010
as the baseline.  Snowflake is the only publicly available engine
that additionally implements Feature~R020.  Feature~R030 has not
been adopted by any major production system to date.

The following paragraphs briefly characterise each implementation
included in Table~\ref{tab:bg-systems}.

\textbf{Oracle} introduced \texttt{MATCH\_RECOGNIZE} in
Database~12c (2013) and remains the primary reference
implementation~\cite{oracle2022,oracle2021dwhsg,book}.
It implements Feature~R010 using an NFA-based engine with
backtracking optimisations for complex patterns.  Feature~R020 is
not supported; the clause cannot appear inside a \texttt{WINDOW}
function.  Oracle also enforces several additional restrictions:
Boolean expressions are not permitted in the \texttt{MEASURES}
clause, and \texttt{DISTINCT} may not precede aggregate functions.
Despite its technical depth, Oracle operates under a commercial
licence, making it inaccessible for many users.

\textbf{Trino} is an open-source distributed query engine whose
\texttt{MATCH\_RECOGNIZE} implementation is DFA-based, guaranteeing
deterministic pattern matching in linear time without
backtracking~\cite{trino2022,trino2021row}.  It supports
Feature~R010 only---the clause may only appear in the
\texttt{FROM} clause.  Trino's architecture is designed for
parallelised, federated query execution across large clusters, which
differentiates it from single-node systems.

\textbf{Snowflake} is a cloud-native data platform that covers
both Feature~R010 and Feature~R020~\cite{snowflake2025match_recognize,hoffa2021snowflake}.
In contrast to the SQL standard, Snowflake makes no syntactic
distinction between the \texttt{FROM} and \texttt{WINDOW} usage
contexts: the two are syntactically equivalent, and Feature~R020
is realised purely through semantic interpretation.  Snowflake
charges by compute time (warehouse size), so query costs scale
with workload.

\textbf{Apache Flink} targets real-time stream processing and
implements \texttt{MATCH\_RECOGNIZE} through its Complex Event
Processing (CEP) library rather than as a native SQL
sub-clause~\cite{vsprem2024building,georgiou2025streaming}.
This makes Flink highly effective for high-velocity continuous
data, but it introduces a separate programming model that is
less familiar to analysts accustomed to batch SQL.

\textbf{DuckDB} is an embeddable, in-process analytical database
that does not yet provide native \texttt{MATCH\_RECOGNIZE}
support~\cite{raasveldt2019duckdb,findling2025bringing}.  Because
DuckDB requires no separate server and exposes a Python API, it
is architecturally close to this thesis's target environment.
Ongoing research investigates transpiling \texttt{MATCH\_RECOGNIZE}
to \texttt{WITH~RECURSIVE} inside DuckDB as a path toward
Feature~R010 compliance.

Table~\ref{tab:bg-systems} summarises these five systems alongside
this work.

\begin{table}[htbp]
  \centering
  \caption{Existing \texttt{MATCH\_RECOGNIZE} implementations}
  \label{tab:bg-systems}
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{llccl}
    \toprule
    \textbf{Engine} & \textbf{Feature Level} & \textbf{Open Source} & \textbf{Deployment} & \textbf{Licence} \\
    \midrule
    Oracle       & R010 (no R020)  & No  & On-premise / Cloud & Commercial  \\
    Trino        & R010            & Yes & Distributed        & Apache 2.0  \\
    Snowflake    & R010 + R020     & No  & Cloud-native       & Commercial  \\
    Apache Flink & CEP library     & Yes & Streaming cluster  & Apache 2.0  \\
    DuckDB       & None (planned)  & Yes & In-process         & MIT         \\
    \midrule
    \textbf{This work} & \textbf{R010 (full)} & \textbf{Yes} & \textbf{In-process Python} & \textbf{Open-source} \\
    \bottomrule
  \end{tabular}
\end{table}

Two structural gaps are visible from this overview.  First, every
system with native \texttt{MATCH\_RECOGNIZE} support---Oracle,
Trino, Snowflake, and Flink---requires a separate server, a
distributed cluster, or a paid cloud subscription, making them
unsuitable for lightweight or in-process data science.  Second,
none of these systems integrates with Python's Pandas ecosystem,
which is the dominant platform for in-process analytical workloads.
This thesis addresses both gaps by delivering a full
Feature~R010-compliant \texttt{MATCH\_RECOGNIZE} engine that runs
entirely inside a Pandas workflow with zero additional deployment.

Chapter~\ref{chap:related} examines each of these systems in
greater depth, documenting the precise restrictions each imposes,
comparing their execution strategies, and situating the contribution
of this work within the broader landscape of pattern-matching
research.
